---
title: "Pedigree Diagram: kinship2 Reference Comparison"
subtitle: "Decision-support research for issues #142 (rectilinear mate-line/sibship-bar style, shipped), #143/#144 (founder-positioning fixes, shipped), and #145 (sire/dam left-right ordering, open) -- not a package vignette, not shipped"
date: today
format:
  html:
    embed-resources: true
    toc: true
    df-print: kable
---

## Purpose

The owner asked, after seeing `nprcgenekeepr`'s live Pedigree Diagram tab
render a real family, whether the drawing is right -- pointing at
<https://rpubs.com/dliupress/pedigreedemo> as the reference for what a
"correct" pedigree drawing looks like. That page uses the **kinship2** R
package (a real, well-known pedigree-drawing package, not part of
`nprcgenekeepr`'s own dependencies) to plot a small set of example pedigrees.

This document reproduces those *exact same* example pedigrees -- same
individuals, same parent/child/sex relationships -- using
`nprcgenekeepr`'s own current pedigree-diagram capability
(`makePedigreeMatingLayout()` + `visNetwork`, the identical render chain
`R/modPedigree.R`'s Diagram tab uses today), placed directly next to
kinship2's own rendering of the identical data. The goal is to give the owner
a fair, apples-to-apples visual and factual basis for deciding whether (and
how) to pursue diagram-layout changes -- **no decision is made in this
document, and no package code is changed here.**

Originally written for [issue #142](https://github.com/rmsharp/nprcgenekeepr/issues/142)
(shipped since), this comparison also caught a real founder-positioning
defect along the way -- independent of #142, fixed via issues
[#143](https://github.com/rmsharp/nprcgenekeepr/issues/143)/[#144](https://github.com/rmsharp/nprcgenekeepr/issues/144)
(see Examples 1-2) -- and now includes a fourth worked example investigating
[issue #145](https://github.com/rmsharp/nprcgenekeepr/issues/145)'s sire/dam
left-right ordering question directly against kinship2's own default
behavior (Example 4), refreshed per
`docs/audits/PEDIGREE_DIAGRAM_BACKLOG_SEQUENCING_AUDIT_2026-08-08.md`'s Tier
1 step 3 recommendation.

`kinship2` is installed only into this project's local `renv` library for
this one-off comparison (not added to `DESCRIPTION`, not snapshotted into
`renv.lock`) -- it is reference material, not a dependency.

## Setup

```{r setup}
#| message: false
suppressWarnings(suppressMessages(library(kinship2)))
library(nprcgenekeepr)
library(visNetwork)
```

### Helper: convert a kinship2-style data frame to nprcgenekeepr's pedigree format

kinship2's own examples use `id`/`father`/`mother`/`sex` with `0` meaning "no
parent" and `sex` coded `1` = male, `2` = female. `nprcgenekeepr` uses
`id`/`sire`/`dam`/`sex` (`"M"`/`"F"`/`"H"`/`"U"`, `NA` for no parent) plus a
`gen` column from its own exported `findGeneration()`. `convertSexCodes()` --
already exported by `nprcgenekeepr` -- converts kinship2's `1`/`2` codes
directly.

```{r helper-convert}
to_nprc_ped <- function(id, father, mother, sex) {
  ped <- data.frame(
    id = as.character(id),
    sire = ifelse(father == 0, NA_character_, as.character(father)),
    dam = ifelse(mother == 0, NA_character_, as.character(mother)),
    sex = as.character(convertSexCodes(sex)),
    stringsAsFactors = FALSE
  )
  ped$gen <- findGeneration(ped$id, ped$sire, ped$dam)
  ped
}
```

### Helper: render via nprcgenekeepr, matching `R/modPedigree.R`'s exact current chain

```{r helper-render}
render_nprc_diagram <- function(ped) {
  layout <- makePedigreeMatingLayout(ped)
  visNetwork::visNetwork(layout$nodes, layout$edges) |>
    visNetwork::visPhysics(enabled = FALSE) |>
    visNetwork::visNodes(physics = FALSE) |>
    visNetwork::visEdges(smooth = FALSE) |>
    visNetwork::visLegend(
      addNodes = data.frame(
        label = c(
          "Female", "Male", "Hermaphrodite", "Unknown", "Other / Unrecorded"
        ),
        shape = c("dot", "square", "star", "triangle", "diamond"),
        stringsAsFactors = FALSE
      ),
      useGroups = FALSE, position = "right", main = "Sex",
      width = 0.28, stepY = 65L
    ) |>
    visNetwork::visOptions(
      nodesIdSelection = list(
        enabled = TRUE,
        values = layout$nodes$id[
          !grepl("^__union_|^__dup_", layout$nodes$id)
        ]
      ),
      highlightNearest = list(enabled = TRUE, hover = TRUE, degree = 1L,
                               algorithm = "all")
    )
}
```

## How to read each example

For each pedigree below:

1. **kinship2's own rendering** (`plot.pedigree()`) -- the reference
   convention: mates joined by a horizontal line at their own level; from its
   midpoint, a vertical line drops to a horizontal sibship bar; each child
   rises to meet that bar with its own vertical line. Strict right angles,
   generations in clean rows.
2. **nprcgenekeepr's current rendering** -- the exact code path
   `R/modPedigree.R`'s Diagram tab runs today, fed the *identical* data.

## Example 1: `sample.ped` family 2 (14 people, no multi-mate individuals)

The cleanest comparison -- a single 3-generation family tree with no
duplicated/multi-mate individuals, so the ONLY variable being compared is the
mate-line/sibship-bar edge-routing convention itself.

```{r fam2-data}
data(sample.ped)
fam2 <- sample.ped[sample.ped$ped == 2, ]
nprcPed2 <- to_nprc_ped(fam2$id, fam2$father, fam2$mother, fam2$sex)
nprcPed2
```

### kinship2

```{r fam2-kinship2}
#| fig-width: 9
#| fig-height: 5
ped2 <- kinship2::pedigree(
  id = fam2$id, dadid = fam2$father, momid = fam2$mother, sex = fam2$sex
)
plot(ped2)
title(main = "kinship2::plot.pedigree() -- sample.ped family 2")
```

### nprcgenekeepr

```{r fam2-nprc}
render_nprc_diagram(nprcPed2)
```

**Founder positioning (fixed since this document was first written -- see
below):** in this rendering, `203` (a founder with no recorded parents, who
mates with `204`, a *daughter* of `201`x`202`, generation 1) now positions
correctly, adjacent to his own mate's row, not inside `201`x`202`'s mate-line
span. This document originally caught a real defect here -- `203` used to
land in the founders' own row (`y = 0`), visually reading as if `201` and
`203`, two males, were the mated pair -- confirmed and fixed via issues
[#143](https://github.com/rmsharp/nprcgenekeepr/issues/143) and
[#144](https://github.com/rmsharp/nprcgenekeepr/issues/144) (S472/S474,
2026-08-04). Current, correct layout data:

```{r fam2-defect-evidence}
layout2 <- makePedigreeMatingLayout(nprcPed2)
n2 <- layout2$nodes[layout2$nodes$id %in% c("201", "202", "203", "204"),
                     c("id", "x", "y")]
n2[order(n2$y, n2$x), ]
```

`203` is now at `y = 150`, next to `__union_2` and `204` -- his own actual
mate union -- not in the founders' `y = 0` row at all. The founders' row now
holds only `202`, `__union_1` (the true midpoint of `201`x`202`'s own
mate-line), and `201`. This matches kinship2's own convention exactly (a
founder who marries into a later generation is positioned adjacent to their
actual mate's row, not at their own generation), closing the gap this
document's Summary table originally flagged. Root cause and a second,
independent confirmation (family 1, ids `116`/`117`) are discussed after
Example 2 below.

## Example 2: `sample.ped` family 1 (41 people, includes real multi-mate individuals)

A larger, more complex family. kinship2's own layout duplicates 2 individuals
here (ids `103` and `138`) purely for layout/crossing-minimization reasons --
each has only ONE mate, but their own birth family is drawn far from their
marital family in kinship2's linear ordering, so kinship2 draws them twice
(dashed line connecting the two occurrences) to avoid a long crossing line.
Individual `113` has no mate and no children, so kinship2's own plot silently
omits it ("Did not plot the following people: 113").

```{r fam1-data}
fam1 <- sample.ped[sample.ped$ped == 1, ]
nprcPed1 <- to_nprc_ped(fam1$id, fam1$father, fam1$mother, fam1$sex)
nrow(nprcPed1)
any(is.na(nprcPed1$gen))
```

### kinship2

```{r fam1-kinship2}
#| fig-width: 11
#| fig-height: 7
ped1 <- kinship2::pedigree(
  id = fam1$id, dadid = fam1$father, momid = fam1$mother, sex = fam1$sex
)
plot(ped1)
title(main = "kinship2::plot.pedigree() -- sample.ped family 1")
```

### nprcgenekeepr

```{r fam1-nprc}
layout1 <- makePedigreeMatingLayout(nprcPed1)
cat(
  "nodes:", nrow(layout1$nodes),
  " edges:", nrow(layout1$edges),
  " duplicate occurrences:", length(layout1$duplicateToReal), "\n"
)
render_nprc_diagram(nprcPed1)
```

**Observation:** `nprcgenekeepr`'s current duplicate mechanism is narrower in
scope than kinship2's. `nprcgenekeepr` only duplicates an individual who has
*multiple mating units of their own* (a real multi-mate case); it has no
mechanism to duplicate a single-mate individual purely to shorten a long
layout-crossing edge the way kinship2 does for `103`/`138` here. That is why
`nprcgenekeepr` produces **zero** duplicate nodes for this family (confirmed
above), even though kinship2 produces two. Individual `113` still gets its own
node in `nprcgenekeepr`'s output (never silently dropped), unlike kinship2's
own behavior.

**The founder-positioning defect found in Example 1 recurred here too when
this document was first written, confirming it was a systematic pattern, not
a one-off quirk of that synthetic example -- both are now fixed.** `117` is a
founder (no recorded parents) who mates with `116` (a *daughter* of
`105`x`106`, generation 1) -- the identical shape as `203`/`204` above:

```{r fam1-defect-evidence}
e1 <- layout1$edges
into_union <- e1[grepl("^__union_", e1$to), ]
u <- into_union$to[into_union$from == "116"]
n1 <- layout1$nodes[layout1$nodes$id %in% c("105", "106", "117", "116", u),
                     c("id", "x", "y")]
n1[order(n1$y, n1$x), ]
```

`117` now sits at `y = 150`, alongside `__union_9` and `116` -- his own
actual mate union -- not at `y = 0` with `105`/`106` at all. Originally,
`117` (a founder, own generation 0) landed at the SAME row as `105`/`106`,
physically between `106` and `105`x`106`'s own mate connector -- the same
illusion as `201`/`203` in Example 1, suggesting `105` and `117` (or `106`
and `117`) were connected when neither was.

**Root cause and fix:** `.positionMatingUnitForest()` places a mating
unit's row at the generation of its *tree-native* parent (the one whose own
subtree the recursion is already walking -- here, `204`/`116`, each a real
child elsewhere in the tree). It previously did not correspondingly move the
*other*, marry-in parent (`203`/`117`, a founder with no tree-native
position of their own) down to that same row -- the marry-in parent kept
their own, earlier generation row, where they could land inside an unrelated
couple's mate-line span. This was independent of the #142 edge-routing style
question below -- a defect in the underlying x/y coordinate ASSIGNMENT, not
in how edges are drawn between already-placed nodes -- so it remained a
problem even for a fully rectilinear implementation of #142 until fixed
directly. **Fixed via two issues, split by anchor role:**
[#143](https://github.com/rmsharp/nprcgenekeepr/issues/143) (S472,
2026-08-04) covers a marry-in parent who is NOT their mating unit's anchor
(the `203`/`117` shape above); [#144](https://github.com/rmsharp/nprcgenekeepr/issues/144)
(S474, 2026-08-04) covers the anchor-side case the #143 fix did not reach. A
founder marrying into a family at a generation other than their own is an
ordinary, common case in real breeding-colony pedigrees, not a
synthetic-data curiosity -- both fixes were verified against realistic
dangling-parent and multi-generation data, not just this document's two
synthetic families (see `docs/planning/issue143-founder-positioning-fix-plan.md`
and `docs/planning/issue144-anchor-row-mismatch-fix-plan.md`).

**Separately, not a defect:** every node in every rendering above uses the
same solid fill color, regardless of any affected-status value. This is not
"everyone is marked affected" -- `nprcgenekeepr`'s diagram has no
affected-status encoding at all today (tracked: issue #133); the uniform
color is simply vis.js's own default node styling with nothing overriding it.
kinship2's convention (unfilled = normal, filled = affected) has no analog
here yet, so this document's kinship2 renderings using `affected =` (Example
3 below) will show filled/unfilled contrast that the corresponding
`nprcgenekeepr` rendering cannot -- by omission, not by error.

## Example 3: 16-person genotype-annotated pedigree (a real multi-mate case)

The rpubs page's third example (originally following a separate
"Pedigree-drawing with R and graphviz" article) -- individual `2` has TWO
mates (`3` and `10`), a genuine multi-mate case that exercises
`nprcgenekeepr`'s own duplicate-node mechanism directly. This is the raw data
table as printed on the reference page (transcribed verbatim, including its
`aff`/genotype columns used only for the kinship2 reference plot below --
`nprcgenekeepr`'s current Diagram tab has no analog for affected-status or
genotype-label annotation; see issues
[#133](https://github.com/rmsharp/nprcgenekeepr/issues/133) (affected/
phenotype/genotype status) and the Marker Genetics tab (a separate, non-diagram
feature) for genotype data specifically).

```{r geno-data}
geno <- data.frame(
  id = 1:16,
  fid = c(2, 0, 0, 2, 2, 2, 2, 0, 8, 0, 2, 2, 0, 13, 0, 15),
  mid = c(3, 0, 0, 3, 3, 3, 3, 0, 4, 0, 10, 10, 0, 11, 0, 12),
  sex = c(2, 1, 2, 2, 2, 1, 2, 1, 1, 2, 2, 2, 1, 1, 1, 2),
  aff = c(2, 1, 2, 2, 1, 1, 1, 1, 1, 1, 1, 2, 1, 1, 1, 1),
  GABRB1 = c(
    "7/7", "-/-", "7/9", "7/9", "7/7", "7/7", "7/7", "-/-",
    "7/9", "-/-", "7/7", "6/7", "-/-", "7/8", "-/-", "6/6"
  ),
  D4S1645 = c(
    "7/10", "-/-", "3/10", "3/7", "7/10", "7/10", "7/10", "-/-",
    "3/10", "-/-", "7/7", "7/7", "-/-", "7/8", "-/-", "7/7"
  ),
  stringsAsFactors = FALSE
)
geno
nprcPed3 <- to_nprc_ped(geno$id, geno$fid, geno$mid, geno$sex)
nprcPed3
```

### kinship2 (with its own affected-status + genotype-label annotation)

```{r geno-kinship2}
#| fig-width: 10
#| fig-height: 6
ped3 <- kinship2::pedigree(
  id = geno$id, dadid = geno$fid, momid = geno$mid, sex = geno$sex,
  affected = geno$aff
)
strid <- paste(geno$id, geno$GABRB1, geno$D4S1645, sep = "\n")
par(xpd = TRUE)
plot(ped3, id = strid)
title(main = "kinship2::plot.pedigree() -- genotype-annotated pedigree")
```

### nprcgenekeepr

```{r geno-nprc}
layout3 <- makePedigreeMatingLayout(nprcPed3)
cat(
  "nodes:", nrow(layout3$nodes),
  " edges:", nrow(layout3$edges),
  " duplicate occurrences:", length(layout3$duplicateToReal), "\n"
)
layout3$duplicateToReal
render_nprc_diagram(nprcPed3)
```

**Observation:** for this genuine multi-mate case (individual `2`, mates `3`
and `10`), kinship2 draws `2` **once**, with two separate mate-lines (to `3`
and to `10`) meeting at the same node. `nprcgenekeepr`'s current algorithm
instead **duplicates** `2` (one occurrence anchoring the `2`x`3` mating unit,
a second `__dup_2_1` occurrence anchoring the `2`x`10` mating unit, connected
by a dashed line) -- this is `nprcgenekeepr`'s existing, already-shipped
duplicate-node convention working as designed, just visibly different from
kinship2's own single-node-two-mate-lines approach.

## Example 4: multi-mate "crowding" -- does kinship2 place sires left of dams?

[Issue #145](https://github.com/rmsharp/nprcgenekeepr/issues/145) asks
whether the Pedigree Diagram should place a sire to the left of a dam,
citing "standard genetic counseling conventions." A dedicated verification
spike (S482, 2026-08-08 --
`docs/research/issue-145-kinship2-sire-dam-placement-spike-2026-08-08.md`)
read kinship2's own alignment source directly and found **no such rule
anywhere in the package**: the ordinary direct-pedigree spouse-pairing code
path has zero `sex` check (unlike two other, narrower branches that do), and
the apparent "sire-left" result for a simple single-mate pair (Examples 1-3
above all happen to show this) is an indexing artifact, not a rule -- one
that stops holding the moment an individual has more than one mate. This
example reproduces that spike's decisive counter-example directly, side by
side with `nprcgenekeepr`'s own handling of the same data: a dam mated to
two sires (deliberately the mirror image of Example 3's mother-and-daughter
mate-count asymmetry above, this time crowding the FEMALE parent).

```{r crowd-data}
crowd <- data.frame(
  id   = c("S1", "D1", "S2", "K1", "K2"),
  sire = c(NA,   NA,   NA,   "S1", "S2"),
  dam  = c(NA,   NA,   NA,   "D1", "D1"),
  sex  = c("M",  "F",  "M",  "M",  "F"),
  stringsAsFactors = FALSE
)
crowd$gen <- findGeneration(crowd$id, crowd$sire, crowd$dam)
crowd
```

### kinship2

```{r crowd-kinship2}
#| fig-width: 6
#| fig-height: 4
pedCrowd <- kinship2::pedigree(
  id = crowd$id,
  dadid = ifelse(is.na(crowd$sire), "0", crowd$sire),
  momid = ifelse(is.na(crowd$dam), "0", crowd$dam),
  sex = ifelse(crowd$sex == "M", 1, 2),
  missid = "0"
)
plot(pedCrowd)
title(main = "kinship2::plot.pedigree() -- 1 dam, 2 sires (role-reversed crowding)")
```

```{r crowd-kinship2-order}
al <- kinship2::align.pedigree(pedCrowd)
pedCrowd$id[al$nid[1L, al$nid[1L, ] > 0]]
```

kinship2's own default (no `hints` supplied) places the founders' row, left
to right, as **`S1`, `D1`, `S2`** -- the dam **centered**, with her two sires
flanking her on either side, split by the order their mating pairs were
discovered while scanning the pedigree (`S1`x`D1` before `S2`x`D1`), not by
sex. Nothing about this placement is sex-driven: swapping which sire mated
first in the input data swaps which side each sire lands on, with no code
path anywhere in kinship2 inspecting `sex` to decide the split (see the
research doc's Method/Finding sections for the source-level argument this
example only demonstrates empirically).

### nprcgenekeepr

```{r crowd-nprc}
layoutCrowd <- makePedigreeMatingLayout(crowd)
cat(
  "nodes:", nrow(layoutCrowd$nodes),
  " edges:", nrow(layoutCrowd$edges),
  " duplicate occurrences:", length(layoutCrowd$duplicateToReal), "\n"
)
layoutCrowd$duplicateToReal
render_nprc_diagram(crowd)
```

**Observation:** `nprcgenekeepr` takes a structurally different approach to
this same "one parent, two mates" shape than kinship2's centered-single-node
split. Rather than keeping `D1` as one node with two mate-lines,
`nprcgenekeepr` **duplicates** her (`__dup_D1_1`) -- the same
already-shipped convention Example 3 demonstrated for a multi-mate father,
now confirmed to apply identically regardless of which parent's sex has the
extra mate. Neither `nprcgenekeepr`'s existing duplicate-node placement nor
kinship2's centered/discovery-order split makes any sex-based left-right
decision for this case -- consistent with the research doc's conclusion that
issue #145's premise (kinship2 enforces sire-left placement) does not hold
once an individual has more than one mate, the exact scenario this example
constructs.

## Summary of factual differences observed

| Aspect | kinship2 (reference) | nprcgenekeepr (current, shipped) |
|---|---|---|
| Sex shapes | square = male, circle = female (this dataset only exercises 2 of kinship2's sex codes) | dot = Female, square = Male, star = Hermaphrodite, triangle = Unknown, diamond = Other/Unrecorded (5 shapes) |
| Mate connection | horizontal line directly between the two mates, at their shared row | a small, separate "mating-unit" node (a dot), connected to each parent by its own edge |
| Parent(s) -> children edge | vertical drop from the mate-line's midpoint to a horizontal sibship bar, then vertical drops to each child -- strict right angles throughout | straight (diagonal, unless coordinates happen to align) edges from each parent to the mating-unit node, and from the mating-unit node to each child |
| Generations | strict horizontal rows | rows by generation number, but the diagonal edges make row alignment visually less obvious |
| Individual with multiple mates | drawn once, with one mate-line per mate | duplicated once per mating unit beyond the first (anchor), connected by a dashed line |
| Individual needed twice for layout/crossing reasons only (single mate, distant birth family) | duplicated, dashed line | not duplicated -- no mechanism for this case currently exists |
| Isolated individual (no mate, no children) | silently omitted from the plot | still rendered as its own node |
| Affected/phenotype status | filled shape | not supported on the diagram (tracked: issue #133) |
| Twin/zygosity encoding | a relationship-code triangle connector | not supported on the diagram (tracked: issue #137) |
| Genotype labels on nodes | supported via a custom `id=` label string | not supported on the diagram (a separate Marker Genetics tab covers genotype-based analysis, not diagram annotation) |
| Founder who mates into a later generation (e.g. `203`x`204`, `117`x`116`) | positioned adjacent to their actual mate's row | positioned adjacent to their actual mate's row -- **matches kinship2, fixed via issues #143/#144** (originally a confirmed defect in both example families above; see Examples 1-2) |
| Sire/dam left-right ordering for a single-mate pair | sire tends to render left of dam -- an indexing artifact of `dad`/`mom` slot order, not a sex-based rule (no `hints` case) | sire tends to render left of dam -- same underlying discovery-order mechanism, coincidentally similar-looking result (neither package has a sex-based rule; see Example 4 / issue #145) |
| Sire/dam left-right ordering once an individual has multiple mates ("crowding") | the multi-mated parent is centered, mates split by pedigree data-row discovery order -- **not sex** (Example 4's decisive counter-example: `S1, D1, S2`) | the multi-mated parent is duplicated (existing convention, Example 3/4) rather than centered with a split -- also not sex-based |

**Several different kinds of gap/question are visible here, and they should
not be conflated:**

1. **A styling/convention gap (issue #142's scope, now shipped):** kinship2's
   strict horizontal-then-vertical mate-line/sibship-bar routing versus
   `nprcgenekeepr`'s direct diagonal edges to/from a small connector node.
   This was a "how do already-correctly-placed nodes get connected"
   question -- resolved by issue #142's `edgeStyle = "rectilinear"` option,
   which reproduces kinship2's own right-angle convention as an alternative
   to the (still-default) direct style shown in Examples 1-4 above.
2. **A positioning defect, found while preparing this document and now
   fixed (issues #143/#144):** a founder who mated into a later generation
   used to be placed at the wrong row entirely, which could make the diagram
   visually imply an incorrect mating pair. This was a "where do nodes get
   placed in the first place" question -- independent of #142's edge-routing
   question, and it reproduced in both of kinship2's own reference families,
   not just a contrived case. See Examples 1-2 above for the current,
   corrected behavior.
3. **A sire/dam left-right ordering question (issue #145's premise,
   investigated and found unsupported -- S482/Example 4 above):** issue #145
   asks `nprcgenekeepr` to enforce sire-left-of-dam placement, citing
   "standard genetic counseling conventions." Neither package actually
   implements a sex-based left-right rule -- both packages' apparent
   "sire-left" result for a simple single-mate pair is an artifact of
   internal parent-slot/discovery order, not a rule, and Example 4's dam-
   with-two-sires counter-example shows kinship2 abandoning even that
   appearance the moment an individual has more than one mate. Any future
   #145 design work should treat "male typically left" as a default worth
   choosing for the simple case, not as behavior either package already
   enforces and `nprcgenekeepr` merely needs to "correct."
4. **The duplicate-individual differences** (narrower trigger condition than
   kinship2's; no single-node-multiple-mate-line support) are a fourth,
   related but analytically separate question -- see Examples 2-4.

No resolution is recommended for any of these in this document -- it exists
to inform decisions on each, not make them.
