---
title: "Pedigree Diagram: kinship2 Reference Comparison"
subtitle: "Decision-support research for issue #142 (rectilinear mate-line/sibship-bar style) -- not a package vignette, not shipped"
date: today
format:
  html:
    embed-resources: true
    toc: true
    df-print: kable
---

## Purpose

The owner asked, after seeing `nprcgenekeepr`'s live Pedigree Diagram tab
render a real family, whether the drawing is right -- pointing at
<https://rpubs.com/dliupress/pedigreedemo> as the reference for what a
"correct" pedigree drawing looks like. That page uses the **kinship2** R
package (a real, well-known pedigree-drawing package, not part of
`nprcgenekeepr`'s own dependencies) to plot a small set of example pedigrees.

This document reproduces those *exact same* example pedigrees -- same
individuals, same parent/child/sex relationships -- using
`nprcgenekeepr`'s own current pedigree-diagram capability
(`makePedigreeMatingLayout()` + `visNetwork`, the identical render chain
`R/modPedigree.R`'s Diagram tab uses today), placed directly next to
kinship2's own rendering of the identical data. The goal is to give the owner
a fair, apples-to-apples visual and factual basis for deciding whether (and
how) to pursue [issue #142](https://github.com/rmsharp/nprcgenekeepr/issues/142)
-- **no decision is made in this document, and no package code is changed.**

`kinship2` is installed only into this project's local `renv` library for
this one-off comparison (not added to `DESCRIPTION`, not snapshotted into
`renv.lock`) -- it is reference material, not a dependency.

## Setup

```{r setup}
#| message: false
suppressWarnings(suppressMessages(library(kinship2)))
library(nprcgenekeepr)
library(visNetwork)
```

### Helper: convert a kinship2-style data frame to nprcgenekeepr's pedigree format

kinship2's own examples use `id`/`father`/`mother`/`sex` with `0` meaning "no
parent" and `sex` coded `1` = male, `2` = female. `nprcgenekeepr` uses
`id`/`sire`/`dam`/`sex` (`"M"`/`"F"`/`"H"`/`"U"`, `NA` for no parent) plus a
`gen` column from its own exported `findGeneration()`. `convertSexCodes()` --
already exported by `nprcgenekeepr` -- converts kinship2's `1`/`2` codes
directly.

```{r helper-convert}
to_nprc_ped <- function(id, father, mother, sex) {
  ped <- data.frame(
    id = as.character(id),
    sire = ifelse(father == 0, NA_character_, as.character(father)),
    dam = ifelse(mother == 0, NA_character_, as.character(mother)),
    sex = as.character(convertSexCodes(sex)),
    stringsAsFactors = FALSE
  )
  ped$gen <- findGeneration(ped$id, ped$sire, ped$dam)
  ped
}
```

### Helper: render via nprcgenekeepr, matching `R/modPedigree.R`'s exact current chain

```{r helper-render}
render_nprc_diagram <- function(ped) {
  layout <- makePedigreeMatingLayout(ped)
  visNetwork::visNetwork(layout$nodes, layout$edges) |>
    visNetwork::visPhysics(enabled = FALSE) |>
    visNetwork::visNodes(physics = FALSE) |>
    visNetwork::visEdges(smooth = FALSE) |>
    visNetwork::visLegend(
      addNodes = data.frame(
        label = c(
          "Female", "Male", "Hermaphrodite", "Unknown", "Other / Unrecorded"
        ),
        shape = c("dot", "square", "star", "triangle", "diamond"),
        stringsAsFactors = FALSE
      ),
      useGroups = FALSE, position = "right", main = "Sex",
      width = 0.28, stepY = 65L
    ) |>
    visNetwork::visOptions(
      nodesIdSelection = list(
        enabled = TRUE,
        values = layout$nodes$id[
          !grepl("^__union_|^__dup_", layout$nodes$id)
        ]
      ),
      highlightNearest = list(enabled = TRUE, hover = TRUE, degree = 1L,
                               algorithm = "all")
    )
}
```

## How to read each example

For each pedigree below:

1. **kinship2's own rendering** (`plot.pedigree()`) -- the reference
   convention: mates joined by a horizontal line at their own level; from its
   midpoint, a vertical line drops to a horizontal sibship bar; each child
   rises to meet that bar with its own vertical line. Strict right angles,
   generations in clean rows.
2. **nprcgenekeepr's current rendering** -- the exact code path
   `R/modPedigree.R`'s Diagram tab runs today, fed the *identical* data.

## Example 1: `sample.ped` family 2 (14 people, no multi-mate individuals)

The cleanest comparison -- a single 3-generation family tree with no
duplicated/multi-mate individuals, so the ONLY variable being compared is the
mate-line/sibship-bar edge-routing convention itself.

```{r fam2-data}
data(sample.ped)
fam2 <- sample.ped[sample.ped$ped == 2, ]
nprcPed2 <- to_nprc_ped(fam2$id, fam2$father, fam2$mother, fam2$sex)
nprcPed2
```

### kinship2

```{r fam2-kinship2}
#| fig-width: 9
#| fig-height: 5
ped2 <- kinship2::pedigree(
  id = fam2$id, dadid = fam2$father, momid = fam2$mother, sex = fam2$sex
)
plot(ped2)
title(main = "kinship2::plot.pedigree() -- sample.ped family 2")
```

### nprcgenekeepr

```{r fam2-nprc}
render_nprc_diagram(nprcPed2)
```

**Finding -- a founder-positioning defect (distinct from the edge-routing
question below):** in this rendering, `203` (a founder with no recorded
parents, who mates with `204`, a *daughter* of `201`x`202`, generation 1)
visually reads as if `201` and `203` -- two males -- are the mated pair. This
is confirmed in the actual layout data, not just an impression:

```{r fam2-defect-evidence}
layout2 <- makePedigreeMatingLayout(nprcPed2)
n2 <- layout2$nodes[layout2$nodes$y == 0, c("id", "x", "y")]
n2[order(n2$x), ]
```

At `y = 0` (the founders' own row), left to right: `202` (x = -360), `203`
(x = -240), `__union_1` (x = -180, the true midpoint of `201`x`202`'s own
mate-line), `201` (x = 0). `203` sits physically *inside* the visual span of
`201`x`202`'s own mate connector, even though `203` has no relationship to
either of them -- his real mate union (with `204`) is a full generation
row below (`y = 150`), reached by one long edge that crosses back up through
this row. Root cause and a second, independent confirmation (family 1, ids
`116`/`117`) are discussed after Example 2 below.

## Example 2: `sample.ped` family 1 (41 people, includes real multi-mate individuals)

A larger, more complex family. kinship2's own layout duplicates 2 individuals
here (ids `103` and `138`) purely for layout/crossing-minimization reasons --
each has only ONE mate, but their own birth family is drawn far from their
marital family in kinship2's linear ordering, so kinship2 draws them twice
(dashed line connecting the two occurrences) to avoid a long crossing line.
Individual `113` has no mate and no children, so kinship2's own plot silently
omits it ("Did not plot the following people: 113").

```{r fam1-data}
fam1 <- sample.ped[sample.ped$ped == 1, ]
nprcPed1 <- to_nprc_ped(fam1$id, fam1$father, fam1$mother, fam1$sex)
nrow(nprcPed1)
any(is.na(nprcPed1$gen))
```

### kinship2

```{r fam1-kinship2}
#| fig-width: 11
#| fig-height: 7
ped1 <- kinship2::pedigree(
  id = fam1$id, dadid = fam1$father, momid = fam1$mother, sex = fam1$sex
)
plot(ped1)
title(main = "kinship2::plot.pedigree() -- sample.ped family 1")
```

### nprcgenekeepr

```{r fam1-nprc}
layout1 <- makePedigreeMatingLayout(nprcPed1)
cat(
  "nodes:", nrow(layout1$nodes),
  " edges:", nrow(layout1$edges),
  " duplicate occurrences:", length(layout1$duplicateToReal), "\n"
)
render_nprc_diagram(nprcPed1)
```

**Observation:** `nprcgenekeepr`'s current duplicate mechanism is narrower in
scope than kinship2's. `nprcgenekeepr` only duplicates an individual who has
*multiple mating units of their own* (a real multi-mate case); it has no
mechanism to duplicate a single-mate individual purely to shorten a long
layout-crossing edge the way kinship2 does for `103`/`138` here. That is why
`nprcgenekeepr` produces **zero** duplicate nodes for this family (confirmed
above), even though kinship2 produces two. Individual `113` still gets its own
node in `nprcgenekeepr`'s output (never silently dropped), unlike kinship2's
own behavior.

**The founder-positioning defect found in Example 1 recurs here too**,
confirming it is a systematic pattern, not a one-off quirk of that synthetic
example. `117` is a founder (no recorded parents) who mates with `116`
(a *daughter* of `105`x`106`, generation 1) -- the identical shape as
`203`/`204` above:

```{r fam1-defect-evidence}
e1 <- layout1$edges
into_union <- e1[grepl("^__union_", e1$to), ]
u <- into_union$to[into_union$from == "116"]
n1 <- layout1$nodes[layout1$nodes$y == 0, c("id", "x", "y")]
n1[n1$id %in% c("105", "106", "117", u), ][order(n1[n1$id %in% c("105", "106", "117", u), "x"]), ]
```

`117` (a founder, own generation 0) is placed at the SAME row as `105`/`106`,
landing physically between `106` and `105`x`106`'s own mate connector --
exactly the same illusion as `201`/`203` in Example 1, this time suggesting
`105` and `117` (or `106` and `117`) are connected when neither is. `117`'s
actual mate union (with `116`) sits a full row below, at `y = 150`.

**Root cause (not fixed here -- flagged for a future session, not part of
this document's scope):** `.positionMatingUnitForest()` places a mating
unit's row at the generation of its *tree-native* parent (the one whose own
subtree the recursion is already walking -- here, `204`/`116`, each a real
child elsewhere in the tree) but does not correspondingly move the *other*,
marry-in parent (`203`/`117`, a founder with no tree-native position of their
own) down to that same row. The marry-in parent keeps their own, earlier
generation row, where they can land inside an unrelated couple's mate-line
span. **This is independent of the #142 edge-routing style question below --
it is a defect in the underlying x/y coordinate ASSIGNMENT, not in how edges
are drawn between already-placed nodes**, so it would remain a problem even
under a fully rectilinear implementation of #142. A founder marrying into a
family at a generation other than their own is an ordinary, common case in
real breeding-colony pedigrees, not a synthetic-data curiosity -- this
warrants its own investigation regardless of the #142 decision.

**Separately, not a defect:** every node in every rendering above uses the
same solid fill color, regardless of any affected-status value. This is not
"everyone is marked affected" -- `nprcgenekeepr`'s diagram has no
affected-status encoding at all today (tracked: issue #133); the uniform
color is simply vis.js's own default node styling with nothing overriding it.
kinship2's convention (unfilled = normal, filled = affected) has no analog
here yet, so this document's kinship2 renderings using `affected =` (Example
3 below) will show filled/unfilled contrast that the corresponding
`nprcgenekeepr` rendering cannot -- by omission, not by error.

## Example 3: 16-person genotype-annotated pedigree (a real multi-mate case)

The rpubs page's third example (originally following a separate
"Pedigree-drawing with R and graphviz" article) -- individual `2` has TWO
mates (`3` and `10`), a genuine multi-mate case that exercises
`nprcgenekeepr`'s own duplicate-node mechanism directly. This is the raw data
table as printed on the reference page (transcribed verbatim, including its
`aff`/genotype columns used only for the kinship2 reference plot below --
`nprcgenekeepr`'s current Diagram tab has no analog for affected-status or
genotype-label annotation; see issues
[#133](https://github.com/rmsharp/nprcgenekeepr/issues/133) (affected/
phenotype/genotype status) and the Marker Genetics tab (a separate, non-diagram
feature) for genotype data specifically).

```{r geno-data}
geno <- data.frame(
  id = 1:16,
  fid = c(2, 0, 0, 2, 2, 2, 2, 0, 8, 0, 2, 2, 0, 13, 0, 15),
  mid = c(3, 0, 0, 3, 3, 3, 3, 0, 4, 0, 10, 10, 0, 11, 0, 12),
  sex = c(2, 1, 2, 2, 2, 1, 2, 1, 1, 2, 2, 2, 1, 1, 1, 2),
  aff = c(2, 1, 2, 2, 1, 1, 1, 1, 1, 1, 1, 2, 1, 1, 1, 1),
  GABRB1 = c(
    "7/7", "-/-", "7/9", "7/9", "7/7", "7/7", "7/7", "-/-",
    "7/9", "-/-", "7/7", "6/7", "-/-", "7/8", "-/-", "6/6"
  ),
  D4S1645 = c(
    "7/10", "-/-", "3/10", "3/7", "7/10", "7/10", "7/10", "-/-",
    "3/10", "-/-", "7/7", "7/7", "-/-", "7/8", "-/-", "7/7"
  ),
  stringsAsFactors = FALSE
)
geno
nprcPed3 <- to_nprc_ped(geno$id, geno$fid, geno$mid, geno$sex)
nprcPed3
```

### kinship2 (with its own affected-status + genotype-label annotation)

```{r geno-kinship2}
#| fig-width: 10
#| fig-height: 6
ped3 <- kinship2::pedigree(
  id = geno$id, dadid = geno$fid, momid = geno$mid, sex = geno$sex,
  affected = geno$aff
)
strid <- paste(geno$id, geno$GABRB1, geno$D4S1645, sep = "\n")
par(xpd = TRUE)
plot(ped3, id = strid)
title(main = "kinship2::plot.pedigree() -- genotype-annotated pedigree")
```

### nprcgenekeepr

```{r geno-nprc}
layout3 <- makePedigreeMatingLayout(nprcPed3)
cat(
  "nodes:", nrow(layout3$nodes),
  " edges:", nrow(layout3$edges),
  " duplicate occurrences:", length(layout3$duplicateToReal), "\n"
)
layout3$duplicateToReal
render_nprc_diagram(nprcPed3)
```

**Observation:** for this genuine multi-mate case (individual `2`, mates `3`
and `10`), kinship2 draws `2` **once**, with two separate mate-lines (to `3`
and to `10`) meeting at the same node. `nprcgenekeepr`'s current algorithm
instead **duplicates** `2` (one occurrence anchoring the `2`x`3` mating unit,
a second `__dup_2_1` occurrence anchoring the `2`x`10` mating unit, connected
by a dashed line) -- this is `nprcgenekeepr`'s existing, already-shipped
duplicate-node convention working as designed, just visibly different from
kinship2's own single-node-two-mate-lines approach.

## Summary of factual differences observed

| Aspect | kinship2 (reference) | nprcgenekeepr (current, shipped) |
|---|---|---|
| Sex shapes | square = male, circle = female (this dataset only exercises 2 of kinship2's sex codes) | dot = Female, square = Male, star = Hermaphrodite, triangle = Unknown, diamond = Other/Unrecorded (5 shapes) |
| Mate connection | horizontal line directly between the two mates, at their shared row | a small, separate "mating-unit" node (a dot), connected to each parent by its own edge |
| Parent(s) -> children edge | vertical drop from the mate-line's midpoint to a horizontal sibship bar, then vertical drops to each child -- strict right angles throughout | straight (diagonal, unless coordinates happen to align) edges from each parent to the mating-unit node, and from the mating-unit node to each child |
| Generations | strict horizontal rows | rows by generation number, but the diagonal edges make row alignment visually less obvious |
| Individual with multiple mates | drawn once, with one mate-line per mate | duplicated once per mating unit beyond the first (anchor), connected by a dashed line |
| Individual needed twice for layout/crossing reasons only (single mate, distant birth family) | duplicated, dashed line | not duplicated -- no mechanism for this case currently exists |
| Isolated individual (no mate, no children) | silently omitted from the plot | still rendered as its own node |
| Affected/phenotype status | filled shape | not supported on the diagram (tracked: issue #133) |
| Twin/zygosity encoding | a relationship-code triangle connector | not supported on the diagram (tracked: issue #137) |
| Genotype labels on nodes | supported via a custom `id=` label string | not supported on the diagram (a separate Marker Genetics tab covers genotype-based analysis, not diagram annotation) |
| Founder who mates into a later generation (e.g. `203`x`204`, `117`x`116`) | positioned adjacent to their actual mate's row | positioned at their OWN generation row -- can land inside an unrelated couple's mate-line, visually implying the wrong pairing (**confirmed defect**, both example families above) |

**Two different kinds of gap are visible here, and they should not be
conflated:**

1. **A styling/convention gap (issue #142's scope):** kinship2's strict
   horizontal-then-vertical mate-line/sibship-bar routing versus
   `nprcgenekeepr`'s current direct diagonal edges to/from a small connector
   node. This is a "how do already-correctly-placed nodes get connected"
   question.
2. **A confirmed positioning defect (independent of #142, found while
   preparing this document):** a founder who mates into a later generation
   is placed at the wrong row entirely, which can make the diagram visually
   imply an incorrect mating pair. This is a "where do nodes get placed in
   the first place" question -- it exists regardless of how the edges
   connecting them are eventually drawn, and reproduces in both of
   kinship2's own reference families, not just a contrived case.

The duplicate-individual differences (narrower trigger condition than
kinship2's; no single-node-multiple-mate-line support) are a third, related
but analytically separate question. No resolution is recommended for any of
the three in this document -- it exists to inform that decision, not make it.
