---
title: "Pedigree Diagram: Sibling Subtree-Width Asymmetry -- Evidence"
subtitle: "Supporting evidence for docs/planning/pedigree-diagram-sibling-subtree-width-plan.md -- not a package vignette, not shipped"
date: today
format:
  html:
    embed-resources: true
    toc: true
    df-print: kable
---

## Purpose

`BACKLOG.md` (found S576, incidental to Track 6's own empirical validation) flags that 2-3 direct
children of the same mating unit can land far apart in `x` purely because their own
descendant-subtree sizes differ -- independent of where the union itself is positioned. This
document builds a small, self-contained reproduction of that mechanism, renders it via
**kinship2** (the reference convention, matching the project's own
`pedigree-diagram-kinship2-reference-comparison.qmd` precedent) and via `nprcgenekeepr`'s current
shipped algorithm side by side, then empirically tests one candidate mitigation (a bounded-depth
contour-merge lookahead) against both this small example and the real 375-individual bundled
fixture. **No package code is changed here** -- this is investigation feeding the design document's
own Decision section.

`kinship2` is installed only into this project's local `renv` library for this one-off comparison
(not added to `DESCRIPTION`, not snapshotted into `renv.lock`) -- it is reference material, not a
dependency, matching the established precedent.

## Setup

```{r setup}
#| message: false
suppressWarnings(suppressMessages(library(kinship2)))
library(nprcgenekeepr)
```

## The reproduction

A minimal pedigree isolating the mechanism: `P1`x`P2` have two children. `A` is a childless leaf.
`B` anchors a 2-generation-deep, 2-children-per-union subtree (`B`x`BM` -> `B1`,`B2`; `B1`x`B1M` ->
`B1a`,`B1b`; `B2`x`B2M` -> `B2a`,`B2b`) -- 13 individuals total.

```{r build-ped}
build_ped <- function(rows) {
  ped <- do.call(rbind, lapply(rows, as.data.frame, stringsAsFactors = FALSE))
  ped$gen <- findGeneration(ped$id, ped$sire, ped$dam)
  ped
}
r <- function(id, sire = NA, dam = NA, sex) list(id = id, sire = sire, dam = dam, sex = sex)

rows <- list(
  r("P1", sex = "M"), r("P2", sex = "F"),
  r("A", "P1", "P2", "M"), r("B", "P1", "P2", "F"),
  r("BM", sex = "M"),
  r("B1", "BM", "B", "M"), r("B2", "BM", "B", "F"),
  r("B1M", sex = "F"), r("B2M", sex = "M"),
  r("B1a", "B1", "B1M", "F"), r("B1b", "B1", "B1M", "M"),
  r("B2a", "B2M", "B2", "F"), r("B2b", "B2M", "B2", "M")
)
ped <- build_ped(rows)
ped
```

## kinship2's own rendering

```{r kinship2-render}
#| fig-width: 9
#| fig-height: 5.5
sireCode <- ifelse(is.na(ped$sire), "0", ped$sire)
damCode  <- ifelse(is.na(ped$dam), "0", ped$dam)
sexCode  <- ifelse(ped$sex == "M", 1, 2)
pedK <- kinship2::pedigree(id = ped$id, dadid = sireCode, momid = damCode,
                            sex = sexCode, missid = "0")
plot(pedK)
title(main = "kinship2::plot.pedigree() -- sibling subtree-width asymmetry example")
```

`A` renders compact, close to the rest of the family, despite `B`'s much wider subtree.

## nprcgenekeepr's current (shipped) rendering

```{r nprc-render}
#| fig-width: 9
#| fig-height: 5.5
layout <- makePedigreeMatingLayout(ped, edgeStyle = "rectilinear")
nodes <- layout$nodes
edges <- layout$edges
plot(nodes$x, -nodes$y, type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "nprcgenekeepr::makePedigreeMatingLayout() (edgeStyle='rectilinear') -- same data",
     xlim = range(nodes$x) + c(-20, 20))
for (i in seq_len(nrow(edges))) {
  fx <- nodes$x[nodes$id == edges$from[i]]; fy <- -nodes$y[nodes$id == edges$from[i]]
  tx <- nodes$x[nodes$id == edges$to[i]];   ty <- -nodes$y[nodes$id == edges$to[i]]
  if (length(fx) && length(tx)) segments(fx, fy, tx, ty, col = "grey40")
}
realNodes <- nodes[!grepl("^__", nodes$id), ]
points(realNodes$x, -realNodes$y, pch = 1, cex = 2.4)
text(realNodes$x, -realNodes$y, labels = realNodes$id, cex = 0.75, pos = 1, offset = 0.9)

gap <- abs(diff(nodes$x[nodes$id %in% c("A", "B")]))
cat("A x:", nodes$x[nodes$id == "A"], " B x:", nodes$x[nodes$id == "B"], " gap:", gap, "\n")
```

**Confirms the mechanism exactly:** `A`'s own position is unaffected by `B`'s subtree -- `A`
has no descendants to collide with -- yet the union's children-merge step reserves `B`'s full
descendant footprint against `A`'s, pushing `A` far away to avoid an overlap that never occurs at
`A`'s own generation row.

## Candidate mitigation: bounded-depth contour-merge lookahead

The shipped `.positionMatingUnitForest()` (`R/makePedigreeDiagramData.R`) compares two sibling
subtrees' *entire* contour (every generation row, however deep) when deciding how far apart to
place them. One plausible-looking fix: only compare rows within `K` generations of the merge
point, trusting the existing row-based `sweepMinSep()` post-pass (Track 3) to resolve any
resulting same-row crowding at deeper generations. The function below is a modified copy of
`.positionMatingUnitForest()` (investigation-only, not part of the package) with exactly that one
change -- `mergeSubtrees()` gains an `atGen` parameter and its `finite` mask is windowed to
`[atGen, atGen + lookaheadK]`.

```{r candidate-fn}
#| code-fold: true
positionBoundedLookahead <- function(ped, forest, orderBySex = TRUE, lookaheadK = Inf) {
  ped$gen[is.na(ped$gen)] <- 0L
  matingUnits <- forest$matingUnits
  duplicates <- forest$duplicates
  childEdges <- forest$childEdges
  unitIds <- matingUnits$id
  realIds <- as.character(ped$id)
  genOf <- stats::setNames(ped$gen, realIds)
  danglingIds <- setdiff(c(matingUnits$sire, matingUnits$dam), realIds)
  if (length(danglingIds) > 0L) {
    fallbackGen <- vapply(danglingIds, function(x) {
      matingUnits$gen[matingUnits$sire == x | matingUnits$dam == x][1L]
    }, integer(1L))
    genOf <- c(genOf, stats::setNames(fallbackGen, danglingIds))
  }
  unitGenOf <- stats::setNames(matingUnits$gen, unitIds)
  maxGen <- max(ped$gen, if (nrow(matingUnits) > 0L) matingUnits$gen else 0L, na.rm = TRUE)
  if (!is.finite(maxGen)) maxGen <- 0L
  minSep <- 1L

  leafContour <- function(gen) {
    left <- rep(Inf, maxGen + 1L); right <- rep(-Inf, maxGen + 1L)
    left[gen + 1L] <- 0L; right[gen + 1L] <- 0L
    list(x = 0L, contour = list(left = left, right = right))
  }

  ## MODIFIED from the shipped mergeSubtrees(): takes atGen, masks the
  ## finite comparison to rows in [atGen, atGen + lookaheadK] only -- rows
  ## beyond that are treated as "no constraint" for THIS merge decision.
  mergeSubtrees <- function(subResults, atGen) {
    n <- length(subResults)
    xs <- numeric(n)
    contour <- subResults[[1L]]$contour
    rowIdx <- seq_along(contour$right) - 1L
    windowMask <- rowIdx >= atGen & rowIdx <= (atGen + lookaheadK)
    if (n > 1L) {
      for (i in 2L:n) {
        ci <- subResults[[i]]$contour
        finite <- is.finite(contour$right) & is.finite(ci$left) & windowMask
        shift <- minSep
        if (any(finite)) {
          needed <- max(contour$right[finite] - ci$left[finite] + minSep)
          shift <- max(shift, needed)
        }
        xs[i] <- shift
        contour <- list(left = pmin(contour$left, ci$left + shift),
                         right = pmax(contour$right, ci$right + shift))
      }
    }
    list(xs = xs, contour = contour)
  }

  finalizeNode <- function(merged, ownGen) {
    xs <- merged$xs
    ownX <- (xs[1L] + xs[length(xs)]) / 2L
    shiftAmt <- -ownX
    contour <- list(left = merged$contour$left + shiftAmt, right = merged$contour$right + shiftAmt)
    contour$left[ownGen + 1L] <- min(contour$left[ownGen + 1L], 0L)
    contour$right[ownGen + 1L] <- max(contour$right[ownGen + 1L], 0L)
    list(ownX = ownX, childOffsets = xs - ownX, contour = contour)
  }

  anchoredUnits <- matingUnits[!is.na(matingUnits$anchor), , drop = FALSE]
  everAnchor <- unique(anchoredUnits$anchor)
  nonAnchorSides <- c(anchoredUnits$sire, anchoredUnits$dam)
  neverAnchorIds <- setdiff(unique(nonAnchorSides), everAnchor)
  hasOwnDirectChild <- function(id) any(childEdges$from == id & !(childEdges$from %in% unitIds))
  freePassIds <- Filter(function(id) !hasOwnDirectChild(id), neverAnchorIds)
  freePassUnitOf <- stats::setNames(character(length(freePassIds)), freePassIds)
  for (fp in freePassIds) {
    ownUnits <- matingUnits$id[matingUnits$sire == fp | matingUnits$dam == fp]
    dupUnits <- duplicates$matingUnitId[duplicates$realId == fp]
    freePassUnitOf[[fp]] <- setdiff(ownUnits, dupUnits)[1L]
  }
  freePassOfUnit <- split(names(freePassUnitOf), freePassUnitOf)

  dispGenOf <- genOf[realIds]
  realFreePassIds <- intersect(freePassIds, realIds)
  if (length(realFreePassIds) > 0L) {
    dispGenOf[realFreePassIds] <- unname(unitGenOf[freePassUnitOf[realFreePassIds]])
  }

  relNode <- new.env(parent = emptyenv())

  positionUnit <- function(unitId) {
    kidIds <- childEdges$to[childEdges$from == unitId]
    fpHere <- freePassOfUnit[[unitId]]
    subIds <- c(fpHere, kidIds)
    subResults <- lapply(subIds, function(sid) {
      if (!is.null(fpHere) && sid %in% fpHere) leafContour(unitGenOf[[unitId]]) else positionIndividual(sid)
    })
    fin <- finalizeNode(mergeSubtrees(subResults, unitGenOf[[unitId]]), unitGenOf[[unitId]])
    relNode[[unitId]] <- list(childIds = subIds, childOffsets = fin$childOffsets)
    list(x = fin$ownX, contour = fin$contour)
  }

  positionIndividual <- function(id) {
    unitSub <- matingUnits$id[matingUnits$anchor %in% id]
    directSub <- childEdges$to[childEdges$from == id & !(childEdges$from %in% unitIds)]
    subIds <- c(unitSub, directSub)
    if (length(subIds) == 0L) {
      relNode[[id]] <- list(childIds = character(0L), childOffsets = numeric(0L))
      return(leafContour(genOf[[id]]))
    }
    subResults <- lapply(subIds, function(sid) if (sid %in% unitIds) positionUnit(sid) else positionIndividual(sid))
    fin <- finalizeNode(mergeSubtrees(subResults, genOf[[id]]), genOf[[id]])
    relNode[[id]] <- list(childIds = subIds, childOffsets = fin$childOffsets)
    list(x = fin$ownX, contour = fin$contour)
  }

  hasParentEdge <- realIds %in% childEdges$to
  founderIds <- realIds[!hasParentEdge]
  rootIds <- setdiff(founderIds, freePassIds)
  rootIds <- rootIds[order(match(rootIds, realIds))]
  orphanUnitIds <- matingUnits$id[is.na(matingUnits$anchor)]

  rootResults <- c(lapply(rootIds, positionIndividual), lapply(orphanUnitIds, positionUnit))
  allRootIds <- c(rootIds, orphanUnitIds)
  rootMerge <- mergeSubtrees(rootResults, 0L)

  absX <- new.env(parent = emptyenv())
  assignAbs <- function(id, base) {
    absX[[id]] <- base
    node <- relNode[[id]]
    if (!is.null(node) && length(node$childIds) > 0L) {
      for (i in seq_along(node$childIds)) assignAbs(node$childIds[i], base + node$childOffsets[i])
    }
  }
  for (i in seq_along(allRootIds)) assignAbs(allRootIds[i], rootMerge$xs[i])

  realX <- vapply(realIds, function(i) absX[[i]], numeric(1L))

  sweepMinSep <- function(ids, xVals, genVals) {
    x <- stats::setNames(xVals, ids)
    for (g in unique(genVals)) {
      rowIds <- ids[genVals == g]
      if (length(rowIds) < 2L) next
      rowIds <- rowIds[order(x[rowIds], rowIds, method = "radix")]
      for (i in 2L:length(rowIds)) {
        prevX <- x[[rowIds[i - 1L]]]
        if (x[[rowIds[i]]] < prevX + minSep) x[[rowIds[i]]] <- prevX + minSep
      }
    }
    x
  }
  sweepX <- sweepMinSep(realIds, unname(realX), unname(dispGenOf[realIds]))
  realX <- sweepX[realIds]

  data.frame(id = realIds, x = unname(realX), gen = unname(dispGenOf[realIds]),
             stringsAsFactors = FALSE)
}
```

### Candidate result on the small example: `A`-`B` gap by lookahead depth

```{r candidate-small}
forest <- nprcgenekeepr:::.buildMatingUnitForest(ped)
res <- lapply(c(0, 1, 2, Inf), function(K) {
  pos <- positionBoundedLookahead(ped, forest, lookaheadK = K)
  data.frame(K = K, gap_raw = abs(diff(pos$x[pos$id %in% c("A", "B")])))
})
do.call(rbind, res)
```

`K = 0` (the tightest possible bound) closes the `A`-`B` gap from 2.5 raw units (matches the
shipped algorithm, `K = Inf`) down to 1 -- promising in isolation.

### Candidate rendering at `K = 0`

```{r candidate-render}
#| fig-width: 9
#| fig-height: 5.5
pos0 <- positionBoundedLookahead(ped, forest, lookaheadK = 0)
pos0$x <- pos0$x * 120L  # match makePedigreeMatingLayout()'s own xScale
plot(pos0$x, -pos0$gen * 150, type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "CANDIDATE (bounded lookahead K=0) -- same data -- NOT shipped, illustration only",
     xlim = range(pos0$x) + c(-20, 20))
for (i in seq_len(nrow(ped))) {
  cx <- pos0$x[pos0$id == ped$id[i]]; cy <- -pos0$gen[pos0$id == ped$id[i]] * 150
  if (!is.na(ped$sire[i])) {
    sx <- pos0$x[pos0$id == ped$sire[i]]; sy <- -pos0$gen[pos0$id == ped$sire[i]] * 150
    if (length(sx) && length(cx)) segments(sx, sy, cx, cy, col = "grey60", lty = 2)
  }
  if (!is.na(ped$dam[i])) {
    dx <- pos0$x[pos0$id == ped$dam[i]]; dy <- -pos0$gen[pos0$id == ped$dam[i]] * 150
    if (length(dx) && length(cx)) segments(dx, dy, cx, cy, col = "grey60", lty = 2)
  }
}
points(pos0$x, -pos0$gen * 150, pch = 1, cex = 2.4)
text(pos0$x, -pos0$gen * 150, labels = pos0$id, cex = 0.75, pos = 1, offset = 0.9)
```

**The tighter `A`-`B` gap comes at a real cost: `B1`'s and `B2`'s own connecting lines now cross**
(visible in the plot above, in the `B1`/`B2M` and `B2`/`BM` region) -- an edge-crossing defect that
did not exist in the shipped algorithm's rendering above. Edge crossings are a strictly worse
legibility problem than "correct but wide": a reader can no longer trace parentage by following an
uninterrupted line.

### Candidate result on the real 375-individual fixture

```{r candidate-real}
pedReal <- read.csv("../../inst/extdata/examples/obfuscated_rhesus_mhc_ped.csv",
                     stringsAsFactors = FALSE)
pedReal$id <- as.character(pedReal$id)
pedReal$sire <- as.character(pedReal$sire)
pedReal$dam <- as.character(pedReal$dam)
pedReal$gen <- findGeneration(pedReal$id, pedReal$sire, pedReal$dam)
forestReal <- nprcgenekeepr:::.buildMatingUnitForest(pedReal)
childEdges <- forestReal$childEdges
matingUnits <- forestReal$matingUnits

measure <- function(posDf) {
  xOf <- stats::setNames(posDf$x, posDf$id)
  offsets <- c()
  for (u in matingUnits$id) {
    kids <- childEdges$to[childEdges$from == u]
    kidsReal <- intersect(kids, posDf$id)
    if (length(kidsReal) == 0) next
    kx <- xOf[kidsReal]
    unitX <- (min(kx) + max(kx)) / 2
    for (k in kidsReal) offsets <- c(offsets, abs(xOf[[k]] - unitX))
  }
  c(n = length(offsets), pct_over_threshold = mean(offsets > 200 / 120) * 100,
    max_raw_offset = max(offsets))
}

resReal <- lapply(c(0, 1, 2, 3, Inf), function(K) {
  pos <- positionBoundedLookahead(pedReal, forestReal, lookaheadK = K)
  m <- measure(pos)
  data.frame(K = K, n_edges = m[["n"]], pct_over_threshold = round(m[["pct_over_threshold"]], 1),
             max_raw_offset = round(m[["max_raw_offset"]], 2))
})
do.call(rbind, resReal)
```

**Caveat:** this `measure()` proxy is simplified relative to the shipped algorithm's actual final
positions (it stops after the first `sweepMinSep()` pass and does not apply the `orderBySex` swap
or the final de-collision pass), so its absolute percentages should not be compared directly against
Track 6's own published baseline figures (`docs/planning/pedigree-diagram-track6-child-centered-
union-position-plan.md` \S1.4: 9/251, 3.6%, measured with the full pipeline). The trend across `K`
under this *same* simplified methodology is still meaningful: **tightening the lookahead makes the
real-fixture measure worse, not better** (`K = 0`'s `pct_over_threshold` is higher than `K = Inf`'s)
-- the opposite direction from the small example. A full, faithful re-measurement against the
shipped pipeline's exact final positions is deferred to an implementation session, should one ever
be scheduled.

## Conclusion

The bounded-lookahead candidate is **rejected**: it trades a "correct but wide" layout for a
"narrower but crossing" one on the very example constructed to show it working best, and it
regresses (does not improve) the real fixture's own measure under the same methodology. No other
low-risk candidate was identified this session. See the design document's own Decision/Rationale
sections for the resulting recommendation.
