---
title: "Pedigree Diagram: Non-Rigid Layout Feasibility Spike -- Evidence"
subtitle: "Supporting evidence for docs/planning/pedigree-diagram-nonrigid-layout-spike-plan.md -- not a package vignette, not shipped"
date: today
format:
  html:
    embed-resources: true
    toc: true
    df-print: kable
---

## Purpose

`BACKLOG.md` (found S588, HIGH PRIORITY) scoped a bounded, one-session feasibility spike:
prototype ONE non-rigid/constraint-aware layout candidate for `.positionMatingUnitForest()`
(`R/makePedigreeDiagramData.R`) and test it against (a) the sibling design document's own
13-individual synthetic example and (b) the real 375-individual bundled fixture, measured with a
**faithful** reproduction of the shipped algorithm's full final-position pipeline (not the prior
design session's own simplified proxy). Owner selected the candidate paradigm --
**barycenter/median layered-DAG compaction** (a Sugiyama-style technique, the `Graphviz dot`
family) -- over a force-directed alternative, via `AskUserQuestion`.

**No package code is changed here** -- this is investigation feeding the spike's own verdict.
`kinship2` is installed only into this project's local `renv` library for this one-off comparison
(not added to `DESCRIPTION`, not snapshotted into `renv.lock`), matching established precedent.

## Setup

```{r setup}
#| message: false
suppressWarnings(suppressMessages(library(kinship2)))
library(nprcgenekeepr)
```

## The candidate implementation

Seed from the shipped recursive contour-merge's own output (a valid, non-crossing layout --
byte-identical logic to `.positionMatingUnitForest()` through Track 3's first `sweepMinSep()`),
fix the left-right order within each generation row from that seed (never changes after this
point), then iteratively refine `x` via an alternating down-sweep/up-sweep barycenter/median pull,
re-enforcing order-preserving `minSep` after every row update. The refined positions are then fed
through the **identical** downstream pipeline the shipped function uses (`orderBySex` swap, Track
6 `finalUnitX`-from-children, duplicate offset, final de-collision, final sweep) -- verbatim from
`R/makePedigreeDiagramData.R:884-1025` -- for a fair, faithful comparison.

Two real implementation bugs were found and fixed during this session's own development (both
documented inline below): (1) an un-damped, fully-simultaneous ("Jacobi") update let a pair of
nodes sharing their only pull source (e.g. two parents of the same one mating unit) chase an
identical target forever, and the order-preserving sweep -- which can only push right, never left
-- re-separated them by exactly `minSep` every iteration, an unbounded linear drift; (2) an
earlier version used the SAME "unit x" (mean of children) for both the down-sweep and up-sweep
targets, which made a child's down-pull self-referential (toward its own sibling-group mean, not
its actual parents) rather than toward its real parents.

```{r candidate-fn}
#| code-fold: true
## ---- SEED: verbatim copy of .positionMatingUnitForest()'s recursive
## contour-merge machinery, through Track 3's FIRST sweepMinSep() call.
buildSeed <- function(ped, forest) {
  ped$gen[is.na(ped$gen)] <- 0L
  matingUnits <- forest$matingUnits
  duplicates <- forest$duplicates
  childEdges <- forest$childEdges
  unitIds <- matingUnits$id
  realIds <- as.character(ped$id)
  genOf <- stats::setNames(ped$gen, realIds)

  danglingIds <- setdiff(c(matingUnits$sire, matingUnits$dam), realIds)
  if (length(danglingIds) > 0L) {
    fallbackGen <- vapply(danglingIds, function(x) {
      matingUnits$gen[matingUnits$sire == x | matingUnits$dam == x][1L]
    }, integer(1L))
    genOf <- c(genOf, stats::setNames(fallbackGen, danglingIds))
  }

  unitGenOf <- stats::setNames(matingUnits$gen, unitIds)
  maxGen <- max(ped$gen, if (nrow(matingUnits) > 0L) matingUnits$gen else 0L,
                na.rm = TRUE)
  if (!is.finite(maxGen)) maxGen <- 0L
  minSep <- 1L

  leafContour <- function(gen) {
    left <- rep(Inf, maxGen + 1L); right <- rep(-Inf, maxGen + 1L)
    left[gen + 1L] <- 0L; right[gen + 1L] <- 0L
    list(x = 0L, contour = list(left = left, right = right))
  }

  mergeSubtrees <- function(subResults) {
    n <- length(subResults)
    xs <- numeric(n)
    contour <- subResults[[1L]]$contour
    if (n > 1L) {
      for (i in 2L:n) {
        ci <- subResults[[i]]$contour
        finite <- is.finite(contour$right) & is.finite(ci$left)
        shift <- minSep
        if (any(finite)) {
          needed <- max(contour$right[finite] - ci$left[finite] + minSep)
          shift <- max(shift, needed)
        }
        xs[i] <- shift
        contour <- list(left = pmin(contour$left, ci$left + shift),
                         right = pmax(contour$right, ci$right + shift))
      }
    }
    list(xs = xs, contour = contour)
  }

  finalizeNode <- function(merged, ownGen) {
    xs <- merged$xs
    ownX <- (xs[1L] + xs[length(xs)]) / 2L
    shiftAmt <- -ownX
    contour <- list(left = merged$contour$left + shiftAmt,
                     right = merged$contour$right + shiftAmt)
    contour$left[ownGen + 1L] <- min(contour$left[ownGen + 1L], 0L)
    contour$right[ownGen + 1L] <- max(contour$right[ownGen + 1L], 0L)
    list(ownX = ownX, childOffsets = xs - ownX, contour = contour)
  }

  anchoredUnits <- matingUnits[!is.na(matingUnits$anchor), , drop = FALSE]
  everAnchor <- unique(anchoredUnits$anchor)
  nonAnchorSides <- c(anchoredUnits$sire, anchoredUnits$dam)
  neverAnchorIds <- setdiff(unique(nonAnchorSides), everAnchor)
  hasOwnDirectChild <- function(id) {
    any(childEdges$from == id & !(childEdges$from %in% unitIds))
  }
  freePassIds <- Filter(function(id) !hasOwnDirectChild(id), neverAnchorIds)
  freePassUnitOf <- stats::setNames(character(length(freePassIds)), freePassIds)
  for (fp in freePassIds) {
    ownUnits <- matingUnits$id[matingUnits$sire == fp | matingUnits$dam == fp]
    dupUnits <- duplicates$matingUnitId[duplicates$realId == fp]
    freePassUnitOf[[fp]] <- setdiff(ownUnits, dupUnits)[1L]
  }
  freePassOfUnit <- split(names(freePassUnitOf), freePassUnitOf)

  dispGenOf <- genOf[realIds]
  realFreePassIds <- intersect(freePassIds, realIds)
  if (length(realFreePassIds) > 0L) {
    dispGenOf[realFreePassIds] <-
      unname(unitGenOf[freePassUnitOf[realFreePassIds]])
  }

  relNode <- new.env(parent = emptyenv())

  positionUnit <- function(unitId) {
    kidIds <- childEdges$to[childEdges$from == unitId]
    fpHere <- freePassOfUnit[[unitId]]
    subIds <- c(fpHere, kidIds)
    subResults <- lapply(subIds, function(sid) {
      if (!is.null(fpHere) && sid %in% fpHere) {
        leafContour(unitGenOf[[unitId]])
      } else {
        positionIndividual(sid)
      }
    })
    fin <- finalizeNode(mergeSubtrees(subResults), unitGenOf[[unitId]])
    relNode[[unitId]] <- list(childIds = subIds, childOffsets = fin$childOffsets)
    list(x = fin$ownX, contour = fin$contour)
  }

  positionIndividual <- function(id) {
    unitSub <- matingUnits$id[matingUnits$anchor %in% id]
    directSub <- childEdges$to[childEdges$from == id &
                                  !(childEdges$from %in% unitIds)]
    subIds <- c(unitSub, directSub)
    if (length(subIds) == 0L) {
      relNode[[id]] <- list(childIds = character(0L), childOffsets = numeric(0L))
      return(leafContour(genOf[[id]]))
    }
    subResults <- lapply(subIds, function(sid) {
      if (sid %in% unitIds) positionUnit(sid) else positionIndividual(sid)
    })
    fin <- finalizeNode(mergeSubtrees(subResults), genOf[[id]])
    relNode[[id]] <- list(childIds = subIds, childOffsets = fin$childOffsets)
    list(x = fin$ownX, contour = fin$contour)
  }

  hasParentEdge <- realIds %in% childEdges$to
  founderIds <- realIds[!hasParentEdge]
  rootIds <- setdiff(founderIds, freePassIds)
  rootIds <- rootIds[order(match(rootIds, realIds))]
  orphanUnitIds <- matingUnits$id[is.na(matingUnits$anchor)]

  rootResults <- c(lapply(rootIds, positionIndividual),
                    lapply(orphanUnitIds, positionUnit))
  allRootIds <- c(rootIds, orphanUnitIds)
  rootMerge <- mergeSubtrees(rootResults)

  absX <- new.env(parent = emptyenv())
  assignAbs <- function(id, base) {
    absX[[id]] <- base
    node <- relNode[[id]]
    if (!is.null(node) && length(node$childIds) > 0L) {
      for (i in seq_along(node$childIds)) {
        assignAbs(node$childIds[i], base + node$childOffsets[i])
      }
    }
  }
  for (i in seq_along(allRootIds)) assignAbs(allRootIds[i], rootMerge$xs[i])

  realX <- vapply(realIds, function(i) absX[[i]], numeric(1L))

  sweepMinSep <- function(ids, xVals, genVals) {
    x <- stats::setNames(xVals, ids)
    for (g in unique(genVals)) {
      rowIds <- ids[genVals == g]
      if (length(rowIds) < 2L) next
      rowIds <- rowIds[order(x[rowIds], rowIds, method = "radix")]
      for (i in 2L:length(rowIds)) {
        prevX <- x[[rowIds[i - 1L]]]
        if (x[[rowIds[i]]] < prevX + minSep) x[[rowIds[i]]] <- prevX + minSep
      }
    }
    x
  }
  sweepGen <- unname(dispGenOf[realIds])
  sweepX <- sweepMinSep(realIds, unname(realX), sweepGen)
  realX <- unname(sweepX[realIds])

  list(realIds = realIds, realX = realX, dispGenOf = dispGenOf,
       matingUnits = matingUnits, childEdges = childEdges,
       duplicates = duplicates, unitIds = unitIds, unitGenOf = unitGenOf,
       minSep = minSep, sweepGen = sweepGen, sweepMinSep = sweepMinSep,
       hasOwnDirectChild = hasOwnDirectChild, ped = ped)
}

## ---- CANDIDATE: order-preserving barycenter/median compaction refinement.
refineBarycenter <- function(seed, rounds = 30L, tol = 1e-6, alpha = 0.6) {
  realIds <- seed$realIds
  x <- stats::setNames(seed$realX, realIds)
  gen <- unname(seed$dispGenOf[realIds])
  matingUnits <- seed$matingUnits
  childEdges <- seed$childEdges
  unitIds <- seed$unitIds
  minSep <- seed$minSep

  rowOrder <- split(realIds, gen)
  rowOrder <- lapply(rowOrder, function(ids) {
    ids[order(x[ids], ids, method = "radix")]
  })
  genLevels <- sort(as.numeric(names(rowOrder)))

  unitChildrenOf <- stats::setNames(
    lapply(matingUnits$id, function(u) {
      intersect(childEdges$to[childEdges$from == u], realIds)
    }),
    matingUnits$id
  )
  memberUnitsOf <- split(matingUnits$id, matingUnits$sire)
  memberUnitsOfDam <- split(matingUnits$id, matingUnits$dam)
  for (nm in names(memberUnitsOfDam)) {
    memberUnitsOf[[nm]] <- union(memberUnitsOf[[nm]], memberUnitsOfDam[[nm]])
  }
  parentOf <- split(childEdges$from, childEdges$to)
  unitSireOf <- stats::setNames(matingUnits$sire, matingUnits$id)
  unitDamOf <- stats::setNames(matingUnits$dam, matingUnits$id)

  liveUnitChildX <- function() {
    stats::setNames(
      vapply(matingUnits$id, function(u) {
        kids <- unitChildrenOf[[u]]
        if (length(kids) == 0L) NA_real_ else mean(x[kids])
      }, numeric(1L)),
      matingUnits$id
    )
  }
  liveUnitParentX <- function() {
    stats::setNames(
      vapply(matingUnits$id, function(u) {
        sireId <- unitSireOf[[u]]; damId <- unitDamOf[[u]]
        vals <- c(if (sireId %in% realIds) x[[sireId]] else NA_real_,
                  if (damId %in% realIds) x[[damId]] else NA_real_)
        vals <- vals[is.finite(vals)]
        if (length(vals)) mean(vals) else NA_real_
      }, numeric(1L)),
      matingUnits$id
    )
  }

  sweepRow <- function(g) {
    ids <- rowOrder[[as.character(g)]]
    if (length(ids) < 2L) return(invisible())
    for (i in 2:length(ids)) {
      prevX <- x[[ids[i - 1L]]]
      if (x[[ids[i]]] < prevX + minSep) x[[ids[i]]] <<- prevX + minSep
    }
  }

  updateRow <- function(g, direction) {
    ids <- rowOrder[[as.character(g)]]
    uX <- if (identical(direction, "down")) liveUnitParentX() else liveUnitChildX()
    for (id in ids) {
      target <- NA_real_
      if (identical(direction, "down")) {
        parents <- parentOf[[id]]
        if (!is.null(parents)) {
          vals <- vapply(parents, function(p) {
            if (p %in% unitIds) uX[[p]] else x[[p]]
          }, numeric(1L))
          vals <- vals[is.finite(vals)]
          if (length(vals)) target <- mean(vals)
        }
      } else {
        units <- memberUnitsOf[[id]]
        if (!is.null(units)) {
          vals <- uX[units]
          vals <- vals[is.finite(vals)]
          if (length(vals)) target <- mean(vals)
        }
      }
      if (is.finite(target)) {
        x[[id]] <<- x[[id]] + alpha * (target - x[[id]])
      }
    }
    sweepRow(g)
  }

  history <- numeric(0L)
  for (round in seq_len(rounds)) {
    xOld <- x
    for (g in genLevels) updateRow(g, "down")
    for (g in rev(genLevels)) updateRow(g, "up")
    delta <- max(abs(x - xOld))
    history <- c(history, delta)
    if (delta < tol) break
  }

  list(realX = unname(x[realIds]), roundsRun = length(history),
       lastDelta = if (length(history)) utils::tail(history, 1L) else NA_real_)
}

## ---- FINISH PIPELINE: verbatim copy of .positionMatingUnitForest()'s
## downstream stages -- R/makePedigreeDiagramData.R:884-1025.
finishPipeline <- function(seed, realX, orderBySex = TRUE) {
  realIds <- seed$realIds
  matingUnits <- seed$matingUnits
  duplicates <- seed$duplicates
  childEdges <- seed$childEdges
  unitIds <- seed$unitIds
  unitGenOf <- seed$unitGenOf
  dispGenOf <- seed$dispGenOf
  minSep <- seed$minSep
  sweepGen <- seed$sweepGen
  sweepMinSep <- seed$sweepMinSep
  hasOwnDirectChild <- seed$hasOwnDirectChild
  ped <- seed$ped

  nodes <- data.frame(
    id = c(realIds, unitIds),
    x = c(unname(realX), rep(NA_real_, length(unitIds))),
    gen = c(unname(dispGenOf), matingUnits$gen),
    stringsAsFactors = FALSE
  )

  if (isTRUE(orderBySex)) {
    sexOf <- stats::setNames(as.character(ped$sex), realIds)
    mateCount <- table(c(matingUnits$sire, matingUnits$dam))
    for (u in seq_len(nrow(matingUnits))) {
      sireId <- matingUnits$sire[u]
      damId <- matingUnits$dam[u]
      if (!(sireId %in% realIds) || !(damId %in% realIds)) next
      if (mateCount[[sireId]] != 1L || mateCount[[damId]] != 1L) next
      if (hasOwnDirectChild(sireId) || hasOwnDirectChild(damId)) next
      sireSex <- sexOf[[sireId]]
      damSex <- sexOf[[damId]]
      isQualifying <- (identical(sireSex, "M") && identical(damSex, "F")) ||
        (identical(sireSex, "F") && identical(damSex, "M"))
      if (!isQualifying) next
      maleId <- if (identical(sireSex, "M")) sireId else damId
      femaleId <- if (identical(sireSex, "F")) sireId else damId
      mIdx <- which(nodes$id == maleId)
      fIdx <- which(nodes$id == femaleId)
      if (nodes$x[mIdx] > nodes$x[fIdx]) {
        tmp <- nodes$x[mIdx]
        nodes$x[mIdx] <- nodes$x[fIdx]
        nodes$x[fIdx] <- tmp
      }
    }
  }

  finalUnitX <- stats::setNames(numeric(nrow(matingUnits)), matingUnits$id)
  if (nrow(matingUnits) > 0L) {
    for (i in seq_len(nrow(matingUnits))) {
      uid <- matingUnits$id[i]
      kids <- childEdges$to[childEdges$from == uid]
      kidX <- nodes$x[match(kids, nodes$id)]
      finalUnitX[[uid]] <- (min(kidX) + max(kidX)) / 2L
    }
  }
  nodes$x[match(matingUnits$id, nodes$id)] <- unname(finalUnitX[matingUnits$id])

  dupX <- numeric(nrow(duplicates))
  if (nrow(duplicates) > 0L) {
    dupX <- unname(finalUnitX[duplicates$matingUnitId]) + minSep * 0.4
  }
  nodes <- rbind(nodes, data.frame(
    id = duplicates$id, x = dupX,
    gen = unname(unitGenOf[duplicates$matingUnitId]),
    stringsAsFactors = FALSE
  ))

  ord <- order(nodes$gen, nodes$id, method = "radix")
  seenAtGen <- new.env(parent = emptyenv())
  for (i in ord) {
    genKey <- as.character(nodes$gen[i])
    used <- seenAtGen[[genKey]]
    xx <- nodes$x[i]
    while (!is.null(used) && any(abs(used - xx) < 1e-9)) {
      xx <- xx + 1e-3
    }
    nodes$x[i] <- xx
    seenAtGen[[genKey]] <- c(used, xx)
  }

  nodes$x[match(realIds, nodes$id)] <-
    unname(sweepMinSep(realIds, nodes$x[match(realIds, nodes$id)], sweepGen))

  nodes
}

positionCandidate <- function(ped, forest, orderBySex = TRUE,
                               rounds = 30L, alpha = 0.6) {
  seed <- buildSeed(ped, forest)
  refined <- refineBarycenter(seed, rounds = rounds, alpha = alpha)
  nodes <- finishPipeline(seed, refined$realX, orderBySex = orderBySex)
  attr(nodes, "roundsRun") <- refined$roundsRun
  attr(nodes, "lastDelta") <- refined$lastDelta
  nodes
}

positionSeedOnly <- function(ped, forest, orderBySex = TRUE) {
  seed <- buildSeed(ped, forest)
  finishPipeline(seed, seed$realX, orderBySex = orderBySex)
}
```

## Cross-check: the seed reproduction is byte-identical to the shipped algorithm

Before trusting any comparison built on it, confirm the "seed" (verbatim-copied recursive
contour-merge) matches `.positionMatingUnitForest()`'s own live output exactly -- both on the
small synthetic example and the real fixture.

```{r crosscheck-small}
build_ped <- function(rows) {
  ped <- do.call(rbind, lapply(rows, as.data.frame, stringsAsFactors = FALSE))
  ped$gen <- findGeneration(ped$id, ped$sire, ped$dam)
  ped
}
r <- function(id, sire = NA, dam = NA, sex) list(id = id, sire = sire, dam = dam, sex = sex)
rows <- list(
  r("P1", sex = "M"), r("P2", sex = "F"),
  r("A", "P1", "P2", "M"), r("B", "P1", "P2", "F"),
  r("BM", sex = "M"),
  r("B1", "BM", "B", "M"), r("B2", "BM", "B", "F"),
  r("B1M", sex = "F"), r("B2M", sex = "M"),
  r("B1a", "B1", "B1M", "F"), r("B1b", "B1", "B1M", "M"),
  r("B2a", "B2M", "B2", "F"), r("B2b", "B2M", "B2", "M")
)
ped <- build_ped(rows)
forest <- nprcgenekeepr:::.buildMatingUnitForest(ped)

shipped <- nprcgenekeepr:::.positionMatingUnitForest(ped, forest, orderBySex = TRUE)
mySeed <- positionSeedOnly(ped, forest, orderBySex = TRUE)
mySeed <- mySeed[order(mySeed$id), ]
shippedOrd <- shipped[order(shipped$id), c("id", "x", "gen")]
cat("Synthetic example: max |x| diff =", max(abs(mySeed$x - shippedOrd$x)), "\n")
```

```{r crosscheck-real}
pedReal <- read.csv("../../inst/extdata/examples/obfuscated_rhesus_mhc_ped.csv",
                     stringsAsFactors = FALSE)
pedReal$id <- as.character(pedReal$id)
pedReal$sire <- as.character(pedReal$sire)
pedReal$dam <- as.character(pedReal$dam)
pedReal$gen <- findGeneration(pedReal$id, pedReal$sire, pedReal$dam)
forestReal <- nprcgenekeepr:::.buildMatingUnitForest(pedReal)

shippedReal <- nprcgenekeepr:::.positionMatingUnitForest(pedReal, forestReal, orderBySex = TRUE)
seedNodesReal <- positionSeedOnly(pedReal, forestReal, orderBySex = TRUE)
seedOrd <- seedNodesReal[order(seedNodesReal$id), ]
shipOrd <- shippedReal[order(shippedReal$id), ]
cat("Real 375-individual fixture: max |x| diff =",
    max(abs(seedOrd$x - shipOrd$x)), "\n")
```

Both are 0 (or effectively 0, floating-point exact) -- the seed is a faithful, byte-identical
reproduction at both scales, so downstream comparisons are trustworthy.

## Synthetic example: candidate vs. shipped

```{r synth-render}
#| fig-width: 11
#| fig-height: 5.5
seedNodes <- positionSeedOnly(ped, forest, orderBySex = TRUE)
candNodes <- positionCandidate(ped, forest, orderBySex = TRUE, rounds = 60L, alpha = 0.3)

renderNodes <- function(nodes, forest, title) {
  childEdges <- forest$childEdges
  matingUnits <- forest$matingUnits
  xOf <- stats::setNames(nodes$x, nodes$id)
  genOf <- stats::setNames(nodes$gen, nodes$id)
  plot(nodes$x, -nodes$gen, type = "n", axes = FALSE, xlab = "", ylab = "",
       main = title, xlim = range(nodes$x) + c(-1, 1))
  for (i in seq_len(nrow(matingUnits))) {
    uid <- matingUnits$id[i]
    for (p in c(matingUnits$sire[i], matingUnits$dam[i])) {
      if (p %in% names(xOf)) {
        segments(xOf[[p]], -genOf[[p]], xOf[[uid]], -genOf[[uid]], col = "grey50")
      }
    }
  }
  for (i in seq_len(nrow(childEdges))) {
    fromId <- childEdges$from[i]; toId <- childEdges$to[i]
    if (fromId %in% names(xOf) && toId %in% names(xOf)) {
      segments(xOf[[fromId]], -genOf[[fromId]], xOf[[toId]], -genOf[[toId]], col = "grey20")
    }
  }
  realNodes <- nodes[!grepl("^__", nodes$id), ]
  points(realNodes$x, -realNodes$gen, pch = 1, cex = 2.2)
  text(realNodes$x, -realNodes$gen, labels = realNodes$id, cex = 0.7, pos = 1, offset = 0.8)
}

par(mfrow = c(1, 2))
gapS <- abs(diff(seedNodes$x[match(c("A", "B"), seedNodes$id)]))
gapC <- abs(diff(candNodes$x[match(c("A", "B"), candNodes$id)]))
renderNodes(seedNodes, forest, sprintf("SHIPPED -- A-B gap = %.2f", gapS))
renderNodes(candNodes, forest,
            sprintf("CANDIDATE barycenter/median -- A-B gap = %.2f", gapC))
```

The candidate closes the `A`-`B` gap from `r round(gapS,2)` to `r round(gapC,2)` (a
`r round(100*(1-gapC/gapS))`% reduction) with **no edge crossings** -- the left-right order within
each generation row is provably unchanged from the seed by construction (verified below), so no
NEW crossing beyond what the seed already had is possible.

```{r synth-order-check}
checkOrderPreserved <- function(seedNodes, candNodes) {
  realIds <- intersect(seedNodes$id[!grepl("^__", seedNodes$id)],
                        candNodes$id[!grepl("^__", candNodes$id)])
  seedX <- stats::setNames(seedNodes$x, seedNodes$id)
  candX <- stats::setNames(candNodes$x, candNodes$id)
  seedGen <- stats::setNames(seedNodes$gen, seedNodes$id)
  mismatches <- 0L
  for (g in unique(seedGen[realIds])) {
    ids <- realIds[seedGen[realIds] == g]
    if (length(ids) < 2L) next
    if (!identical(ids[order(seedX[ids])], ids[order(candX[ids])])) mismatches <- mismatches + 1L
  }
  mismatches
}
cat("Row-order mismatches (synthetic example):", checkOrderPreserved(seedNodes, candNodes), "\n")
```

## Real 375-individual fixture: the faithful full-pipeline measurement

This uses the **exact** methodology Track 6 established
(`tests/testthat/test_positionMatingUnitForest.R`, ~line 1030): for each mating unit, each child
edge's `|childX - unitX|`, where `unitX` is the FULL-pipeline `finalUnitX` (post-`orderBySex`,
post-final-de-collision) -- not the prior design session's simplified proxy. `xScale = 120`
(`R/makePedigreeDiagramData.R:1166`) -- Track 6's own "200 units" threshold is stated in SCALED
(rendered) units, applied here to this function's raw `x` accordingly.

```{r real-measure}
xScale <- 120
measureFaithful <- function(nodes, forest) {
  matingUnits <- forest$matingUnits
  childEdges <- forest$childEdges
  xOf <- stats::setNames(nodes$x, nodes$id)
  offsets <- c()
  for (i in seq_len(nrow(matingUnits))) {
    uid <- matingUnits$id[i]
    kids <- intersect(childEdges$to[childEdges$from == uid], names(xOf))
    if (length(kids) == 0L) next
    unitX <- xOf[[uid]]
    for (k in kids) offsets <- c(offsets, abs(xOf[[k]] - unitX))
  }
  offsetsScaled <- offsets * xScale
  c(n = length(offsetsScaled), n_over_200 = sum(offsetsScaled > 200),
    pct_over_200 = mean(offsetsScaled > 200) * 100,
    max_offset_scaled = max(offsetsScaled))
}

baselineMetric <- measureFaithful(seedNodesReal, forestReal)
baselineMetric
```

This exactly reproduces Track 6's own published baseline (9/251 edges, 3.6%, max 4,121.25) --
confirming the harness is faithful, not an approximation.

```{r real-candidate}
t0 <- Sys.time()
candNodesReal <- positionCandidate(pedReal, forestReal, orderBySex = TRUE,
                                    rounds = 60L, alpha = 0.3)
candTime <- as.numeric(Sys.time() - t0, units = "secs")
cat("Candidate compute time (60 rounds, 375 individuals):", round(candTime, 2), "s\n")
cat("lastDelta after 60 rounds:", attr(candNodesReal, "lastDelta"),
    "(still moving -- see 'why it fails to converge' below)\n\n")

candidateMetric <- measureFaithful(candNodesReal, forestReal)
candidateMetric
```

```{r real-comparison}
cat(sprintf("Baseline:  %d/%d edges >200 units (%.2f%%), max offset %.2f\n",
            baselineMetric[["n_over_200"]], baselineMetric[["n"]],
            baselineMetric[["pct_over_200"]], baselineMetric[["max_offset_scaled"]]))
cat(sprintf("Candidate: %d/%d edges >200 units (%.2f%%), max offset %.2f\n",
            candidateMetric[["n_over_200"]], candidateMetric[["n"]],
            candidateMetric[["pct_over_200"]], candidateMetric[["max_offset_scaled"]]))

realSeed <- seedNodesReal[!grepl("^__", seedNodesReal$id), ]
realCand <- candNodesReal[!grepl("^__", candNodesReal$id), ]
cat(sprintf("\nOverall real-node x range -- baseline: %.1f, candidate: %.1f (%.1fx)\n",
            diff(range(realSeed$x)), diff(range(realCand$x)),
            diff(range(realCand$x)) / diff(range(realSeed$x))))

cat("\nRow-order mismatches (real fixture, should be 0):",
    checkOrderPreserved(seedNodesReal, candNodesReal), "\n")
```

**The candidate regresses the real fixture**: 15/251 edges exceed the 200-unit threshold (5.98%,
vs. baseline 3.6%), the worst single offset grows from 4,121 to 5,344, and the overall layout
width for real individuals more than sextuples (126.8 -> 773.7 raw units). This is not a subtle
regression -- it is a structural instability. Row order IS still fully preserved (0 mismatches),
so no new edge crossing was introduced -- the failure mode here is different from the previously-
rejected bounded-lookahead candidate's (which introduced a crossing): this one fails on
convergence/stability, not on crossing-freedom.

### Hyperparameter sweep -- the regression is not a tuning artifact

```{r hp-sweep}
sweepResults <- lapply(list(
  list(rounds = 30, alpha = 0.3), list(rounds = 150, alpha = 0.3),
  list(rounds = 150, alpha = 0.15), list(rounds = 300, alpha = 0.15),
  list(rounds = 60, alpha = 0.6)
), function(p) {
  cand <- positionCandidate(pedReal, forestReal, orderBySex = TRUE,
                             rounds = p$rounds, alpha = p$alpha)
  m <- measureFaithful(cand, forestReal)
  data.frame(rounds = p$rounds, alpha = p$alpha,
             lastDelta = round(attr(cand, "lastDelta"), 2),
             n_over_200 = m[["n_over_200"]], pct = round(m[["pct_over_200"]], 2),
             max_offset = round(m[["max_offset_scaled"]], 1))
})
do.call(rbind, sweepResults)
```

Every combination tested regresses relative to the 9/251 (3.6%) baseline -- never better than
15/251. `lastDelta` stays well above 0 in every row: the algorithm has not reached a true fixed
point at this scale within a practical round budget, unlike the small synthetic example (which
stabilizes cleanly within ~30-60 rounds).

### Why it fails to converge at real-fixture scale: high-mate-count hub individuals

```{r hub-diagnostic}
mu <- forestReal$matingUnits
worst <- mu[mu$id == "__union_69", ]
worst
cat("Sire", worst$sire, "own mate count:",
    sum(mu$sire == worst$sire | mu$dam == worst$sire), "\n")
```

The single worst-regressed edge pair (offset 195 -> 5,344 scaled units, a 27x worsening) traces to
a mating unit whose sire has **5 separate mating unions** -- a highly-connected "hub" individual
entirely absent from the 13-individual synthetic example (max 1 mate per individual there). A hub
like this couples many different subtrees' barycenter targets together each round, and this
session's row-sequential damped-relaxation design has no formal convergence guarantee for that
topology -- unlike a properly-implemented, published algorithm (e.g. Brandes & Kopf 2002's
horizontal coordinate assignment, or `igraph::layout_with_sugiyama()`, which was checked and found
not installed in this environment but is a well-established CRAN package implementing exactly this
technique with its own convergence properties). This is the concrete, structural reason a second
spike -- if pursued -- should adapt a proven implementation rather than tune this one further.

## Conclusion

The barycenter/median candidate is **rejected as prototyped**: it achieves a modest, real
improvement on the small synthetic example (20% gap reduction, provably no new crossings) but
**regresses the real 375-individual fixture** under the faithful full-pipeline measurement (9/251
-> 15/251 edges over threshold, 6x overall width growth), consistently across a 5-point
hyperparameter sweep, with a concrete, diagnosed cause (unstable convergence at high-mate-count
hub individuals, a topology the small synthetic example never exercises). This is the SECOND
independently-designed candidate (after the sibling design session's bounded-lookahead candidate)
to show the same qualitative pattern -- improves the toy example, regresses the real fixture --
via a different failure mechanism each time (edge crossings there; convergence instability here).
See the spike's own plan document for the resulting verdict and recommendation.
