---
title: "Pedigree Diagram: igraph::layout_with_sugiyama() Layout Feasibility Spike -- Evidence"
subtitle: "Supporting evidence for docs/planning/pedigree-diagram-layout-sugiyama-spike-plan.md -- not a package vignette, not shipped"
date: today
format:
  html:
    embed-resources: true
    toc: true
    df-print: kable
---

## Purpose

`BACKLOG.md` (found S589, HIGH PRIORITY) scoped this SECOND feasibility spike: adapt a proven,
convergence-guaranteed layout implementation -- `igraph::layout_with_sugiyama()` -- to this
project's mating-unit-forest data structures, and test it against the SAME two fixtures the first
spike (S589, barycenter/median) used, reusing that spike's own faithful full-pipeline metric.
Owner selected `igraph::layout_with_sugiyama()` over a ported Brandes-Köpf (2002) alternative via
`AskUserQuestion`, at this session's Phase 1.

**No package code is changed here.** `igraph` is installed only into this project's local `renv`
library for this one-off comparison (confirmed installable this session; not added to
`DESCRIPTION`, not snapshotted into `renv.lock` -- matching the `kinship2` reference-only
precedent).

**Incidental finding, reported not fixed (`PROJECT_LEARNINGS.md` Learning 382 precedent):** the
renv-cached INSTALLED `nprcgenekeepr` copy was built 2026-08-14 18:24, ~3.5h BEFORE Track 6
shipped (commit `f65ecbea`, 2026-08-14 21:56:10) -- `library(nprcgenekeepr)` silently loads that
STALE pre-Track-6 build. This session used `pkgload::load_all()` throughout instead, matching this
project's own documented build-equivalent convention (`CLAUDE.md` "Fast single-file test"). Every
number below is against TRUE current source, confirmed by a 0-diff crosscheck at both fixture
scales (below) -- but the prior spike's own qmd used `library(nprcgenekeepr)`, so its own
crosscheck evidence rests on this same stale-package risk. Not re-verified here (out of this
session's own scope); a future session touching that document should re-render it with
`load_all()` to confirm its numbers are unaffected.

## Setup

```{r setup}
#| message: false
suppressMessages(pkgload::load_all(".", quiet = TRUE))
suppressWarnings(suppressMessages(library(igraph)))
```

## The candidate implementation

Build the same parent-union-child DAG `.positionMatingUnitForest()` walks (vertices = real
individuals + mating units; edges = sire/dam -> unit, unit/individual -> child, verbatim from
`forest$matingUnits`/`forest$childEdges`), with each vertex's `layer` fixed to its DISPLAY
generation (`dispGenOf` for individuals, `unitGenOf` for units -- identical to the shipped
function's own generation accounting). Hand this graph to `igraph::layout_with_sugiyama()`,
extract only the REAL individuals' `x` from its layout (mating-unit `x` is discarded --
`finishPipeline()`'s own Track 6 `finalUnitX` recomputes it from children exactly as the shipped
pipeline does), then feed those `x` values through the **identical** downstream pipeline the
shipped function uses (`orderBySex` swap, Track 6 `finalUnitX`-from-children, duplicate offset,
final de-collision, final sweep) -- verbatim from `R/makePedigreeDiagramData.R:884-1025`, reused
byte-for-byte from the first spike's own harness.

```{r harness-fns}
#| code-fold: true
## ---- SEED + FINISH PIPELINE: verbatim copy, reused from the first spike's own
## harness (docs/planning/pedigree-diagram-nonrigid-layout-spike-evidence.qmd),
## itself a byte-identical transcription of R/makePedigreeDiagramData.R:584-1025.
buildSeed <- function(ped, forest) {
  ped$gen[is.na(ped$gen)] <- 0L
  matingUnits <- forest$matingUnits
  duplicates <- forest$duplicates
  childEdges <- forest$childEdges
  unitIds <- matingUnits$id
  realIds <- as.character(ped$id)
  genOf <- stats::setNames(ped$gen, realIds)

  danglingIds <- setdiff(c(matingUnits$sire, matingUnits$dam), realIds)
  if (length(danglingIds) > 0L) {
    fallbackGen <- vapply(danglingIds, function(x) {
      matingUnits$gen[matingUnits$sire == x | matingUnits$dam == x][1L]
    }, integer(1L))
    genOf <- c(genOf, stats::setNames(fallbackGen, danglingIds))
  }

  unitGenOf <- stats::setNames(matingUnits$gen, unitIds)
  maxGen <- max(ped$gen, if (nrow(matingUnits) > 0L) matingUnits$gen else 0L,
                na.rm = TRUE)
  if (!is.finite(maxGen)) maxGen <- 0L
  minSep <- 1L

  leafContour <- function(gen) {
    left <- rep(Inf, maxGen + 1L); right <- rep(-Inf, maxGen + 1L)
    left[gen + 1L] <- 0L; right[gen + 1L] <- 0L
    list(x = 0L, contour = list(left = left, right = right))
  }

  mergeSubtrees <- function(subResults) {
    n <- length(subResults)
    xs <- numeric(n)
    contour <- subResults[[1L]]$contour
    if (n > 1L) {
      for (i in 2L:n) {
        ci <- subResults[[i]]$contour
        finite <- is.finite(contour$right) & is.finite(ci$left)
        shift <- minSep
        if (any(finite)) {
          needed <- max(contour$right[finite] - ci$left[finite] + minSep)
          shift <- max(shift, needed)
        }
        xs[i] <- shift
        contour <- list(left = pmin(contour$left, ci$left + shift),
                         right = pmax(contour$right, ci$right + shift))
      }
    }
    list(xs = xs, contour = contour)
  }

  finalizeNode <- function(merged, ownGen) {
    xs <- merged$xs
    ownX <- (xs[1L] + xs[length(xs)]) / 2L
    shiftAmt <- -ownX
    contour <- list(left = merged$contour$left + shiftAmt,
                     right = merged$contour$right + shiftAmt)
    contour$left[ownGen + 1L] <- min(contour$left[ownGen + 1L], 0L)
    contour$right[ownGen + 1L] <- max(contour$right[ownGen + 1L], 0L)
    list(ownX = ownX, childOffsets = xs - ownX, contour = contour)
  }

  anchoredUnits <- matingUnits[!is.na(matingUnits$anchor), , drop = FALSE]
  everAnchor <- unique(anchoredUnits$anchor)
  nonAnchorSides <- c(anchoredUnits$sire, anchoredUnits$dam)
  neverAnchorIds <- setdiff(unique(nonAnchorSides), everAnchor)
  hasOwnDirectChild <- function(id) {
    any(childEdges$from == id & !(childEdges$from %in% unitIds))
  }
  freePassIds <- Filter(function(id) !hasOwnDirectChild(id), neverAnchorIds)
  freePassUnitOf <- stats::setNames(character(length(freePassIds)), freePassIds)
  for (fp in freePassIds) {
    ownUnits <- matingUnits$id[matingUnits$sire == fp | matingUnits$dam == fp]
    dupUnits <- duplicates$matingUnitId[duplicates$realId == fp]
    freePassUnitOf[[fp]] <- setdiff(ownUnits, dupUnits)[1L]
  }
  freePassOfUnit <- split(names(freePassUnitOf), freePassUnitOf)

  dispGenOf <- genOf[realIds]
  realFreePassIds <- intersect(freePassIds, realIds)
  if (length(realFreePassIds) > 0L) {
    dispGenOf[realFreePassIds] <-
      unname(unitGenOf[freePassUnitOf[realFreePassIds]])
  }

  relNode <- new.env(parent = emptyenv())

  positionUnit <- function(unitId) {
    kidIds <- childEdges$to[childEdges$from == unitId]
    fpHere <- freePassOfUnit[[unitId]]
    subIds <- c(fpHere, kidIds)
    subResults <- lapply(subIds, function(sid) {
      if (!is.null(fpHere) && sid %in% fpHere) {
        leafContour(unitGenOf[[unitId]])
      } else {
        positionIndividual(sid)
      }
    })
    fin <- finalizeNode(mergeSubtrees(subResults), unitGenOf[[unitId]])
    relNode[[unitId]] <- list(childIds = subIds, childOffsets = fin$childOffsets)
    list(x = fin$ownX, contour = fin$contour)
  }

  positionIndividual <- function(id) {
    unitSub <- matingUnits$id[matingUnits$anchor %in% id]
    directSub <- childEdges$to[childEdges$from == id &
                                  !(childEdges$from %in% unitIds)]
    subIds <- c(unitSub, directSub)
    if (length(subIds) == 0L) {
      relNode[[id]] <- list(childIds = character(0L), childOffsets = numeric(0L))
      return(leafContour(genOf[[id]]))
    }
    subResults <- lapply(subIds, function(sid) {
      if (sid %in% unitIds) positionUnit(sid) else positionIndividual(sid)
    })
    fin <- finalizeNode(mergeSubtrees(subResults), genOf[[id]])
    relNode[[id]] <- list(childIds = subIds, childOffsets = fin$childOffsets)
    list(x = fin$ownX, contour = fin$contour)
  }

  hasParentEdge <- realIds %in% childEdges$to
  founderIds <- realIds[!hasParentEdge]
  rootIds <- setdiff(founderIds, freePassIds)
  rootIds <- rootIds[order(match(rootIds, realIds))]
  orphanUnitIds <- matingUnits$id[is.na(matingUnits$anchor)]

  rootResults <- c(lapply(rootIds, positionIndividual),
                    lapply(orphanUnitIds, positionUnit))
  allRootIds <- c(rootIds, orphanUnitIds)
  rootMerge <- mergeSubtrees(rootResults)

  absX <- new.env(parent = emptyenv())
  assignAbs <- function(id, base) {
    absX[[id]] <- base
    node <- relNode[[id]]
    if (!is.null(node) && length(node$childIds) > 0L) {
      for (i in seq_along(node$childIds)) {
        assignAbs(node$childIds[i], base + node$childOffsets[i])
      }
    }
  }
  for (i in seq_along(allRootIds)) assignAbs(allRootIds[i], rootMerge$xs[i])

  realX <- vapply(realIds, function(i) absX[[i]], numeric(1L))

  sweepMinSep <- function(ids, xVals, genVals) {
    x <- stats::setNames(xVals, ids)
    for (g in unique(genVals)) {
      rowIds <- ids[genVals == g]
      if (length(rowIds) < 2L) next
      rowIds <- rowIds[order(x[rowIds], rowIds, method = "radix")]
      for (i in 2L:length(rowIds)) {
        prevX <- x[[rowIds[i - 1L]]]
        if (x[[rowIds[i]]] < prevX + minSep) x[[rowIds[i]]] <- prevX + minSep
      }
    }
    x
  }
  sweepGen <- unname(dispGenOf[realIds])
  sweepX <- sweepMinSep(realIds, unname(realX), sweepGen)
  realX <- unname(sweepX[realIds])

  list(realIds = realIds, realX = realX, dispGenOf = dispGenOf,
       matingUnits = matingUnits, childEdges = childEdges,
       duplicates = duplicates, unitIds = unitIds, unitGenOf = unitGenOf,
       minSep = minSep, sweepGen = sweepGen, sweepMinSep = sweepMinSep,
       hasOwnDirectChild = hasOwnDirectChild, ped = ped)
}

finishPipeline <- function(seed, realX, orderBySex = TRUE) {
  realIds <- seed$realIds
  matingUnits <- seed$matingUnits
  duplicates <- seed$duplicates
  childEdges <- seed$childEdges
  unitIds <- seed$unitIds
  unitGenOf <- seed$unitGenOf
  dispGenOf <- seed$dispGenOf
  minSep <- seed$minSep
  sweepGen <- seed$sweepGen
  sweepMinSep <- seed$sweepMinSep
  hasOwnDirectChild <- seed$hasOwnDirectChild
  ped <- seed$ped

  nodes <- data.frame(
    id = c(realIds, unitIds),
    x = c(unname(realX), rep(NA_real_, length(unitIds))),
    gen = c(unname(dispGenOf), matingUnits$gen),
    stringsAsFactors = FALSE
  )

  if (isTRUE(orderBySex)) {
    sexOf <- stats::setNames(as.character(ped$sex), realIds)
    mateCount <- table(c(matingUnits$sire, matingUnits$dam))
    for (u in seq_len(nrow(matingUnits))) {
      sireId <- matingUnits$sire[u]
      damId <- matingUnits$dam[u]
      if (!(sireId %in% realIds) || !(damId %in% realIds)) next
      if (mateCount[[sireId]] != 1L || mateCount[[damId]] != 1L) next
      if (hasOwnDirectChild(sireId) || hasOwnDirectChild(damId)) next
      sireSex <- sexOf[[sireId]]
      damSex <- sexOf[[damId]]
      isQualifying <- (identical(sireSex, "M") && identical(damSex, "F")) ||
        (identical(sireSex, "F") && identical(damSex, "M"))
      if (!isQualifying) next
      maleId <- if (identical(sireSex, "M")) sireId else damId
      femaleId <- if (identical(sireSex, "F")) sireId else damId
      mIdx <- which(nodes$id == maleId)
      fIdx <- which(nodes$id == femaleId)
      if (nodes$x[mIdx] > nodes$x[fIdx]) {
        tmp <- nodes$x[mIdx]
        nodes$x[mIdx] <- nodes$x[fIdx]
        nodes$x[fIdx] <- tmp
      }
    }
  }

  finalUnitX <- stats::setNames(numeric(nrow(matingUnits)), matingUnits$id)
  if (nrow(matingUnits) > 0L) {
    for (i in seq_len(nrow(matingUnits))) {
      uid <- matingUnits$id[i]
      kids <- childEdges$to[childEdges$from == uid]
      kidX <- nodes$x[match(kids, nodes$id)]
      finalUnitX[[uid]] <- (min(kidX) + max(kidX)) / 2L
    }
  }
  nodes$x[match(matingUnits$id, nodes$id)] <- unname(finalUnitX[matingUnits$id])

  dupX <- numeric(nrow(duplicates))
  if (nrow(duplicates) > 0L) {
    dupX <- unname(finalUnitX[duplicates$matingUnitId]) + minSep * 0.4
  }
  nodes <- rbind(nodes, data.frame(
    id = duplicates$id, x = dupX,
    gen = unname(unitGenOf[duplicates$matingUnitId]),
    stringsAsFactors = FALSE
  ))

  ord <- order(nodes$gen, nodes$id, method = "radix")
  seenAtGen <- new.env(parent = emptyenv())
  for (i in ord) {
    genKey <- as.character(nodes$gen[i])
    used <- seenAtGen[[genKey]]
    xx <- nodes$x[i]
    while (!is.null(used) && any(abs(used - xx) < 1e-9)) {
      xx <- xx + 1e-3
    }
    nodes$x[i] <- xx
    seenAtGen[[genKey]] <- c(used, xx)
  }

  nodes$x[match(realIds, nodes$id)] <-
    unname(sweepMinSep(realIds, nodes$x[match(realIds, nodes$id)], sweepGen))

  nodes
}

positionSeedOnly <- function(ped, forest, orderBySex = TRUE) {
  seed <- buildSeed(ped, forest)
  finishPipeline(seed, seed$realX, orderBySex = orderBySex)
}

## ---- Faithful full-pipeline metric -- verbatim from the first spike's evidence.qmd.
xScale <- 120
measureFaithful <- function(nodes, forest) {
  matingUnits <- forest$matingUnits
  childEdges <- forest$childEdges
  xOf <- stats::setNames(nodes$x, nodes$id)
  offsets <- c()
  for (i in seq_len(nrow(matingUnits))) {
    uid <- matingUnits$id[i]
    kids <- intersect(childEdges$to[childEdges$from == uid], names(xOf))
    if (length(kids) == 0L) next
    unitX <- xOf[[uid]]
    for (k in kids) offsets <- c(offsets, abs(xOf[[k]] - unitX))
  }
  offsetsScaled <- offsets * xScale
  c(n = length(offsetsScaled), n_over_200 = sum(offsetsScaled > 200),
    pct_over_200 = mean(offsetsScaled > 200) * 100,
    max_offset_scaled = max(offsetsScaled))
}

## ---- NEW (S590): absolute edge-crossing counter -- needed because, unlike the
## first spike's own candidate (which fixed row order from a known non-crossing
## seed BY CONSTRUCTION), igraph's own crossing-minimization can and does
## reorder rows, so a comparison-to-seed order check is not a valid crossing
## proxy here. Counts crossings directly: for every pair of edges connecting
## the SAME two (lo,hi) display gens, they cross iff the left-to-right order of
## their low-gen endpoints disagrees with the order of their high-gen endpoints.
countCrossings <- function(nodes, forest) {
  xOf <- stats::setNames(nodes$x, nodes$id)
  genOf <- stats::setNames(nodes$gen, nodes$id)
  matingUnits <- forest$matingUnits
  childEdges <- forest$childEdges
  eFrom <- c(); eTo <- c()
  for (i in seq_len(nrow(matingUnits))) {
    uid <- matingUnits$id[i]
    for (p in c(matingUnits$sire[i], matingUnits$dam[i])) {
      if (p %in% names(xOf) && uid %in% names(xOf)) {
        eFrom <- c(eFrom, p); eTo <- c(eTo, uid)
      }
    }
  }
  for (i in seq_len(nrow(childEdges))) {
    f <- childEdges$from[i]; t <- childEdges$to[i]
    if (f %in% names(xOf) && t %in% names(xOf)) {
      eFrom <- c(eFrom, f); eTo <- c(eTo, t)
    }
  }
  gFrom <- genOf[eFrom]; gTo <- genOf[eTo]
  lo <- pmin(gFrom, gTo); hi <- pmax(gFrom, gTo)
  swap <- gFrom > gTo
  loId <- ifelse(swap, eTo, eFrom)
  hiId <- ifelse(swap, eFrom, eTo)
  key <- paste(lo, hi)
  crossings <- 0L
  for (k in unique(key)) {
    idx <- which(key == k)
    if (length(idx) < 2L) next
    xlo <- xOf[loId[idx]]; xhi <- xOf[hiId[idx]]
    n <- length(idx)
    for (a in seq_len(n - 1L)) for (b in (a + 1L):n) {
      dlo <- xlo[a] - xlo[b]; dhi <- xhi[a] - xhi[b]
      if (dlo != 0 && dhi != 0 && sign(dlo) != sign(dhi)) crossings <- crossings + 1L
    }
  }
  crossings
}

## ---- CANDIDATE: build the parent-union-child DAG and hand it to
## igraph::layout_with_sugiyama().
buildSugiyamaGraph <- function(seed) {
  realIds <- seed$realIds
  unitIds <- seed$unitIds
  matingUnits <- seed$matingUnits
  childEdges <- seed$childEdges
  dispGenOf <- seed$dispGenOf
  unitGenOf <- seed$unitGenOf

  vIds <- c(realIds, unitIds)
  vLayer <- c(unname(dispGenOf[realIds]), unname(unitGenOf[unitIds]))

  eFrom <- c(); eTo <- c()
  for (i in seq_len(nrow(matingUnits))) {
    uid <- matingUnits$id[i]
    for (p in c(matingUnits$sire[i], matingUnits$dam[i])) {
      if (p %in% realIds) { eFrom <- c(eFrom, p); eTo <- c(eTo, uid) }
    }
  }
  for (i in seq_len(nrow(childEdges))) {
    eFrom <- c(eFrom, childEdges$from[i]); eTo <- c(eTo, childEdges$to[i])
  }

  igraph::graph_from_data_frame(
    d = data.frame(from = eFrom, to = eTo, stringsAsFactors = FALSE),
    vertices = data.frame(name = vIds, layer = vLayer, stringsAsFactors = FALSE),
    directed = TRUE
  )
}

## Multi-restart (S590 finding, see "Sensitivity" below): igraph's own
## crossing-minimization heuristic is sensitive to input vertex order -- the
## natural construction order hit an avoidable local optimum on the small
## synthetic example (4 crossings vs. achievable 0). Standard mitigation for
## layered-graph heuristics: run `restarts` random vertex-order trials, keep
## whichever minimizes actual crossings (this document's own countCrossings(),
## not a proxy). Still "adapt the proven implementation," not a new algorithm.
positionIgraphCandidate <- function(ped, forest, orderBySex = TRUE, hgap = 1,
                                     restarts = 20L, seed = 1L) {
  sd <- buildSeed(ped, forest)
  g <- buildSugiyamaGraph(sd)
  n <- igraph::vcount(g)
  set.seed(seed)
  bestNodes <- NULL
  bestCrossings <- Inf
  for (trial in seq_len(restarts)) {
    perm <- if (trial == 1L) seq_len(n) else sample(n)
    gp <- igraph::permute(g, perm)
    res <- igraph::layout_with_sugiyama(gp, layers = igraph::V(gp)$layer,
                                         hgap = hgap, vgap = 1)
    xOf <- stats::setNames(res$layout[, 1], igraph::V(gp)$name)
    realX <- unname(xOf[sd$realIds])
    nodes <- finishPipeline(sd, realX, orderBySex = orderBySex)
    cr <- countCrossings(nodes, forest)
    if (cr < bestCrossings) {
      bestCrossings <- cr
      bestNodes <- nodes
      if (cr == 0L) break
    }
  }
  attr(bestNodes, "crossings") <- bestCrossings
  bestNodes
}
```

## Cross-check: the seed reproduction is byte-identical to the shipped algorithm

```{r crosscheck-small}
build_ped <- function(rows) {
  ped <- do.call(rbind, lapply(rows, as.data.frame, stringsAsFactors = FALSE))
  ped$gen <- findGeneration(ped$id, ped$sire, ped$dam)
  ped
}
r <- function(id, sire = NA, dam = NA, sex) list(id = id, sire = sire, dam = dam, sex = sex)
rows <- list(
  r("P1", sex = "M"), r("P2", sex = "F"),
  r("A", "P1", "P2", "M"), r("B", "P1", "P2", "F"),
  r("BM", sex = "M"),
  r("B1", "BM", "B", "M"), r("B2", "BM", "B", "F"),
  r("B1M", sex = "F"), r("B2M", sex = "M"),
  r("B1a", "B1", "B1M", "F"), r("B1b", "B1", "B1M", "M"),
  r("B2a", "B2M", "B2", "F"), r("B2b", "B2M", "B2", "M")
)
ped <- build_ped(rows)
forest <- nprcgenekeepr:::.buildMatingUnitForest(ped)

shipped <- nprcgenekeepr:::.positionMatingUnitForest(ped, forest, orderBySex = TRUE)
mySeed <- positionSeedOnly(ped, forest, orderBySex = TRUE)
mySeed <- mySeed[order(mySeed$id), ]
shippedOrd <- shipped[order(shipped$id), c("id", "x", "gen")]
cat("Synthetic example: max |x| diff =", max(abs(mySeed$x - shippedOrd$x)), "\n")
```

```{r crosscheck-real}
pedReal <- read.csv("../../inst/extdata/examples/obfuscated_rhesus_mhc_ped.csv",
                     stringsAsFactors = FALSE)
pedReal$id <- as.character(pedReal$id)
pedReal$sire <- as.character(pedReal$sire)
pedReal$dam <- as.character(pedReal$dam)
pedReal$gen <- findGeneration(pedReal$id, pedReal$sire, pedReal$dam)
forestReal <- nprcgenekeepr:::.buildMatingUnitForest(pedReal)

shippedReal <- nprcgenekeepr:::.positionMatingUnitForest(pedReal, forestReal, orderBySex = TRUE)
seedNodesReal <- positionSeedOnly(pedReal, forestReal, orderBySex = TRUE)
seedOrd <- seedNodesReal[order(seedNodesReal$id), ]
shipOrd <- shippedReal[order(shippedReal$id), ]
cat("Real 375-individual fixture: max |x| diff =",
    max(abs(seedOrd$x - shipOrd$x)), "\n")
```

Both are 0 -- the harness is faithful at both scales, so downstream comparisons are trustworthy.

## Synthetic example: candidate vs. shipped

```{r synth-render}
#| fig-width: 11
#| fig-height: 5.5
seedNodes <- positionSeedOnly(ped, forest, orderBySex = TRUE)
candNodes <- positionIgraphCandidate(ped, forest, orderBySex = TRUE, restarts = 20L, seed = 1L)

renderNodes <- function(nodes, forest, title) {
  childEdges <- forest$childEdges
  matingUnits <- forest$matingUnits
  xOf <- stats::setNames(nodes$x, nodes$id)
  genOf <- stats::setNames(nodes$gen, nodes$id)
  plot(nodes$x, -nodes$gen, type = "n", axes = FALSE, xlab = "", ylab = "",
       main = title, xlim = range(nodes$x) + c(-1, 1))
  for (i in seq_len(nrow(matingUnits))) {
    uid <- matingUnits$id[i]
    for (p in c(matingUnits$sire[i], matingUnits$dam[i])) {
      if (p %in% names(xOf)) {
        segments(xOf[[p]], -genOf[[p]], xOf[[uid]], -genOf[[uid]], col = "grey50")
      }
    }
  }
  for (i in seq_len(nrow(childEdges))) {
    fromId <- childEdges$from[i]; toId <- childEdges$to[i]
    if (fromId %in% names(xOf) && toId %in% names(xOf)) {
      segments(xOf[[fromId]], -genOf[[fromId]], xOf[[toId]], -genOf[[toId]], col = "grey20")
    }
  }
  realNodes <- nodes[!grepl("^__", nodes$id), ]
  points(realNodes$x, -realNodes$gen, pch = 1, cex = 2.2)
  text(realNodes$x, -realNodes$gen, labels = realNodes$id, cex = 0.7, pos = 1, offset = 0.8)
}

par(mfrow = c(1, 2))
gapS <- abs(diff(seedNodes$x[match(c("A", "B"), seedNodes$id)]))
gapC <- abs(diff(candNodes$x[match(c("A", "B"), candNodes$id)]))
renderNodes(seedNodes, forest, sprintf("SHIPPED -- A-B gap = %.2f", gapS))
renderNodes(candNodes, forest,
            sprintf("igraph sugiyama (20 restarts) -- A-B gap = %.2f, crossings = %d",
                    gapC, attr(candNodes, "crossings")))
```

The candidate closes the `A`-`B` gap from `r round(gapS,2)` to `r round(gapC,2)` (a
`r round(100*(1-gapC/gapS))`% reduction) -- matching the FIRST spike's own barycenter/median
candidate's exact toy-example improvement -- with **0 crossings**, achieved via a properly
published, proven crossing-minimization algorithm.

### Sensitivity: the natural vertex order alone hits an avoidable local optimum

```{r sensitivity-order}
seedS_small <- buildSeed(ped, forest)
g_small <- buildSugiyamaGraph(seedS_small)
res_natural <- igraph::layout_with_sugiyama(g_small, layers = igraph::V(g_small)$layer,
                                             hgap = 1, vgap = 1)
xOf_natural <- stats::setNames(res_natural$layout[, 1], igraph::V(g_small)$name)
nodes_natural <- finishPipeline(seedS_small, unname(xOf_natural[seedS_small$realIds]),
                                 orderBySex = TRUE)
cat("Natural vertex order (no restart): crossings =",
    countCrossings(nodes_natural, forest), "\n")
cat("Best of 20 random-order restarts: crossings =",
    attr(candNodes, "crossings"), "\n")
```

`igraph::layout_with_sugiyama()`'s crossing-minimization heuristic is sensitive to input vertex
order: the natural construction order (matching `.buildMatingUnitForest()`'s own id order) landed
on 4 avoidable crossings on this 13-individual example; a single random reorder found 0 on the
first trial, and 30/30 random trials in ad hoc testing found at least one 0-crossing layout.
Multi-restart (standard practice for layered-graph heuristics -- not a new algorithm, just
mitigating a known local-optimum risk in the proven one) is used for every result below.

## Real 375-individual fixture: the faithful full-pipeline measurement

Reusing the first spike's own methodology exactly: for each mating unit, each child edge's
`|childX - unitX|` against the FULL-pipeline `finalUnitX`, scaled by `xScale = 120`.

```{r real-baseline}
baselineMetric <- measureFaithful(seedNodesReal, forestReal)
baselineMetric
baselineCrossings <- countCrossings(seedNodesReal, forestReal)
cat("Baseline (shipped) crossings:", baselineCrossings, "\n")
```

This exactly reproduces the first spike's own published baseline (9/251 edges, 3.6%, max
4,121.25), confirming the harness is faithful.

```{r real-candidate}
t0 <- Sys.time()
candNodesReal <- positionIgraphCandidate(pedReal, forestReal, orderBySex = TRUE,
                                          restarts = 20L, seed = 1L)
candTime <- as.numeric(Sys.time() - t0, units = "secs")
cat("Candidate compute time (20 restarts, 375 individuals):", round(candTime, 2), "s\n")

candidateMetric <- measureFaithful(candNodesReal, forestReal)
candidateMetric
cat("Candidate crossings:", attr(candNodesReal, "crossings"), "\n")
```

```{r real-comparison}
cat(sprintf("Baseline:  %d/%d edges >200 units (%.2f%%), max offset %.2f, crossings %d\n",
            baselineMetric[["n_over_200"]], baselineMetric[["n"]],
            baselineMetric[["pct_over_200"]], baselineMetric[["max_offset_scaled"]],
            baselineCrossings))
cat(sprintf("Candidate: %d/%d edges >200 units (%.2f%%), max offset %.2f, crossings %d\n",
            candidateMetric[["n_over_200"]], candidateMetric[["n"]],
            candidateMetric[["pct_over_200"]], candidateMetric[["max_offset_scaled"]],
            attr(candNodesReal, "crossings")))

realSeed <- seedNodesReal[!grepl("^__", seedNodesReal$id), ]
realCand <- candNodesReal[!grepl("^__", candNodesReal$id), ]
cat(sprintf("\nOverall real-node x range -- baseline: %.1f, candidate: %.1f (%.1fx)\n",
            diff(range(realSeed$x)), diff(range(realCand$x)),
            diff(range(realCand$x)) / diff(range(realSeed$x))))
```

**The candidate regresses the real fixture on every axis measured**: 25/251 edges exceed the
200-unit threshold (9.96%, vs. baseline 3.6%), the worst single offset grows from 4,121 to 10,110,
the overall layout width for real individuals more than doubles, and -- notably, since crossing
minimization is this algorithm's own stated objective -- **crossings are also worse than
baseline** (5,916 vs. 3,174), even after picking the best of 20 restarts.

### Sensitivity: restart count, seed, and edge weights -- not a tuning artifact

```{r hp-sweep}
sweepResults <- lapply(list(
  list(restarts = 5, seed = 1), list(restarts = 10, seed = 2),
  list(restarts = 40, seed = 3), list(restarts = 40, seed = 99)
), function(p) {
  cand <- positionIgraphCandidate(pedReal, forestReal, orderBySex = TRUE,
                                   restarts = p$restarts, seed = p$seed)
  m <- measureFaithful(cand, forestReal)
  data.frame(restarts = p$restarts, seed = p$seed,
             crossings = attr(cand, "crossings"),
             n_over_200 = m[["n_over_200"]], pct = round(m[["pct_over_200"]], 2),
             max_offset = round(m[["max_offset_scaled"]], 1))
})
do.call(rbind, sweepResults)
```

Every restart-count/seed combination regresses relative to the 9/251 (3.6%) baseline -- never
better than 25/251 (9.96%). A separate check confirmed `weights` passed to
`layout_with_sugiyama()` (tried favoring union-to-child edges up to 10:1 over parent-to-union
edges, intended to encourage tighter full-sibling grouping) had **zero measurable effect** on the
resulting layout in this implementation when explicit `layers` are also supplied -- a real,
checked limitation of this specific knob for this use case, not exploited further.

### Why it fails: sugiyama's own objective doesn't preserve full-sibling compactness

```{r hub-diagnostic}
mu <- forestReal$matingUnits
xOf <- stats::setNames(candNodesReal$x, candNodesReal$id)
worstOff <- -1; worstU <- NULL
for (i in seq_len(nrow(mu))) {
  uid <- mu$id[i]
  kids <- intersect(forestReal$childEdges$to[forestReal$childEdges$from == uid], names(xOf))
  if (!length(kids)) next
  for (k in kids) {
    off <- abs(xOf[[k]] - xOf[[uid]]) * xScale
    if (off > worstOff) { worstOff <- off; worstU <- uid }
  }
}
worst <- mu[mu$id == worstU, ]
worst
cat("Sire", worst$sire, "own mate count:",
    sum(mu$sire == worst$sire | mu$dam == worst$sire), "\n")
kids <- intersect(forestReal$childEdges$to[forestReal$childEdges$from == worstU], names(xOf))
cat("This union's own children's x:", paste(round(xOf[kids], 2), collapse = ", "), "\n")
```

The single worst-regressed edge again traces to a hub individual (**4 separate mating unions** --
a different individual from the first spike's own hub, but the same STRUCTURAL feature). Unlike
the first spike's convergence-instability failure, here the mechanism is different: sugiyama's
crossing-minimization and horizontal-coordinate-assignment optimize for overall graph
straightness/crossing count across the WHOLE graph, with no explicit preference for keeping one
mating unit's own full siblings close together -- a hub individual's several subtrees compete for
horizontal space across the whole layout, and the algorithm can (and does) scatter one union's
full-sibling group widely if doing so serves its own global objective. The shipped rigid-subtree
model achieves tight full-sibling grouping BY CONSTRUCTION (each subtree recursively reserves a
compact, contiguous span); a proven algorithm optimizing a genuinely different objective does not
inherit that property for free.

## Conclusion

**This is the THIRD independently-designed candidate** (after the sibling design session's
bounded-lookahead candidate, S588, and the first spike's barycenter/median candidate, S589) **to
show the same qualitative pattern**: improves the small synthetic example, regresses the real
375-individual fixture -- via a THIRD distinct failure mechanism each time (edge crossings;
convergence instability; full-sibling scattering under a proven library's own global objective).
Not a tuning artifact: confirmed across a 4-point restart/seed sweep and an edge-weight check that
had no effect. See the spike's own plan document for the resulting verdict.
