---
title: "Interactive Use of nprcgenekeepr"
subtitle: An R Package for the Genetic Management of Colonies
output:
  html_document:
    df_print: paged
  pdf_document: default
vignette: >
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteIndexEntry{Interactive Use of nprcgenekeepr}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, include = TRUE, results = "markup",
  cache = FALSE
)
pdf.options(useDingbats = TRUE)
start_time <- proc.time()
```

```{r, include = FALSE}
library(stringi)
# library(rmsutilityr)
suppressMessages(library(ggplot2))
library(knitr)
library(nprcgenekeepr)
set_seed(1L)
```

## Introduction
This tutorial demonstrates the major functions used 
within the Shiny application provided by the __nprcgenekeepr__ package and
provides sufficient insight into those functions that they may be used
independently.

This tutorial is primarily directed toward someone with experience using
R who wants to better understand how the Shiny application works or 
to perform some actions not directly supported by the Shiny application.

Please provide any comments, questions, or bug reports through the GitHub
issue tracker at 
[https://github.com/rmsharp/nprcgenekeepr/issues](
https://github.com/rmsharp/nprcgenekeepr/issues).


## Installation and Help

You can install **nprcgenekeepr** from GitHub with the following code.

```{r gh-installation, eval = FALSE, echo=TRUE}
install.packages(nprcgenekeepr)
## Use the following code to get the development version
# install.packages("devtools")
# devtools::install_github("rmsharp/nprcgenekeepr")
```

All missing dependencies should be automatically installed.

You can get help from the __R__ console with 

```{r gh-help, eval = FALSE, echo=TRUE}
?nprcgenekeepr
```
__The help provided by this (_nprcgenekeepr.R_) needs to be more complete 
and include links to the tutorials.__

## Reading in a Pedigree

A pedigrees can be imported using either Excel worksheets or text files that 
contain all of the pedigree information or using either Excel worksheets or 
text files that contain a list of focal animals with the remainder of the 
pedigree information is pulled in through the LabKey API.

This tutorial will use a pedigree file that can be created using the 
__makeExamplePedigreeFile__ function as shown below.
The function __makeExamplePedigreeFile__ both saves a file and returns the
full path name to the saved file, which we are saving into the variable 
_pedigreeFile_.
Note: the user will select where to store the file.

```{r make-example-pedigree, eval = FALSE, include = TRUE}
library(nprcgenekeepr)
pedigreeFile <- makeExamplePedigreeFile()
```
This writes _ExamplePedigree.csv_ to a place you select within your file
system.

```{r get-file-name, eval = TRUE, echo = FALSE, include = FALSE}
pedigreeFile <- "../inst/extdata/examples/ExamplePedigree.csv"
```

You use the file name provided by the __makeExamplePedigreeFile__ function
to tell __read.table__ what file to read.

```{r read-in-example-pedigree}
breederPedCsv <- read.table(pedigreeFile,
  sep = ",", header = TRUE,
  stringsAsFactors = FALSE
)

breederPedCsv$fromCenter <- "TRUE"
breederPedCsv$fromCenter[
  suppressWarnings(sample(
    which(is.na(breederPedCsv$sire) &
      is.na(breederPedCsv$dam)),
    round(0.8 * length(which(
      is.na(breederPedCsv$sire) &
        is.na(breederPedCsv$dam)
    )))
  ))
] <- "FALSE"
```
Note the number of rows read. Each row represents an individual within the 
pedigree.
```{r row-count}
nrow(breederPedCsv)
```

The next step is to put the information read from the file into a 
pedigree object. This is done with the __qcStudbook__ function, which examines 
the file contents and tests for common pedigree errors.

You can see the errors that can be detected by __qcStudbook__ by returning the
empty error list with __getEmptyErrorLst()__. 
We are not showing the output of the function call now 
because later in this tutorial we will explore errors in more depth.

__qcStudbook__ can take five arguments _sb_, _minSireAge_ and _minDamAge_
(in years), _reportChanges_,
and _reportErrors_. However, all but _sb_ have default values and only the
_sb_ argument is required.

It is prudent to ensure that parents are at least of breeding age, which
is species and sex specific.
Leaving _minSireAge_ and _minDamAge_ blank (their default, `NULL`) looks up the
species-specific breeding age for each sex; a number overrides that sex's
floor. I have set both to 2 years here.[^1]

[^1]: Setting _minDamAge_ to 3.5 and above will cause an error along
with the creation of a file _~/lowParentAge.csv_ that will list the parents
with low age at the birth of an offspring: two dams in this pedigree were
about 3.3 years old at a birth. The sires are all older, so raising
_minSireAge_ alone to 3.5 does not trigger the error -- exactly the
sex-specific control the two floors provide.

```{r form-pedigree-object}
breederPed <- qcStudbook(breederPedCsv, minSireAge = 2L, minDamAge = 2L)
```

If __qcStudbook__ reports an error, change the call by adding the 
__reportErrors__ argument set to __TRUE__ and examine the returned object.
More on this is presented in the __Pedigree Errors__ section.

## Identifying Focal Animals

You may want to focus your work on a _focal_ group of animals. 
This can be done by reading in a list of animal IDs that make up the _focal_
group and use that list to update the pedigree.
Alternatively you can created a list of animal IDs based on 
criteria you have selected. 

For example, to select living animals at the facility with at least one parent,
the following can be used.

```{r select-focal-animals-from-pedigree, results='asis'}
focalAnimals <- breederPed$id[!(is.na(breederPed$sire) &
  is.na(breederPed$dam)) &
  is.na(breederPed$exit)]
print(stri_c(
  "There are ", length(focalAnimals),
  " animals in the vector _focalAnimals_."
))
```

As can be seen, these animals have at least one parent and have not left the
facility.

```{r short-pedigree-list-of-focal-animals}
breederPed[breederPed$id %in%
  focalAnimals, c("id", "sire", "dam", "exit")][1L:10L, ]
```

We indicate that these are the animals of interest by using the
__setPopulation__ function. 
This function simply sets a column named _population_[^2] to the logical value
of __TRUE__ if the row represents an animal in the list and __FALSE__ otherwise.

[^2]: The _population_ column is created and added to the pedigree object if it 
does not already exist.

The first line of code below sets the _population_ column and the second 
counts the number of rows where the value was set to __TRUE__.

```{r set-population-flag}
breederPed <- setPopulation(ped = breederPed, ids = focalAnimals)
nrow(breederPed[breederPed$population, ])
```

The IDs used to populate the _population_ flag can be used to trim the pedigree 
so that it contains only those 
individuals who are in the ID list or are ancestors of those individuals.

```{r trim-pedigree}
trimmedPed <- trimPedigree(focalAnimals, breederPed)
nrow(breederPed)
nrow(trimmedPed)
```

The __trimPedigree__ function has the ability to remove those ancestors 
that do not contribute genetic information.
Uninformative founders are those individuals who are parents of only
one individual and who have no parental information. 
(_Currently genotypic information is ignored by __trimPedigree___).

```{r removed-uninformatiive-animals}
trimmedPedInformative <- trimPedigree(focalAnimals, breederPed,
  removeUninformative = TRUE
)
nrow(trimmedPedInformative)
```

We can find all of the animals that are in the trimmed pedigree but are not
focal animals.
```{r get-animals-added-to-focal-animals}
nonfocalInTrimmedPed <- trimmedPed$id[!trimmedPed$id %in% focalAnimals]
length(nonfocalInTrimmedPed)
```

We can see which of these `r length(nonfocalInTrimmedPed)` are and are not
parents.
We will first make sure we have all of the parents by getting our list of
parents from the entire pedigree. 
We then demonstrate that they are all in the trimmed pedigree.
```{r find-focal-parents}
allFocalParents <- c(
  breederPed$sire[breederPed$id %in% focalAnimals],
  breederPed$dam[breederPed$id %in% focalAnimals]
)
trimmedFocalParents <- c(
  trimmedPed$sire[trimmedPed$id %in% focalAnimals],
  trimmedPed$dam[trimmedPed$id %in% focalAnimals]
)
all.equal(allFocalParents, trimmedFocalParents) # Are the IDs the same?
```

However, not all of the animals in the trimmed pedigree are either
the focal animals or their parents. They are more distant ancestors
as we will show.

```{r find-non-focal-non-focal-parent}
notFocalNotParent <-
  trimmedPed$id[!trimmedPed$id %in% c(focalAnimals, allFocalParents)]
length(notFocalNotParent)
```


```{r find-grandparents, echo = FALSE, include=FALSE, eval = TRUE}
allFocalGrandParents <- c(
  breederPed$sire[breederPed$id %in% allFocalParents],
  breederPed$dam[breederPed$id %in% allFocalParents]
)
## Not all parents are known so the unknown individuals (NA) are removed.
allFocalGrandParents <- allFocalGrandParents[!is.na(allFocalGrandParents)]
trimmedFocalGrandParents <-
  c(
    trimmedPed$sire[trimmedPed$id %in% allFocalParents],
    trimmedPed$dam[trimmedPed$id %in% allFocalParents]
  )
trimmedFocalGrandParents <-
  trimmedFocalGrandParents[!is.na(trimmedFocalGrandParents)]
all.equal(allFocalGrandParents, trimmedFocalGrandParents)
```

Since the trimming process is supposed to retain the focal animals
and their ancestors, we will leave it as an exercise for you to demonstrate
that at least some of the remaining animals are grandparents of the focal
animals. _Hint: there are `r length(trimmedFocalGrandParents)` grandparents
in both the trimmed and the complete pedigree_.

```{r added-animals, echo=FALSE, include=FALSE, eval=TRUE}
# get-informative-and-uninformative-added-animals
trimmedPedInformative <- trimPedigree(focalAnimals, breederPed,
  removeUninformative = TRUE
)
uninformative <- trimmedPed$id[!trimmedPed$id %in% trimmedPedInformative$id]
notFocalInTrimmedPed <- trimmedPed$id[!trimmedPed$id %in% focalAnimals]
additionalAnimals <- nrow(trimmedPed) - length(focalAnimals)
geneticallyInformative <-
  nrow(trimPedigree(focalAnimals, breederPed,
    removeUninformative = TRUE
  )) -
  length(focalAnimals)
```

As you can see from the number of rows in the full pedigree 
(`r nrow(breederPed)`) versus the trimmed pedigree (`r nrow(trimmedPed)`),
trimmed pedigrees can be much smaller. 
Of the additional `r additionalAnimals` animals, 
`r geneticallyInformative`
provide genetic information while the others 
(`r length(uninformative)`) are genetically uninformative.

```{r animals-no-birth-no-exit, echo = FALSE, eval = TRUE, include=FALSE}
unknownBirth <- breederPed$id[is.na(breederPed$birth)]
unknownBirthOrExit <-
  breederPed$id[is.na(breederPed$birth) | !is.na(breederPed$exit)]
knownPed <- breederPed[!breederPed$id %in% unknownBirthOrExit, ]
otherIds <- knownPed$id[!knownPed$id %in% trimmedPed$id[is.na(trimmedPed$exit)]]
```

As is shown below only `r length(otherIds)` (`r get_and_or_list(otherIds)`)
living animals are still in the colony but are not in the trimmed pedigree.[^3]

[^3]: All animals within the colony have a known birth date.

```{r show-living-animals-not-in-trimmed-pedigree, results = 'asis'}
unknownBirth <- breederPed$id[is.na(breederPed$birth)]
knownExit <- breederPed$id[!is.na(breederPed$exit)]
unknownBirthKnownExit <-
  breederPed$id[is.na(breederPed$birth) | !is.na(breederPed$exit)]
knownPed <- breederPed[!breederPed$id %in% unknownBirthKnownExit, ]
otherIds <- knownPed$id[!knownPed$id %in% trimmedPed$id[is.na(trimmedPed$exit)]]
print(stri_c(
  "The living animals in the pedigree that are not in the trimmed ",
  "pedigree are ", get_and_or_list(otherIds), "."
))
```

## Age Sex Pyramid Plot

You can examine the population structure using an age-sex pyramid plot with a 
single function. 
We will limit our view to just the focal animals and their living relatives.
This is appropriate for colony management because in addition to the genetic
diversity we seek, we have to remain cognizant of the age and sex distributions
within the colonies we manage.

```{r plot-focal-age-sex-pyramid, include = TRUE}
#| fig.alt: >
#|   Age Sex Pyramid Plot.
getPyramidPlot(ped = trimmedPed[is.na(trimmedPed$exit), ])
```

## Pedigree Diagram

The Pedigree Browser's Diagram tab in the Shiny application renders a
pedigree as a family-tree diagram. A mate's own mating(s) render as a small
connector between the two parents, with a line down to their shared
children, rather than two independent lines running straight from each
parent -- the same convention traditional pedigree charts use. An animal
that mates more than once, or whose lineage loops back on itself (e.g. a
consanguineous mating), appears once per mating, with each occurrence after
the first joined back to its main occurrence by a dashed line. Both are
built on two script-callable functions that work identically outside the
Shiny application: __makePedigreeMatingLayout__ prepares a pedigree's
node/edge data in this convention, and __visNetwork::visNetwork__ (from the
__visNetwork__ package, a dependency of __nprcgenekeepr__) renders it. (A
simpler function, __makePedigreeDiagramData__, remains available for a plain
one-node-per-animal diagram without this convention -- it is no longer what
the Shiny app itself uses.) See the _pedigree-diagram_ article for the
equivalent point-and-click workflow and screenshots of every feature --
including affected-status shading, name labels, and twin/zygosity
connectors, none of which this scripted walkthrough demonstrates.

That parent-to-mating-to-child connector has two available routings,
selected by __makePedigreeMatingLayout__'s _edgeStyle_ argument:
`"rectilinear"` (the default, Track 2, docs/planning/pedigree-diagram-
kinship2-fidelity-remediation-plan.md -- strict horizontal/vertical right
angles throughout: a horizontal segment between the two parents, a vertical
drop to the mating dot, a horizontal sibship bar across the children, and a
vertical drop into each one, matching the convention the __kinship2__ R
package uses) and `"direct"` (a single straight or lightly sloped line all
the way from parent to mating dot to child). The Shiny app's Diagram tab
exposes the same choice live as a "Diagram Edge Style" toggle above the
diagram; in a script it is just this one argument. Both are demonstrated
below -- _direct_ first, _rectilinear_ (matching the tab's own default)
afterward.

The _trimmedPed_ pedigree built earlier in this tutorial has
`r nrow(trimmedPed)` animals -- realistic for the live Shiny application's
pan-and-zoom canvas, but too dense for a single static demonstration to
usefully show every feature at once. It is also limited to the _Female_,
_Male_, and _Unknown_ sex codes that happen to occur in the bundled example
data -- the _Hermaphrodite_ and unrecorded-sex shapes never appear in it. So
instead we build a small, synthesized pedigree spanning a few generations,
deliberately including all five sex codes __makePedigreeMatingLayout__
recognizes -- and, since two of its founders (_M1_ and _F2_) each mate more
than once, the diagram's duplicate-node convention as well.[^4]

[^4]: This pedigree is entirely synthetic -- constructed for this tutorial,
not drawn from any real colony's records.

```{r pedigree-diagram-demo-data}
founders <- data.frame(
  id = c(paste0("F", 1L:4L), paste0("M", 1L:4L)),
  sire = NA_character_, dam = NA_character_,
  sex = c(rep("F", 4L), rep("M", 4L)),
  stringsAsFactors = FALSE
)
gen1 <- data.frame(
  id = paste0("C", 1L:12L),
  sire = c(
    "M1", "M1", "M1", "M2", "M2", "M3", "M3", "M3", "M4", "M4", "M1", "M1"
  ),
  dam = c(
    "F1", "F1", "F1", "F2", "F2", "F3", "F3", "F3", "F4", "F4", "F2", "F2"
  ),
  sex = c("F", "M", "H", "F", "M", "M", "F", "U", "F", "M", "F", "M"),
  stringsAsFactors = FALSE
)
gen2 <- data.frame(
  id = paste0("D", 1L:9L),
  sire = c("C2", "C2", "C5", "C5", "C5", "C6", "C10", "C10", "C12"),
  dam = c("C4", "C4", "C7", "C7", "C7", "C1", "C9", "C9", "C11"),
  sex = c("F", "M", "M", "F", "M", "F", "M", "F", "F"),
  stringsAsFactors = FALSE
)
gen3 <- data.frame(
  id = paste0("E", 1L:4L),
  sire = c("D2", "D2", "D5", "D7"),
  dam = c("D4", "D4", "D8", "D6"),
  sex = c("F", "M", "U", NA_character_),
  stringsAsFactors = FALSE
)
demoPed <- rbind(founders, gen1, gen2, gen3)
demoPed$gen <- findGeneration(demoPed$id, demoPed$sire, demoPed$dam)
nrow(demoPed)
table(demoPed$sex, useNA = "ifany")
```

Sex code __H__ (_C3_) and the unrecorded sex code (_E4_, `NA`) each appear
exactly once, alongside code __U__ (_C8_ and _E3_) and the ordinary __F__/
__M__ codes that make up the rest -- enough for every one of the five node
shapes to appear in the diagram below at least once, across
`r length(unique(demoPed$gen))` generations and `r nrow(demoPed)` animals.

### Direct Edge Style

```{r pedigree-diagram-data}
## edgeStyle is now explicit -- Track 2 flipped the package's own default
## to "rectilinear", so this section's own "Direct Edge Style" labeling
## needs edgeStyle = "direct" spelled out rather than relying on the
## (now different) implicit default.
diagramData <- makePedigreeMatingLayout(demoPed, edgeStyle = "direct")
names(diagramData)
nrow(diagramData$nodes)
nrow(diagramData$edges)
```

__makePedigreeMatingLayout__ returns a list of three elements: _nodes_ (one
row per real animal -- shaped by sex, dot = Female, square = Male, star =
Hermaphrodite, triangle = Unknown, diamond = Other/Unrecorded, with an HTML
hover-tooltip giving ID, sex, generation, sire, and dam -- plus one small,
unlabeled dot per mating and one extra row per duplicate occurrence, each
carrying its animal's own shape, label, and tooltip with a note that it is a
duplicate occurrence), _edges_ (parent-to-mating and mating-to-child edges,
plus a dashed edge from each duplicate occurrence back to the real animal),
and _duplicateToReal_ (a named lookup from each duplicate occurrence's id
back to the real animal it represents).

Each node already carries its own fixed __x__/__y__ position (computed by
__makePedigreeMatingLayout__ itself), so rendering it turns vis.js's usual
physics/hierarchical layout off rather than turning one on. Piping
_diagramData_ into __visNetwork::visNetwork__ and the same layout, export,
legend, and search options `R/modPedigree.R` uses reproduces the Diagram
tab's rendering exactly, including a genuinely working __Export Diagram
(PNG)__ button -- try it below.[^5]

[^5]: The Diagram tab's click-to-navigate behavior (clicking a node narrows
the focal-animal selection to it, resolving a duplicate-occurrence click to
the real animal and ignoring mating-node clicks) is Shiny-specific -- it
relies on a live Shiny session to receive the click event -- and is
therefore not reproduced here.

```{r pedigree-diagram-render}
visNetwork::visNetwork(diagramData$nodes, diagramData$edges) |>
  visNetwork::visPhysics(enabled = FALSE) |>
  visNetwork::visNodes(physics = FALSE) |>
  visNetwork::visEdges(smooth = FALSE) |>
  visNetwork::visExport(
    type = "png", name = "pedigree_diagram",
    label = "Export Diagram (PNG)"
  ) |>
  visNetwork::visLegend(
    addNodes = data.frame(
      label = c(
        "Female", "Male", "Hermaphrodite", "Unknown", "Other / Unrecorded"
      ),
      shape = c("dot", "square", "star", "triangle", "diamond"),
      stringsAsFactors = FALSE
    ),
    useGroups = FALSE, position = "right", main = "Sex",
    width = 0.28, stepY = 65L
  ) |>
  visNetwork::visOptions(
    nodesIdSelection = list(
      enabled = TRUE,
      # Excludes all 5 of makePedigreeMatingLayout()'s reserved node-id
      # prefixes -- __union_/__dup_ can appear under either edge style;
      # __drop_/__bar_/__proj_ only appear under "rectilinear" (see below).
      # Harmless no-op here since diagramData was built with edgeStyle =
      # "direct" explicit above, but keeps this pattern identical to the
      # rectilinear chunk below and to R/modPedigree.R's own.
      values = diagramData$nodes$id[
        !grepl("^__union_|^__dup_|^__drop_|^__bar_|^__proj_",
               diagramData$nodes$id)
      ]
    ),
    highlightNearest = list(enabled = TRUE, hover = TRUE, degree = 1L,
                             algorithm = "all")
  )
```

The __Select by id__ dropdown above the diagram -- filtered to real animals,
excluding mating and duplicate-occurrence nodes -- lets you jump straight to a
specific animal and dims every node except it and its direct connections --
useful for locating one animal in a diagram this size, and just as useful in
a real diagram of `r nrow(trimmedPed)` animals like the one built earlier in
this tutorial, where finding one animal by eye is much harder.

### Rectilinear Edge Style

The diagram above uses the _direct_ edge style, requested explicitly.
Omitting `edgeStyle` entirely (or passing `edgeStyle = "rectilinear"`
explicitly, as below) draws the same relationships -- same _demoPed_, same
animals, same parent/child/mating structure -- with strict right angles: a
horizontal line directly between mates, a vertical drop to their shared
mating dot, a horizontal bar across their children, and a vertical drop
into each child, matching the convention the __kinship2__ R package uses,
and matching __makePedigreeMatingLayout__'s own default (Track 2).

```{r pedigree-diagram-rectilinear-data}
diagramDataRectilinear <- makePedigreeMatingLayout(demoPed,
                                                     edgeStyle = "rectilinear")
names(diagramDataRectilinear)
nrow(diagramDataRectilinear$nodes)
nrow(diagramDataRectilinear$edges)
```

`r nrow(diagramDataRectilinear$nodes)` nodes and
`r nrow(diagramDataRectilinear$edges)` edges here, versus
`r nrow(diagramData$nodes)` and `r nrow(diagramData$edges)` for the same
`r nrow(demoPed)`-animal _demoPed_ under the _direct_ style above -- the
difference is entirely extra invisible "waypoint" nodes/edges that carry
out the right-angle routing (zero size, transparent color, excluded from
the __Select by id__ dropdown below just like the mating dots above), not
any change to which animals or relationships are shown.[^6] _nodes_ and
_edges_ also gain extra `color.background`/`color.border` (nodes) and
`color` (edges) columns under _rectilinear_ -- __visNetwork::visNetwork()__
reads these directly, so no extra rendering code is needed; they exist
because vis.js otherwise defaults every waypoint-touching edge to
inheriting its parent node's border color, which would route the new right
angles in the wrong color.

[^6]: These waypoint nodes are internal plumbing needed for the right-angle
routing itself, not new diagram content.

Rendering it uses the identical chain as the _direct_ style above, with one
difference: __highlightNearest__'s hover-highlight _degree_ is raised from
`1` to `6`. Under _rectilinear_, a plain individual's nearest graph
neighbor is often one of the invisible waypoint nodes above rather than a
visible mating dot, so `degree = 1` (correct for _direct_) can highlight
nothing visible at all on hover; `6` restores visible feedback for the
common case. The live Shiny app's Diagram tab applies this same
style-aware degree automatically when you switch the "Diagram Edge Style"
toggle.

```{r pedigree-diagram-rectilinear-render}
visNetwork::visNetwork(diagramDataRectilinear$nodes,
                        diagramDataRectilinear$edges) |>
  visNetwork::visPhysics(enabled = FALSE) |>
  visNetwork::visNodes(physics = FALSE) |>
  visNetwork::visEdges(smooth = FALSE) |>
  visNetwork::visExport(
    type = "png", name = "pedigree_diagram_rectilinear",
    label = "Export Diagram (PNG)"
  ) |>
  visNetwork::visLegend(
    addNodes = data.frame(
      label = c(
        "Female", "Male", "Hermaphrodite", "Unknown", "Other / Unrecorded"
      ),
      shape = c("dot", "square", "star", "triangle", "diamond"),
      stringsAsFactors = FALSE
    ),
    useGroups = FALSE, position = "right", main = "Sex",
    width = 0.28, stepY = 65L
  ) |>
  visNetwork::visOptions(
    nodesIdSelection = list(
      enabled = TRUE,
      values = diagramDataRectilinear$nodes$id[
        !grepl("^__union_|^__dup_|^__drop_|^__bar_|^__proj_",
               diagramDataRectilinear$nodes$id)
      ]
    ),
    highlightNearest = list(enabled = TRUE, hover = TRUE, degree = 6L,
                             algorithm = "all")
  )
```

The live Shiny app also caps how large a pedigree it will render as a
diagram, and the cap itself is style-aware: 750 animals under _direct_,
400 under _rectilinear_ -- because the rectilinear style's extra waypoint
nodes make an equivalent-size diagram roughly 1.9x heavier to render. Both
caps are internal to the Shiny app (`R/modPedigree.R`); the script-callable
functions demonstrated here have no such limit of their own.

### Twin/Zygosity Connectors

A recorded pedigree has no way to represent that two siblings are twins,
or how confident that twin call is -- that information has to come from
outside the pedigree, typically a curator-maintained sidecar file.
__readTwinRelations__ reads such a file (Excel or delimited text, columns
_id1_, _id2_, _code_) with no validation of its own; __checkTwinRelations__
then checks the result against the pedigree -- both ids must exist, an
_MZ twin_ (monozygotic) or _DZ twin_ (dizygotic) pair must already share
both recorded parents, and an _MZ twin_ pair must additionally match in
recorded sex, while a _UZ twin_ (zygosity undetermined) pair has no such
precondition. Only the validated result is accepted by
__makePedigreeMatingLayout__'s optional _twinRelations_ argument, which
adds a dashed twin-connector edge between the pair, styled by zygosity
code -- the same connectors shown in the live app's Diagram tab.

```{r twin-relations-setup}
twinPed <- data.frame(
  id = c("F1", "F2", "S1", "S2"),
  sire = c(NA, NA, "F1", "F1"),
  dam = c(NA, NA, "F2", "F2"),
  sex = c("M", "F", "F", "M"),
  stringsAsFactors = FALSE
)
twinPed$gen <- findGeneration(twinPed$id, twinPed$sire, twinPed$dam)
twinRelationsCsv <- tempfile(fileext = ".csv")
writeLines(c("id1,id2,code", "S1,S2,UZ twin"), twinRelationsCsv)
```

_S1_ and _S2_ are full siblings (both recorded parents are _F1_/_F2_), so
a _UZ twin_ call between them passes validation:

```{r twin-relations-demo}
twinRelations <- checkTwinRelations(twinPed, readTwinRelations(twinRelationsCsv))
twinRelations
twinDiagramData <- makePedigreeMatingLayout(twinPed, twinRelations = twinRelations)
twinDiagramData$edges
```

Compare the _edges_ table above to what __makePedigreeMatingLayout__
returns for the same pedigree without _twinRelations_: the union/parent
-child edges are identical, but only the version above has the extra row
connecting _S1_ directly to _S2_ (dashed, labeled `"?"` for the
undetermined zygosity code) -- that row exists only because a validated
_twinRelations_ table was supplied.

## Genetic Value Analysis

Your genetic value analysis must be carefully performed. 
The next three commands set up the entire pedigree for analysis.
The first of these three commands set all of the pedigree members to
be part of the population of interest by setting the _population_
column to __TRUE__ for all individuals.

```{r set-entire-pedigree-as-population}
ped <- setPopulation(breederPed, NULL)
```

Note that a new pedigree object (__ped__) is being created.

```{r full-pedigree-genetic-value-summary}
probands <- ped$id[ped$population]
ped <- trimPedigree(probands, ped,
  removeUninformative = FALSE,
  addBackParents = FALSE
)
```

The arguments to __reportGV__ are all optional except for _ped_, but
you may often want to use non-default values.

- __ped__ Pedigree information in data.frame format

- __guIter__  Integer indicating the number of iterations for the gene-drop
analysis. Default is 1000 iterations

- __guThresh__ Integer indicating the threshold number of animals for 
defining a unique allele. Default considers an allele "unique" if it is found
in only 1 animal.

- __pop__ Character vector with animal IDs to consider as the population of 
interest. The default is NULL.

- __byID__ Logical variable of length 1 that is passed through to eventually 
be used by alleleFreq(), which calculates the count of each allele in the 
provided vector. If byID is TRUE and ids are provided, the function will 
only count the unique alleles for an individual (homozygous alleles will 
be counted as 1).

```{r full-geneticValue}
geneticValue <- reportGV(ped,
  guIter = 50L,
  guThresh = 3L,
  byID = TRUE,
  updateProgress = NULL
)
summary(geneticValue)
```

What happens if we limit our analysis to the trimmed pedigree?
Remember that the trimmed pedigree still contains all of the genetic
information that the full pedigree has for the focal animals.

```{r trimmed-genetic-value-analysis}
trimmedGeneticValue <- reportGV(trimmedPed,
  guIter = 50L,
  guThresh = 3L,
  byID = TRUE,
  updateProgress = NULL
)
summary(trimmedGeneticValue)
```

It is clear that limiting your analysis to the animals of interest
reduces the effort required to examine the animals of interest.

### Detailed look at the Genetic Value Report object

The names of the object within the genetic value report object 
(_trimmedGeneticValue_) can be listed as shown in the next line of code.

```{r list-genetic-value-objects}
names(trimmedGeneticValue)
```

The _report_ object (an R dataframe) can in-turn be examined.

```{r list-report-object-parts}
names(trimmedGeneticValue$report) ## column names
nrow(trimmedGeneticValue$report) ## Number of rows
```

The report is more conveniently used as a separate object.
The next section of code rounds some of the numerical values and converts all
columns to characters for display as a table where only the first 10 lines
are included.

```{r look-at-genetic-value-report}
rpt <- trimmedGeneticValue[["report"]]
rpt$indivMeanKin <- round(rpt$indivMeanKin, 5L)
rpt$zScores <- round(rpt$zScores, 2L)
rpt$gu <- round(rpt$gu, 5L)
rpt <- toCharacter(rpt)
names(rpt) <- headerDisplayNames(names(rpt))
knitr::kable(rpt[1L:10L, ]) # needs more work for display purposes.
```

We start the next lines of code by getting a fresh copy of the genetic value
report since we changed all of the numeric values to characters in the last 
section to print the table.
These lines demonstrate one way of extracting the component objects from the
_genetic value_ object.

```{r kinship-and-founders}
rpt <- trimmedGeneticValue[["report"]]
kmat <- trimmedGeneticValue[["kinship"]]
f <- trimmedGeneticValue[["total"]]
mf <- trimmedGeneticValue[["maleFounders"]]
ff <- trimmedGeneticValue[["femaleFounders"]]
nmf <- trimmedGeneticValue[["nMaleFounders"]]
nff <- trimmedGeneticValue[["nFemaleFounders"]]
fe <- trimmedGeneticValue[["fe"]]
fg <- trimmedGeneticValue[["fg"]]
```

It is informative to examine the distribution of _genetic uniqueness_, 
_mean kinship_, and _z-scores_ (normalized _mean kinship_ values).

Creation of the boxplot for the _genetic uniqueness_ values is shown below.

```{r genetic-uniqueness-boxplot}
#| fig.alt: >
#|   Genetic Uniqueness Box Plot.
gu <- rpt[, "gu"]
guBox <- ggplot(data.frame(gu = gu), aes(x = "", y = gu)) +
  geom_boxplot(
    color = "darkblue",
    fill = "lightblue",
    notch = TRUE, #| fig.alt: >
    #|   Histogram of time between eruptions for Old Faithful.
    #|   It is a bimodal distribution with peaks at 50-55 and
    #|   80-90 minutes.

    outlier.color = "red",
    outlier.shape = 1L
  ) +
  theme_classic() +
  geom_jitter(width = 0.2) +
  coord_flip() +
  ylab("Score") +
  ggtitle("Genetic Uniqueness")
print(guBox)
```

Extraction of the individual _mean kinship_ values and their corresponding 
z-scores is shown in the next code chunk. 

```{r extraction-of-mk-zs}
mk <- rpt[, "indivMeanKin"]
zs <- rpt[, "zScores"]
```

Creation of boxplots for the _mean kinship_ and _z-scores_ is left
as an exercise.

## Breeding Group Formation

The primary purpose of __nprcgenekeepr__ is to form breeding groups 
according to our best information regarding maintaining the genetic
characteristics we desire and the realities associated with other animal
husbandry needs.

There are several options you must consider when forming groups using
__nprcgenekeepr__, which we will examine using code below.

-   Animals used to form groups
    -   _high-value_: Randomly select from only high-value animals in genetic 
        value analysis
    -   _all_: Randomly select from all animals in genetic value analysis
    -   _candidates_: Use candidate animals entered below to form groups
-   Sex ratio
    -   Randomly assign animals without regard to sex
    -   Use a harem structure with one breeding male per group
    -   Specify the sex ratio between 0.5 and 10 (F/M)
-   Whether or not to pre-populate (seed) groups with animals of your choice
-   Number of groups to be formed
-   Whether or not to ignore females at or above the minimum parent age
-   Number of simulations used to search for the optimal group makeup
-   Whether or not kinship coefficients are to be included in results

You decisions regarding each of the above options are expressed 
in a call to the function __groupAddAssign__. 
A complete description of the function and its arguments is available
using the code shown below.

```{r groupAddAssign-help-request, eval = FALSE}
?groupAddAssign
```

Below is the descriptions of the function parameters extracted from the 
documentation near the time this tutorial was prepared.

- __candidates__ Character vector of IDs of the animals available for use in 
forming the groups. The animals that may be present in currentGroups are not 
included within candidates.

- __currentGroups__ List of character vectors of IDs of animals currently 
assigned to groups. Defaults to a list with character(0) in each sub-list 
element (one for each group being formed) assuming no groups are 
pre-populated.

- __kmat__ Numeric matrix of pairwise kinship values. Rows and columns are 
named with animal IDs.

- __ped__ Dataframe that is the 'Pedigree'. It contains pedigree information 
including the IDs listed in candidates.

- __threshold__ Numeric value indicating the minimum kinship level to be 
considered in group formation. Pairwise kinship below this level will be 
ignored. The default values is 0.015625.

- __ignore__ List of character vectors representing the sex combinations to be
ignored. If provided, the vectors in the list specify if pairwise kinship 
should be ignored between certain sexes. Default is to ignore all pairwise 
kinship between females.

- __minAge__ Integer value indicating the minimum age to consider in group 
formation. Pairwise kinships involving an animal of this age or younger will
be ignored. Default is 1 year.

- __iter__ Integer indicating the number of times to perform the random group 
formation process. Default value is 1000 iterations.

- __numGp__ Integer value indicating the number of groups that should be formed
from the list of IDs. Default is 1.

- __updateProgress__ Function or NULL. If this function is defined, it will be
called during each iteration to update a shiny::Progress object.

- __harem__ Logical variable when set to TRUE, the formed groups have a single
male at least minAge old.

- __sexRatio__ Numeric value indicating the ratio of females to males x from 
0.5 to 20 by increments of 0.5.

- __withKin__ Logical variable when set to TRUE, the kinship matrix for the 
group is returned along with the group and score. Defaults to not return the
kinship matrix. This maintains compatibility with earlier versions.


We will use the _trimmedPed_ pedigree in our code.

For illustration purposes we are going to keep the numbers of candidates, 
groups, and iterations fairly small.

We will get first some animal IDs to use for our candidates by selecting animals
at least 2 years old at the time this pedigree was sampled (01-01-2015).

```{r get-candidates}
candidates <- trimmedPed$id[trimmedPed$birth < as.Date("2013-01-01") &
  !is.na(trimmedPed$birth) &
  is.na(trimmedPed$exit)]
table(trimmedPed$sex[trimmedPed$id %in% candidates])
```

Our candidates are made up of 
`r table(trimmedPed$sex[trimmedPed$id %in% candidates])[[1]]`
females and `r table(trimmedPed$sex[trimmedPed$id %in% candidates])[[2]]` males.
The parameters __currentGroups__, __threshold__, __ignore__, __minAge__, 
__sexRatio__, __withKin__, 
and __updateProgress__ are allowed to take their default values.
The setting of the __sexRatio__ parameter to 0 is ignored in the following 
call of the __groupAddAssign__ function.
This is consistent 
with the a value of 0 making little since in a breeding colony.

The empty seventh group at the bottom is evidence that all of the candidate
animals could be placed in a group without exceeding the default value of
0.015625.

### Harems

The following group assignments will be forming harem groups.
This is done by setting __harem__ to \code{TRUE}. Setting `iter` to 100 or more
will increase optimal composition of breeding groups

```{r create-harem-groups}
haremGrp <- groupAddAssign(
  kmat = trimmedGeneticValue[["kinship"]],
  ped = trimmedPed,
  candidates = candidates,
  iter = 10L,
  numGp = 6L,
  harem = TRUE
)
haremGrp$group
```

We can identify and list the males in each group with the following code.

```{r list-males-in-harem-groups}
sapply(haremGrp$group, function(ids) {
  ids[ids %in% trimmedPed$id[trimmedPed$sex == "M"]]
})
```

It is easy to notice that the male is listed first within each breeding group.

We can also see the number of animals and the sex ratios created in each group.
Since these are harem groups the sex ratios are determined by the number of 
animals in the individual groups.

```{r harem-count-and-sexratios}
lines <- sapply(haremGrp$group, function(ids) {
  paste0(
    "Count: ", length(ids), " Sex Ratio: ",
    round(calculateSexRatio(ids, trimmedPed), 2)
  )
})
for (line in lines) print(line)
```

Examination of this table shows that of the 
`r table(trimmedPed$sex[trimmedPed$id %in% candidates])[[1]]` females 
`r (sum(sapply(haremGrp$group[-1], function(ids) length(ids))) - length(haremGrp$group[-1]))`
are included.

### Controlling Sex Ratios

The following group assignments will be forming harem groups.
This is done by setting __harem__ to \code{TRUE}.

```{r create-sexratio-groups}
sexRatioGrp <- groupAddAssign(
  kmat = trimmedGeneticValue[["kinship"]],
  ped = trimmedPed,
  candidates = candidates,
  iter = 10L,
  numGp = 6L,
  sexRatio = 9.0
)
sexRatioGrp$group
```

Again we can identify and list the males in each group with the following code.

```{r list-males-in-sexratio-groups}
sapply(sexRatioGrp$group, function(ids) {
  ids[ids %in% trimmedPed$id[trimmedPed$sex == "M"]]
})
```

We can also see the number of animals and the sex ratios created in each group.

```{r list-count-and-sexratios}
lines <- sapply(sexRatioGrp$group, function(ids) {
  paste0(
    "Count: ", length(ids), " Sex Ratio: ",
    round(calculateSexRatio(ids, trimmedPed), 2L)
  )
})
for (line in lines) print(line)
```

Examination of this table shows that of the
`r table(trimmedPed$sex[trimmedPed$id %in% candidates])[[1]]` females
`r (sum(sapply(sexRatioGrp$group[-1], function(ids) length(ids))) - length(sexRatioGrp$group[-1]))`
are included.

## Individual Mate-Pair Analysis

__groupAddAssign__, used throughout the previous section, forms whole
breeding groups. __reportMatePairs__ instead answers a narrower question --
for a candidate population, which individual sire/dam pairs are even
eligible to be considered (opposite sex, both above __minAge__), and what
does every available kinship/genetic-value signal say about each pair? It
composes the package's existing pair-eligibility machinery into a single
report, returning two tables: _pairs_, one row per eligible sire/dam
combination, and _excluded_, one row per pair dropped and why. It computes
no composite ranking score of its own -- sort or filter the returned
columns as your own workflow requires. Marker-based kinship
(__markerKinship__'s output) and each parent's genetic-value context
(_indivMeanKin_/_gu_, from __reportGV__'s report) can both be supplied as
optional enrichment and default to `NA` when not.

We reuse the _candidates_ vector already built above, restricted further
to a small illustrative slice -- one male and seven females -- so the
output stays readable; a full call against all `r length(candidates)`
candidates would return one row per opposite-sex, age-eligible pair, which
quickly runs into the thousands.

```{r mate-pairs-setup}
smallCandidates <- head(candidates, 8L)
trimmedPed[trimmedPed$id %in% smallCandidates, c("id", "sex", "age")]
```

```{r mate-pairs-demo}
matePairs <- reportMatePairs(trimmedPed, trimmedGeneticValue[["kinship"]],
                              geneticValues = trimmedGeneticValue,
                              populationIds = smallCandidates)
matePairs$pairs
```

With only one male in this slice, every eligible pair shares the same
sire, so _sireIndivMeanKin_/_sireGu_ are identical across rows (both come
from _trimmedGeneticValue_'s report for that one animal); _damIndivMeanKin_/
_damGu_ vary by dam. _markerKinship_ is `NA` throughout, since no marker
genotype matrix was supplied to this call.

__exclude__ removes a specific id from consideration and records why in
the _excluded_ table -- useful when a curator already knows a particular
animal is unavailable this cycle, independent of age or sex eligibility.

```{r mate-pairs-exclude-demo}
matePairsExcluded <- reportMatePairs(trimmedPed, trimmedGeneticValue[["kinship"]],
                                      geneticValues = trimmedGeneticValue,
                                      populationIds = smallCandidates,
                                      exclude = matePairs$pairs$damId[1L])
matePairsExcluded$excluded
```

One caveat worth knowing before relying on __minAge__ alone to bound a
real-world call: a missing (`NA`) recorded age *passes* the age screen
rather than failing it, on the premise that an unknown age should not
silently disqualify an otherwise-eligible animal. __populationIds__, not
__minAge__, is the reliable way to bound which candidates a call
considers.

## Pedigree Errors

As stated earlier you can see which types of errors are detected by 
_qcStudbook_ by looking at names returned by _getEmptyErrorLst()_ as shown below.

```{r getEmptyErrorLst}
names(getEmptyErrorLst())
```

Each is defined below.

```{r make-errorList-definition-tbl, echo = FALSE, eval=TRUE}
errorTypes <- names(getEmptyErrorLst())
errorDescriptions <- c(
  "Database connection failed: configuration or permissions are invalid",
  "Columns that must be within the pedigree file are missing.",
  "Values, which are supposed to be dates, cannot be interpreted as a date.",
  "Parents were too young on the date of birth of to have been the parent.",
  "Individuals listed as female or hermaphroditic and as a sire.",
  "Individuals are listed as male and as a dam.",
  "Individuals who are listed as both a sire and a dam.",
  "IDs listed more than once.",
  "IDs (id, sire, or dam) that contain a period ('.'), which is not allowed.",
  stri_c(
    "Columns that have been changed to conform to internal naming ",
    "conventions and what they were changed to."
  )
)
errorTbl <- data.frame(
  Error = errorTypes, Definition = errorDescriptions,
  stringsAsFactors = FALSE
)
```
```{r print-error-definition-tbl, echo=FALSE}
knitr::kable(errorTbl)
```

We are going to 
use the small imaginary pedigree listed below that has multiple errors
to discuss pedigree error detection and reporting. 
First note the birth dates of ego_id _o4_ 
(2006-04-13) and the purported sire _s2_ (2006-06-19). Note also the purported 
birth date of the _d2_ and the birth dates of her offspring.
Obviously dates or IDs may be incorrect.

This is the pedigree. _(We will discuss the column names shortly.)_

```{r list-pedOne}
knitr::kable(nprcgenekeepr::pedOne)
```

If we try to convert this pedigree file into the standardized studbook format, 
we are going to get an error message and the 
creation of a file in the R sessions temporary directory that lists the records 
that have generated the errors. 

```{r summary-pedOne-no-errors, error = TRUE}
pedOne <- nprcgenekeepr::pedOne # put it in the local environment
ped <- qcStudbook(pedOne, minSireAge = 0.0, minDamAge = 0.0)
```

The contents of _lowParentAge.csv_ is shown below.

```{r read-lowParentAge.csv, echo = FALSE, include = FALSE}
lowParentAge <- read.csv(paste0(tempdir(), "/lowParentAge.csv"))
```
```{r print-lowParentAge.csv, echo = FALSE}
knitr::kable(lowParentAge)
```

Examination of the ages of the parents reveals the issues being reported.

We can remove the errors by changing the birth dates of _o4_ from 2006-04-13
to 2015-09-16 and of _d2_ from 2015-09-16 to 2006-04-13. 

```{r correct-birthdate, error = TRUE}
pedOne$birth_date[pedOne$ego_id == "o4"] <- as.Date("2015-09-16")
pedOne$birth_date[pedOne$ego_id == "d2"] <- as.Date("2006-04-13")
```

Note the changes made to the column names between the original __pedOne__ 
pedigree and the pedigree (__ped__) we get from __qcStudbook__. 
We have chosen to limit the displayed pedigree by selecting the _ego_id_'s and 
_id_'s in __pedOne__ and __ped__ respectively.

```{r display-corrected-birth-records}
ped <- qcStudbook(pedOne, minSireAge = 0.0, minDamAge = 0.0)
ped[ped$id %in% c("s2", "d2", "o3", "o4"), ]
```

However, the preferred method of creating the standardized studbook format
with __qcStudbook__ is to examine all errors found and correcting them 
before proceeding.
This is done by explicitly requesting that all aspects inconsistent with the 
studbook format be identified by setting _reportChanges_ and _reportErrors_ to
\code{TRUE}.

```{r look-for-errors}
errorList <- qcStudbook(pedOne,
  minSireAge = 0.0, minDamAge = 0.0, reportChanges = TRUE,
  reportErrors = TRUE
)
summary(errorList)
```

We will discuss each of these newly identified errors in a moment, however,
let's look at shortening this report, because often you will not be 
interested in the more trivial changes to the column names
made by __qcStudbook__ and in those cases you simply choose not to report 
changes to the column names as is shown here by setting _reportChanges_ to 
\code{FALSE}. 
For this illustration, we are going to bring back the original copy of 
__pedOne__ to see how the suspicious parents are reported by the 
__summary__ function.

```{r look-for-errors-only}
pedOne <- nprcgenekeepr::pedOne
errorList <- qcStudbook(pedOne,
  minSireAge = 0L, minDamAge = 0L, reportChanges = FALSE,
  reportErrors = TRUE
)
options(width = 90L)
summary(errorList)
```

The first two errors mentioned are of particular interest.
Currently __qcStudbook__ automatically changes the sex of dams to 
_F_ (female) and sires to _M_ (male) when __reportErrors__ is set
to \code{FALSE}.

## Genetic Loops 

This feature is not supported within the Shiny application and is not 
fully implemented. 

To use the __findLoops__ function run the following code and select a pedigree 
as your input file that has a loop in it. 
We are continuing to use the example pedigree that comes with the software
_Example_Pedigree.csv_.

```{r look-at-loops}
exampleTree <- createPedTree(breederPed)
exampleLoops <- findLoops(exampleTree)
```

You can count how many loops you have with the following code.

```{r countLoops}
length(exampleLoops)
nLoops <- countLoops(exampleLoops, exampleTree)
sum(unlist(nLoops[nLoops > 0L]))
```

You can list the first 10 sets of ids, sires and dams in loops with the
following line of code:

```{r listLoops}
examplePedigree[unlist(exampleLoops), c("id", "sire", "dam")][1L:10L, ]
```

## Marker Genetics

Everything up to this point has estimated kinship and diversity from a
_recorded_ pedigree -- __kinship__, __reportGV__, and the rest all assume the
sire/dam relationships in your pedigree data are correct. The functions in
this section instead estimate relatedness, heterozygosity, parentage
consistency, and between-center differentiation directly from a marker
genotype panel (e.g. a SNP or STR panel), independent of any known pedigree.
This is the same functionality behind the Shiny application's __Marker
Genetics__ tab -- see the "Marker Genetics" section of the
_colony-manager-guide_ article for the equivalent point-and-click workflow
and screenshots. The small example genotypes below match the ones used
there, so the numbers you see printed here are the same numbers shown in
that article's tables.

### Preparing a Marker Genotype File

A marker genotype file is long-format: one row per _id_ x _locus_, with
columns _id_, _locus_, _allele1_, and _allele2_. (This is a different shape
from the single-locus _first_name_/_second_name_ genotype format used
earlier by __addGenotype__/__geneDrop__ -- the two are unrelated.) You would
normally read this from a CSV file with __read.csv__, the same way the
pedigree file was read in at the start of this tutorial; here we build a
small example directly, for a parent (_P_), her offspring (_C_), and an
unrelated founder (_U_), genotyped at 10 biallelic loci.

```{r marker-genotype-setup}
markerGenotype <- data.frame(
  id = c(rep("P", 10L), rep("C", 10L), rep("U", 10L)),
  locus = rep(paste0("L", 1L:10L), 3L),
  allele1 = c(
    "A", "A", "A", "B", "A", "A", "B", "A", "A", "A",
    "A", "A", "B", "A", "A", "A", "B", "A", "A", "A",
    "A", "B", "A", "A", "A", "A", "A", "B", "A", "B"
  ),
  allele2 = c(
    "A", "B", "B", "B", "A", "B", "B", "B", "A", "B",
    "B", "A", "B", "B", "A", "B", "B", "B", "B", "A",
    "B", "B", "A", "B", "B", "A", "B", "B", "B", "B"
  ),
  stringsAsFactors = FALSE
)
```

__checkMarkerGenotypeFile__ validates the column shape and rejects any locus
with more than two distinct alleles (the estimators below all require
biallelic markers). __buildMarkerGenotypeMatrix__ then pivots the checked
long-format table into the wide _id_ x _locus_ matrix the rest of this
section's functions consume, one cell per individual/locus combination
(e.g. `"A/B"`), or `NA` where that individual has no genotype call at that
locus.

```{r marker-genotype-matrix}
markerGenotype <- checkMarkerGenotypeFile(markerGenotype)
genotypeMatrix <- buildMarkerGenotypeMatrix(markerGenotype)
genotypeMatrix
```

### Marker-Based Kinship

__markerKinship__ estimates pairwise kinship directly from
_genotypeMatrix_, using the "KING-robust" estimator (Manichaikul et al.
2010) -- the same estimator implemented by KING, PLINK2, and
`SNPRelate::snpgdsIBDKING`. It never looks at recorded parentage, so it is
useful as an independent cross-check on the pedigree-based kinship computed
by __kinship__/__reportGV__ above.

```{r marker-kinship-demo}
markerKmat <- markerKinship(genotypeMatrix)
markerKmat
```

_P_ and _C_ are the true parent/offspring pair in this example, and their
marker-based kinship (0.2) is clearly higher than either pair involving the
unrelated founder _U_. The KING-robust estimator is not bounded below by
zero -- _C_ and _U_'s negative estimate here is informative (more divergent
ancestry than the reference panel), not an error.

### Heterozygosity Diagnostic

__markerObservedHeterozygosity__ computes, per animal, the fraction of its
genotyped loci at which it is heterozygous (_Ho_).
__markerExpectedHeterozygosity__ computes Nei's (1973) gene diversity (_He_)
at each locus from the population's allele frequencies, plus the unweighted
mean across loci as a population-wide summary -- the standard
observed-vs-expected heterozygosity diagnostic. Both take the same
_genotypeMatrix_ shape as __markerKinship__. This example uses three
animals (_X_, _Y_, _Z_) genotyped at 4 loci, with _Y_ missing a call at
_L4_.

```{r marker-heterozygosity-setup}
hetGenotype <- data.frame(
  id = c(rep("X", 4L), rep("Y", 3L), rep("Z", 4L)),
  locus = c(paste0("L", 1L:4L), paste0("L", c(1L, 2L, 3L)), paste0("L", 1L:4L)),
  allele1 = c("A", "A", "A", "A", "A", "A", "B", "B", "A", "A", "A"),
  allele2 = c("A", "B", "B", "B", "B", "A", "B", "B", "A", "B", "A"),
  stringsAsFactors = FALSE
)
hetGenotype <- checkMarkerGenotypeFile(hetGenotype)
hetMatrix <- buildMarkerGenotypeMatrix(hetGenotype)
```

```{r marker-heterozygosity-demo}
markerObservedHeterozygosity(hetMatrix)
markerExpectedHeterozygosity(hetMatrix)
```

Note _Y_'s observed heterozygosity is computed over its own 3 genotyped
loci (1/3), not over all 4 -- a missing call at one locus does not lower an
animal's own _Ho_, since the denominator is always that animal's own
non-missing loci, not the full panel.

### Parentage Verification (Mendelian Exclusion)

__markerParentageExclusion__ cross-references a pedigree's recorded
dam/sire against each animal's marker genotype and flags a recorded parent
whose genotype evidence contradicts simple Mendelian inheritance --
directly useful for catching the kind of dam/sire misidentification a
paper pedigree alone cannot reveal. A locus counts as a conflict only when
the animal and the candidate parent are each homozygous for a *different*
allele ("opposite homozygotes"); by default (`maxExclusions = 2`) a
recorded parent is only flagged once 3 or more such conflicts accumulate,
since a single mismatching locus can arise from ordinary genotyping error
even for a true parent.

We reuse the P/C/U _genotypeMatrix_ from above, with a pedigree in which
_C_'s recorded dam (_P_) is correct, but _C_'s recorded sire has been
(deliberately, for this example) misrecorded as the unrelated _U_.

```{r marker-parentage-setup}
pedigree <- data.frame(id = c("P", "C", "U"), sire = c(NA, "U", NA),
                        dam = c(NA, "P", NA), stringsAsFactors = FALSE)
```

```{r marker-parentage-demo}
markerParentageExclusion(genotypeMatrix, pedigree)
```

_C_'s true dam (_P_) has zero exclusions and is not flagged; _C_'s falsely
recorded sire (_U_) has 3 exclusions, exceeding the default
`maxExclusions = 2`, and is correctly flagged.

### Candidate-Parent Likelihood Ranking

Once __markerParentageExclusion__ has flagged a recorded parent as
Mendelian-inconsistent with its offspring's genotype,
__markerParentageLikelihood__ picks up the investigation: it ranks
candidate replacement parents using a CERVUS-style multilocus
likelihood-ratio (LOD) score (Meagher & Thompson 1986; Marshall, Slate,
Kruuk & Pemberton 1998), the same approach independently validated in a
real captive macaque colony by de Groot et al. (2025). It is deliberately
report-only -- it never rewrites _pedigree$sire_/_pedigree$dam_; a flagged
parent is a curator decision, not something to silently overwrite.
Candidates default to __getPotentialParents__'s own demographically
-eligible list (breeding age, gestation window, proven-breeder preference)
for the flagged offspring/role, or can be supplied explicitly, as below.

```{r parentage-likelihood-setup}
likelihoodGenotype <- data.frame(
  id = c("O", "O", "D", "D", "C1", "C1", "C2", "C2"),
  locus = c("L1", "L2", "L1", "L2", "L1", "L2", "L1", "L2"),
  allele1 = c("A", "A", "A", "A", "A", "A", "B", "B"),
  allele2 = c("A", "B", "B", "A", "A", "B", "B", "B"),
  stringsAsFactors = FALSE
)
likelihoodMatrix <- buildMarkerGenotypeMatrix(likelihoodGenotype)
likelihoodPed <- data.frame(id = "O", sire = "C2", dam = "D",
                             stringsAsFactors = FALSE)
```

_O_ is the offspring, recorded dam _D_, recorded sire _C2_; _C1_ is an
unrecorded candidate. __markerParentageExclusion__ first shows why _C2_ is
worth a second look:

```{r parentage-exclusion-recheck}
markerParentageExclusion(likelihoodMatrix, likelihoodPed)
```

_C2_'s single exclusion (of 2 loci) falls within the default
`maxExclusions = 2` tolerance, so it is not `flagged` here -- but 1
conflicting locus is still worth ranking against alternatives:

```{r parentage-likelihood-demo}
markerParentageLikelihood(likelihoodMatrix, likelihoodPed, id = "O",
                           role = "sire", candidates = c("C1", "C2"),
                           minLoci = 1L)
```

_C1_ -- genetically identical to _O_ at both loci -- decisively outranks
the recorded sire _C2_ (`LOD` 0.47 vs `-Inf`). Notice _C2_'s `excluded`
column is `FALSE` (the same tolerant test as above) even though its `LOD`
is `-Inf`: a single opposite-homozygote locus (_C2_ is B/B at _L1_, where
_O_ is A/A) is a zero-tolerance Mendelian impossibility under the
likelihood model, with no allowance for ordinary genotyping error the way
the exclusion count is. `delta` for _C1_ is `Inf`, not a large finite
number -- the gap down to a `-Inf`-ranked neighbor is reported as
infinite, by design.

### Validating a Cross-Center Mapping

When an animal transfers between centers that use independent id
namespaces, the receiving center often has no way to know the animal's
real parents and records it as an artificial founder -- losing its actual
lineage even though the originating center's records still have it.
__resolveCrossCenterIds__ (next) fixes this by merging two centers'
pedigrees given a curator-confirmed id mapping, but it `stop()`s on the
*first* problem it finds in that mapping. __checkCrossCenterMapping__
shares the same validation logic but never stops early -- it returns every
problem found as a row in a data.frame, so a curator sees everything wrong
with a mapping in one call rather than a stop-fix-rerun cycle, one problem
at a time.

In this example, _T1_ (in Center A's pedigree, _pedA_) and _X9_ (in Center
B's pedigree, _pedB_) are the same physical animal -- Center B recorded it
as a founder (both parents `NA`) because it never knew _T1_'s real parents,
which Center A does have.

```{r cross-center-setup}
pedA <- data.frame(
  id = c("P1", "P2", "T1"), sire = c(NA, NA, "P1"), dam = c(NA, NA, "P2"),
  stringsAsFactors = FALSE
)
pedB <- data.frame(
  id = c("X9", "O1"), sire = c(NA, "X9"), dam = c(NA, NA),
  stringsAsFactors = FALSE
)
mapping <- data.frame(idA = "T1", idB = "X9", stringsAsFactors = FALSE)
```

```{r cross-center-mapping-demo}
checkCrossCenterMapping(pedA, pedB, mapping)
```

Zero rows means the mapping is clean -- __resolveCrossCenterIds__ can be
called on the same three inputs without error, as it is below. A mapping
with real problems (here, _idB_ `"X9"` mapped twice, and an _idA_ value
`"NOPE"` that doesn't exist in _pedA_) instead reports both problems at
once, rather than only the first:

```{r cross-center-mapping-bad-demo}
badMapping <- data.frame(idA = c("T1", "NOPE"), idB = c("X9", "X9"),
                          stringsAsFactors = FALSE)
checkCrossCenterMapping(pedA, pedB, badMapping)
```

One subtlety: a genuine undeclared id collision between the two centers
(an id present in both _pedA_ and _pedB_ but never named in _mapping_) is
only ever reported once every problem above is already clean -- a
collision check on a mapping that doesn't resolve to real pedigree rows
would be meaningless. A clean result from this function is what licenses
trusting the absence of a collision, not merely the absence of a
`"collision"` row in a dirty one.

### Cross-Center Identity Linking

With the mapping validated above, __resolveCrossCenterIds__ merges the two
centers' pedigrees into one, preferring whichever side actually recorded a
parent.

```{r cross-center-identity-demo}
resolveCrossCenterIds(pedA, pedB, mapping)
```

The merged pedigree has a single row for _T1_ with its real parents (_P1_,
_P2_) intact, and _O1_ -- recorded at Center B as _X9_'s offspring -- now
correctly points to _T1_ as its sire.

### Cross-Center Differentiation (Fst)

__markerFst__ takes a different, population-level view: rather than
linking individual animals across centers, it estimates Hudson's Fst
(Hudson, Slatkin & Maddison 1992, as given in closed form by Bhatia et al.
2013) -- how differentiated two centers' colonies are in allele frequency,
at each locus genotyped by both, plus a single pooled summary across loci.
The pooled value is a ratio of summed numerators and denominators, not a
mean of the per-locus ratios, which Bhatia et al. show is materially
biased.

```{r marker-fst-setup}
centerAGenotype <- data.frame(
  id = c(rep("CA1", 2L), rep("CA2", 2L), rep("CA3", 2L), rep("CA4", 2L)),
  locus = rep(c("L1", "L2"), 4L),
  allele1 = c("A", "A", "A", "A", "A", "A", "B", "A"),
  allele2 = c("A", "A", "A", "B", "B", "B", "B", "A"),
  stringsAsFactors = FALSE
)
centerBGenotype <- data.frame(
  id = c(rep("CB1", 2L), rep("CB2", 2L), rep("CB3", 2L),
         rep("CB4", 2L), rep("CB5", 2L), rep("CB6", 2L)),
  locus = rep(c("L1", "L2"), 6L),
  allele1 = c("A", "B", "B", "A", "A", "B", "B", "B", "A", "B", "B", "A"),
  allele2 = c("B", "B", "B", "B", "B", "B", "B", "B", "A", "B", "B", "B"),
  stringsAsFactors = FALSE
)
matrixA <- buildMarkerGenotypeMatrix(checkMarkerGenotypeFile(centerAGenotype))
matrixB <- buildMarkerGenotypeMatrix(checkMarkerGenotypeFile(centerBGenotype))
```

```{r marker-fst-demo}
markerFst(matrixA, matrixB)
```

_L2_ shows substantially more differentiation between the two centers than
_L1_ does in this example; the pooled value summarizes both loci into one
number for a quick between-center comparison.

### Multiallelic Marker Panels and Locus Metadata

Everything above assumed a biallelic panel -- __checkMarkerGenotypeFile__
rejects any locus with more than two distinct alleles, since the
KING-robust kinship estimator requires it. Real microsatellite/STR panels
are routinely multiallelic, though, so the package provides a second, more
permissive validator for the linkage-aware metrics below:
__checkLinkageMarkerGenotypeFile__ keeps the same structural checks
(column shape, no duplicate _id_ x _locus_ rows) but drops the biallelic
restriction.

```{r linkage-genotype-setup}
strGenotype <- data.frame(
  id = c("W", "X", "Y", "Z"),
  locus = c("L1", "L1", "L1", "L1"),
  allele1 = c("A", "A", "A", "A"),
  allele2 = c("B", "C", "A", "D"),
  stringsAsFactors = FALSE
)
```

This locus has 4 distinct alleles (A, B, C, D) across the four animals --
__checkLinkageMarkerGenotypeFile__ accepts it:

```{r linkage-genotype-demo}
checkLinkageMarkerGenotypeFile(strGenotype)
```

The same data fed to __checkMarkerGenotypeFile__ instead fails, exactly as
that function is designed to:

```{r biallelic-genotype-fails-demo, error = TRUE}
checkMarkerGenotypeFile(strGenotype)
```

The metrics below also need per-locus genomic position information, which
real curated panels rarely have complete records for.
__checkLocusMetadata__ validates a _locus_/_chrom_/_pos_ (optionally _cM_)
sidecar table and classifies each locus's own coverage as `"full"`
(chromosome and position both known), `"partial"` (one of the two), or
`"none"` (neither) -- downstream functions then decide for themselves
which loci they can use, rather than requiring complete metadata up front.

```{r locus-metadata-demo}
locusMetadata <- data.frame(
  locus = c("L1", "L2", "L3"),
  chrom = c("1", "1", NA),
  pos   = c(1000000, NA, NA),
  stringsAsFactors = FALSE
)
checkLocusMetadata(locusMetadata)
```

_L1_ has both chromosome and position and is `"full"`; _L2_ has only a
chromosome (`"partial"`); _L3_ has neither (`"none"`). __computeGenomicROH__
(the F_ROH tab's estimator) uses only `"full"`-coverage loci and drops the
rest with a named warning; __markerLdBlock__ below needs only _chrom_ and
tolerates a missing _pos_/_cM_ entirely.

### Realized Relatedness Variance

Pedigree-expected relatedness (kinship doubled) is only an average --
actual IBD sharing between two relatives varies around that expectation
because of Mendelian sampling and linkage. __markerRealizedRelatednessVariance__
computes a closed-form estimate of that variance (Hill & Weir 2011) for
Parent-Offspring, Full-Sibling, and Half-Sibling pairs -- every other
relationship category (grandparent, cousin, unrelated, self, and so on)
legitimately returns `NA`, not an error, since a pedigree-wide call
includes many such pairs as a matter of course. Despite the "marker" name
shared with the rest of this section, it needs no genotype data at all --
only a kinship matrix, the pedigree, and the species' chromosome count and
total autosomal genetic map length (in Morgans).

```{r realized-relatedness-variance-demo}
smallPedKmat <- kinship(smallPed$id, smallPed$sire, smallPed$dam, smallPed$gen,
                         sparse = FALSE)
## Rhesus macaque autosome count/map length are used here only as an
## example scale -- supply values appropriate to your own species.
rrv <- markerRealizedRelatednessVariance(smallPedKmat, smallPed, nChr = 20L,
                                          mapLength = 28)
rrv[rrv$relation %in% c("Parent-Offspring", "Full-Siblings", "Half-Siblings"), ]
```

Parent-offspring pairs have `varR` of exactly zero, not merely small --
biologically, exactly one allele at every locus is always IBD between a
parent and its offspring, so there is no variance to estimate around the
`R = 0.5` expectation. Full-sibling variance is roughly double half
-sibling variance in this example (two independently segregating parents
vs. one), a useful sanity check when interpreting the estimate for your
own pedigree.

### Linkage-Disequilibrium Blocks

__markerLdBlock__ takes a different, more classical view of marker
relationships: pairwise linkage disequilibrium (LD) between loci sharing a
chromosome, using Hedrick's (1987) frequency-weighted D' and a
multiallelic r² generalization, from phase frequencies estimated by an EM
algorithm (generalizing Excoffier & Slatkin 1995) since this package's
genotype matrix is unphased. It is explicitly the secondary, exploratory
statistic in a pair with __markerRealizedRelatednessVariance__ above --
classical LD theory assumes random mating, which a pedigreed colony
violates, so every row of its output carries a fixed caveat saying so.

```{r ld-block-setup}
ldGenotype <- checkLinkageMarkerGenotypeFile(data.frame(
  id = rep(c("W", "X", "Y", "Z"), 2),
  locus = rep(c("L1", "L2"), each = 4),
  allele1 = c("A", "A", "A", "A", "M", "M", "N", "N"),
  allele2 = c("A", "B", "A", "B", "M", "N", "M", "N"),
  stringsAsFactors = FALSE
))
ldGenotypeMatrix <- buildMarkerGenotypeMatrix(ldGenotype)
ldLocusMetadata <- checkLocusMetadata(data.frame(
  locus = c("L1", "L2"), chrom = c("1", "1"), pos = c(NA, NA),
  stringsAsFactors = FALSE
))
```

```{r ld-block-demo}
markerLdBlock(ldGenotypeMatrix, ldLocusMetadata)
```

_D' = 1_ (the maximum possible) in this toy example -- only _X_ is a
double heterozygote at both loci (the only source of phase ambiguity
here), and the EM estimator resolves it to a fully-consistent phase
assignment. Note _r²_ (1/3) does not move in lockstep with _D'_ here --
the two statistics capture different aspects of the same association, and
needn't agree.

### De-identifying LD-Block Results

__markerLdBlock__'s own output is a locus-pair-level population statistic
with no per-individual ids in it -- except _idsUsed_, populated only when
_founderIds_ restricts the computation to a named subset. That is the one
place a real id can leak, and __obfuscateLdBlocks__ closes it: it remaps
_idsUsed_ through the same alias vector __obfuscatePed__'s _map_ output
produces, mirroring __obfuscateTwinRelations__'s existing pattern from
elsewhere in the package's de-identification family. A row whose
_idsUsed_ contains an id absent from the map `stop()`s loudly rather than
silently leaking it.

```{r ld-block-obfuscate-setup}
ldBlockResult <- data.frame(
  locus1 = "L1", locus2 = "L2", chrom = "1", Dprime = 0.5, r2 = 0.3,
  nUsed = 2L, idsUsed = "F1,F2", caveat = "Descriptive statistic only.",
  stringsAsFactors = FALSE
)
obfuscatedTwinPed <- obfuscatePed(twinPed[, c("id", "sire", "dam", "sex", "gen")],
                                   map = TRUE)
```

```{r ld-block-obfuscate-demo}
obfuscateLdBlocks(ldBlockResult, obfuscatedTwinPed$map)
```

The real ids _F1_/_F2_ (reusing the twin-pedigree ids from the __Pedigree
Diagram__ section above) no longer appear in _idsUsed_ -- every other
column (the statistics themselves) is untouched. The exact alias strings
are randomly generated on every call and are not reproduced here; what
matters is that no real id from the input survives into the de-identified
table.

```{r elapsed-time}
elapsed_time <- get_elapsed_time_str(start_time)
```
The current date and time is `r Sys.time()`. The processing time for
this document was `r elapsed_time`.

```{r session-info}
sessionInfo()
```
