---
title: "Forming Breeding Groups"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview {#sec-overview}

`nprcgenekeepr` can assemble candidate animals into one or more **breeding
groups** that keep close relatives apart while spreading genetic diversity
across the colony, following the approach of Vinson and Raboin (2015). Groups
can be formed three ways:

- **unconstrained** -- members are drawn subject only to a kinship ceiling;
- **harem** -- each group is seeded with a single male and filled with other
  animals;
- **target sex ratio** -- each group is filled toward a chosen
  females-per-male ratio.

This article walks the whole pipeline -- from a raw studbook to formed groups --
entirely from R, using the `examplePedigree` data set that ships with the
package. The Shiny app (`runGeneKeepR()`) drives the same functions through a
point-and-click interface; here we script them directly.

## Setup {#sec-setup}

```{r}
#| label: load
#| message: false
library(nprcgenekeepr)
set_seed(1L)
```

Group formation makes random draws -- which group to fill next, which animal to
add, which males to use as harem sires -- so two runs differ unless the seed is
fixed. `groupAddAssign()` does **not** set a seed internally, so we call
`set_seed()` once, up front, to make everything below reproducible.
(`set_seed()` is a thin, version-stable wrapper around `set.seed()`.)

## From studbook to candidate list {#sec-pipeline}

Breeding-group formation needs two ingredients: a **list of candidate animals**
and a **kinship matrix** giving the relatedness of every candidate to every
other. Both are derived from a quality-controlled pedigree.

```{r}
#| label: qc
breederPed <- qcStudbook(examplePedigree,
  minParentAge  = 2.0,
  reportChanges = FALSE,
  reportErrors  = FALSE
)
```

`qcStudbook()` validates and standardizes the studbook (checking parentage,
sexes, dates, and minimum parent age). Next we mark the focal population -- here,
the non-founders still in the colony (those with at least one known parent) --
and trim the pedigree to those animals plus the ancestors needed to compute
their kinships.

```{r}
#| label: focal
focalAnimals <- breederPed$id[!(is.na(breederPed$sire) &
  is.na(breederPed$dam)) &
  is.na(breederPed$exit)]
breederPed <- setPopulation(ped = breederPed, ids = focalAnimals)
trimmedPed <- trimPedigree(focalAnimals, breederPed)

length(focalAnimals)
dim(trimmedPed)
```

`reportGV()` runs the genetic value analysis and -- the part we need here --
returns the **kinship matrix** for the trimmed pedigree.

```{r}
#| label: gv
trimmedGeneticValue <- reportGV(trimmedPed,
  guIter   = 50, # genome-uniqueness iterations; use >= 1000 in practice
  guThresh = 3,
  byID     = TRUE,
  updateProgress = NULL
)
kmat <- trimmedGeneticValue[["kinship"]]
dim(kmat)
```

> The small `guIter` above keeps this article quick to render. For real
> analyses use `guIter >= 1000`; the genome-uniqueness estimates stabilize only
> with many iterations.

Finally, choose the candidates -- here, animals born before 2013 that are still
in the colony.

```{r}
#| label: candidates
candidates <- trimmedPed$id[trimmedPed$birth < as.Date("2013-01-01") &
  !is.na(trimmedPed$birth) &
  is.na(trimmedPed$exit)]
length(candidates)
```

## Harem groups {#sec-harem}

With `harem = TRUE`, `groupAddAssign()` seeds each of the `numGp` groups with one
eligible male and then fills it with other animals, never placing two animals
whose kinship exceeds `threshold` (default `0.015625`, i.e. 1/64 -- the kinship
of second cousins) in the same group.

```{r}
#| label: harem
haremGrp <- groupAddAssign(
  kmat = kmat,
  ped  = trimmedPed,
  candidates = candidates,
  iter  = 10, # group-search iterations; use >= 1000 in practice
  numGp = 6,
  harem = TRUE
)
```

The result is a list with two elements, `group` and `score`. `group` holds one
character vector of animal IDs per group, followed by a final element collecting
any candidates left unplaced (a lone `NA` when every candidate was assigned):

```{r}
#| label: harem-sizes
names(haremGrp)
lengths(haremGrp$group) # size of each group; the last is the unplaced pool
haremGrp$group[[1]]     # the first harem group
```

## Groups with a target sex ratio {#sec-sexratio}

Passing a non-zero `sexRatio` (females per male) fills each group toward that
ratio instead of using a single harem sire. Here we aim for nine females per
male.

```{r}
#| label: sexratio
sexRatioGrp <- groupAddAssign(
  kmat = kmat,
  ped  = trimmedPed,
  candidates = candidates,
  iter     = 10,
  numGp    = 6,
  sexRatio = 9.0
)
lengths(sexRatioGrp$group)
```

`harem` and `sexRatio` are alternative strategies -- use one or the other.
Because the sex-ratio constraint caps how many animals of each sex a group can
hold, more candidates may be left in the trailing unplaced element than in the
harem case; raising `iter` lets the search place more of them.

## Key arguments {#sec-arguments}

| Argument | Default | Meaning |
|---|---|---|
| `candidates` | -- | character vector of animal IDs to place |
| `kmat` | -- | kinship matrix from `reportGV()[["kinship"]]` |
| `ped` | -- | the (trimmed) pedigree the candidates belong to |
| `numGp` | `1` | number of groups to form |
| `threshold` | `0.015625` | maximum within-group kinship (1/64 = second cousins) |
| `minAge` | `1.0` | minimum age (years) to be placed |
| `harem` | `FALSE` | seed each group with a single male |
| `sexRatio` | `0.0` | target females per male (0 = unconstrained) |
| `iter` | `1000` | group-search iterations (more = better groups) |

## See also {#sec-seealso}

- The **Studbook Quality Control** article -- `qcStudbook()` validates and
  standardizes a studbook, the first step before forming groups.
- The **Building a Focal-Animal Pedigree Offline** article -- build a focal-animal
  pedigree from files with no database, via `getFocalAnimalPedFromFile()`.
- The **Genetic Value Analysis** article -- `reportGV()` produces the kinship
  matrix `groupAddAssign()` consumes here, and ranks animals by mean kinship and
  genome uniqueness.
- The **Age-Sex Pyramid Plots** article -- picture the colony's age and sex
  structure with `getPyramidPlot()`.
- `rankSubjects()` -- the ranking scheme applied to a genetic value report.
- `kinship()` -- pairwise kinship coefficients from a pedigree.
- `runGeneKeepR()` -- the Shiny app that performs this workflow interactively.

**Reference.** Vinson A, Raboin MJ (2015). "A Practical Approach for Designing
Breeding Groups to Maximize Genetic Diversity in a Large Colony of Captive
Rhesus Macaques (*Macaca mulatta*)." *Journal of the American Association for
Laboratory Animal Science* 54(6):700-707.
