---
title: "Engineering nprcgenekeepr 2.0.0: Modular Architecture, Expanded Capability, and an AI-Assisted Development Process"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  echo = FALSE, warning = FALSE, message = FALSE
)
```

## Abstract {#sec-abstract}

`nprcgenekeepr` provides genetic tools for primate-colony management --
studbook quality control, pedigree construction, genetic value analysis, and
breeding-group formation -- as an R package with a Shiny front end and an
exposed API. Between its two most recent CRAN submissions (v1.0.8,
2025-07-25, through v2.0.0, 2026-07-09; 512 non-merge commits), the package
underwent a substantial modernization. A nine-phase, vertical-slice migration
retired a long-standing duplicate-implementation problem, replacing two
independently maintained Shiny applications with a single modular one (ten
`R/mod*.R` modules, 4,731 lines including `appUI.R`/`appServer.R`). Thirteen
curated new capabilities shipped alongside it, spanning species-aware parent
identification, more honest uncertainty reporting in Genetic Value Analysis,
and two new Shiny dashboards. The test suite grew from 132 to 257 files, and
its browser-driven end-to-end layer went from present-but-inert to
executable and behavior-verifying. Every session in the effort, including
this article's own drafting, was directed by Claude Code (Claude CLI) under
a written, enforced development protocol -- strict red-green-refactor TDD
for production code, single-deliverable sessions, and a durable, append-only
correction ledger -- rather than open-ended prompting. This article
documents that transformation quantitatively: every claim traces to a
commit, `CHANGELOG.md` entry, or frozen extraction, for readers evaluating
the package itself, its testing rigor, or its AI-assisted development
process.

## Introduction {#sec-introduction}

Between its two most recent CRAN submissions -- v1.0.8 (`4548aa1b`,
2025-07-25) and v2.0.0 (`8ca8bb24`, 2026-07-09) -- `nprcgenekeepr` changed
enough to warrant a written account of the work, not just a version-number
bump. This article documents that transformation for domain experts, other
primate-center bioinformatics groups, CRAN reviewers, and potential
contributors: **what** changed, and **how** it was built.

Four pillars structure the account. Section 1 describes retiring a
long-standing duplicate-implementation problem via a nine-phase,
vertical-slice migration to a modular Shiny architecture. Section 2
describes the new analysis capabilities that shipped alongside that
migration. Section 3 describes how the test suite grew and gained an
executable, behavior-verifying, browser-driven layer. Section 4 describes
the development process itself -- every session in this effort, including
the five sessions that drafted this article, was directed by **Claude
Code** (Claude CLI) operating under a written, enforced protocol rather
than open-ended prompting.

**Scope.** Every quantitative or dated claim in this article is bounded to
the 512 non-merge commits between those two submission commits
(`4548aa1b`..`8ca8bb24`) -- not "since the project began." As of this
writing (2026-07-09), CRAN's review of the v2.0.0 submission is still
pending; nothing in this article depends on that outcome.

## Section 1 -- From Monolith to Modules: the Shiny Architecture Transformation {#sec-modules}

For most of its life, `nprcgenekeepr` shipped **two coexisting Shiny
applications** that implemented the same features twice. The original,
`inst/application/` -- a single `server.r` (1,304 lines) plus a short
`ui.r` (53 lines) that sourced eight separate `uitp*.R` panel files --
ran alongside a newer, module-based rewrite (`R/appUI.R`, `R/appServer.R`,
and a growing set of `R/mod*.R` files). The two had already begun to
drift: different tab identifiers, different default parameters, a bug
fixed in one but not the other. Maintaining both was tracked as this
project's single biggest source of friction for adding new features
(issue #27).

"Completing the conversion" meant bringing the modular app to full
feature parity with the monolith, validating it end to end, and then
retiring the monolith outright -- not a gradual coexistence, a hard cutover.
That work ran as a **nine-phase, vertical-slice migration** -- each phase
one TDD session (RED -> GREEN -> REFACTOR with phase gates), each phase
leaving a working application behind it -- from Session 22 to Session 35
(2026-06-03 to 2026-06-06).

```{r}
#| label: fig-commit-pace
#| fig-cap: >
#|   Commits per month across the full v1.0.8 -> v2.0.0 range
#|   (`4548aa1b`..`8ca8bb24`, 512 non-merge commits, as of 2026-07-09). The
#|   June 2026 spike overlaps the nine-phase modularization migration
#|   (Session 22-35, 2026-06-03 to 2026-06-06), but June also carried other,
#|   unrelated work -- this chart is overall project pace, not a count of
#|   module-migration commits alone.
#| fig-width: 7
#| fig-height: 4
timeline <- read.csv("data/commit-activity-timeline.csv")
timeline$month <- factor(timeline$month, levels = timeline$month)
ggplot2::ggplot(timeline, ggplot2::aes(x = month, y = commit_count)) +
  ggplot2::geom_col(fill = "#3B6E8F") +
  ggplot2::geom_text(ggplot2::aes(label = commit_count), vjust = -0.4,
                      size = 3.2) +
  ggplot2::labs(x = NULL, y = "Commits") +
  ggplot2::theme_minimal(base_size = 11)
```

@fig-commit-pace situates the migration inside the wider effort: the
project moved from occasional, pre-methodology commits in late 2025 to a
sustained pace once the SESSION_RUNNER methodology took hold in
Session 1 (2026-05-30), peaking in June 2026 -- the same month the
module migration ran -- before settling into a testing-and-release-hardening
pace in July.

### Architecture, before and after

@fig-architecture contrasts the two designs directly. The monolith's
`server.r` was a single large reactive block reading and writing shared
state through ordinary R variables in one scope. The modular app instead
composes independent `modXServer()` functions, each returning a named
list of `reactive()` closures; `appServer()` wires those returns into one
`shared <- reactiveValues(...)` object and passes `reactive(shared$...)`
down to the modules that need it -- an explicit, if still informal,
module contract (documented as XARCH-2 in the migration plan).

::: {#fig-architecture}
```{mermaid}
flowchart TB
  subgraph Before["Before -- inst/application/, port 6012"]
    direction TB
    ui1["ui.r (53 lines)<br/>sources 8 uitp*.R panel files"] --> srv1
    srv1["server.r (1,304 lines)<br/>one monolithic reactive block"]
  end
  subgraph After["After -- R/appUI.R + R/appServer.R, port 6013"]
    direction TB
    ui2["appUI.R<br/>composes 10 modXXXUI() calls"]
    srv2["appServer.R<br/>shared reactiveValues()"]
    mods["10 R/mod*.R modules:<br/>modInput, modPedigree, modPyramid,<br/>modGeneticValue, modSummaryStats,<br/>modBreedingGroups, modGvAndBgDesc,<br/>modORIPReporting, modGeneticDiversity,<br/>modPotentialParents"]
    ui2 --> mods
    srv2 --> mods
    mods -- "reactive() returns" --> srv2
  end
```

The legacy monolith (`inst/application/`, deleted at Phase 9, commit
`24992e0b`) versus the modular app now launched by `runGeneKeepR()`
(default port 6013, as of 2026-07-09). `runModularApp()` -- the modular
app's original launcher -- is now a soft-deprecated alias that calls
`runGeneKeepR()`, so pre-existing callers keep working.
:::

The cutover was declared in Phase 9 (Session 35): `runGeneKeepR()` became
the canonical launcher via `lifecycle::deprecate_soft()`-aliasing
(`3db018d1`), and, in its own standalone commit immediately after,
`inst/application/` -- 17 files, including `server.r`, `ui.r`, and the
eight `uitp*.R` panels -- was deleted outright (`24992e0b`; docs/vignette
updates and an unrelated fix followed in `53a9e5e0`/`a1618c48`). Because
deletion was its own commit rather than folded into the alias change,
reverting the cutover -- had the modular app proven insufficient -- would
have been a single `git revert` of `24992e0b`, not an unpicking of nine
phases of work. As of 2026-07-09, that revert was never needed.

### The modules today

```{r}
#| label: tbl-modules
#| tbl-cap: >
#|   The ten Shiny modules mounted in the modular app, as of 2026-07-09
#|   (`4,731` total lines across `R/mod*.R`, `R/appUI.R`, and
#|   `R/appServer.R`).
modules <- read.csv("data/module-inventory.csv")
modules <- modules[order(-modules$loc), ]
kableExtra::kbl(
  modules,
  col.names = c("Module", "Responsibility", "Lines of code", "Test files"),
  align = c("l", "l", "r", "r")
) |>
  kableExtra::kable_styling(full_width = FALSE)
```

@tbl-modules lists all ten `R/mod*.R` files that exist in the package as
of 2026-07-09; together with `R/appUI.R` and `R/appServer.R` they total
4,731 lines. Six of the ten -- `modInput`, `modPedigree`, `modPyramid`,
`modGeneticValue`, `modSummaryStats`, and `modBreedingGroups` -- came out
of the nine-phase migration itself. `modGvAndBgDesc` was mounted early in
that migration (Phase 2, Session 23). `modORIPReporting` exists and is
wired into both `appUI()` and `appServer()`, but only mounts at runtime
for an ONPRC site configuration (`shouldShowOripTab()`), matching the
migration plan's explicit decision to keep it unwired-but-undeleted
rather than build out its functionality. `modGeneticDiversity` and
`modPotentialParents` are **not** migration artifacts -- they were added
afterward, as new capabilities in their own right; Section 2 (new
features) covers them.

### The nine-phase migration

```{r}
#| label: tbl-phases
#| tbl-cap: >
#|   The nine-phase, vertical-slice migration from monolith to modular
#|   app, Session 22 to Session 35 (2026-06-03 to 2026-06-06).
phases <- read.csv("data/migration-phases.csv", na.strings = "NA")
phases$highlight <- c(
  paste("Founder table added to the Summary Stats tab; kinship download",
        "moved to the module-internal getKinshipMatrix(); dual-name",
        "z-score lookup fix."),
  paste("Mounted the GvAndBgDesc description tab; resolved an HTML-id",
        "collision with the Genetic Value tab."),
  paste("Re-exposed the genome-uniqueness threshold selector",
        "(default 4); added a subset/filter export."),
  "Genotype-file merge restored for both separate- and common-pedigree input modes.",
  paste("Breeding Groups: per-group downloads, a per-group kinship view,",
        "and a group selector (new \"Group Detail\" tab)."),
  paste("Breeding Groups: seed-group pre-seeding; exposed three",
        "previously inert controls; corrected a 100x default-iteration",
        "drift (1000 -> 10)."),
  paste("Focal-animal / LabKey pedigree build restored (live-EHR path;",
        "verification was environmentally limited to a mocked seam)."),
  paste("shinytest2 E2E harness enabled end-to-end -- expanded into a",
        "4-session subplan (8a-8d, issue #39) plus a 7-part hardening",
        "pass (8e-1..8e-7, issue #40, Session 37-50)."),
  paste("Declared the modular app canonical: runGeneKeepR() now launches",
        "it directly; inst/application/ (17 files) deleted;",
        "runModularApp() became the soft-deprecated alias.")
)
phases$commit_display <- ifelse(
  is.na(phases$commit_sha), "not quoted in source plan", phases$commit_sha
)
kableExtra::kbl(
  phases[, c("phase", "highlight", "risk", "session", "commit_display",
             "status")],
  col.names = c("Phase", "What shipped", "Risk", "Session(s)",
                "Commit sha", "Status"),
  align = c("c", "l", "c", "c", "l", "l")
) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::column_spec(2, width = "26em")
```

@tbl-phases summarizes all nine phases. Phases 1, 2, and 9 are anchored
to directly-verified commit shas; phases 3 through 7 are verified only by
their `CHANGELOG.md` session-close-out entries -- the source migration
plan does not quote a sha for those phases, and this article does not
invent one. Phase 8 deserves a specific correction against an earlier,
looser characterization of this migration ("phases 1-9 all marked
DONE"): it was not a single-session parity phase like its neighbors. It
expanded into its own four-session subplan (8a-8d, Session 31-34, issue
#39, closed at Session 34) and, after that, a further seven-part
hardening pass (8e-1 through 8e-7, Session 37-50, issue #40) --
documented in the subplan's own body text and in `CHANGELOG.md`, not
visible from the migration plan's phase header alone. Phase 9, by
contrast, was the irreversible step: declaring the modular app canonical
and deleting the monolith outright, which is why it carries a HIGH risk
rating even though it shipped in a single session.

## Section 2 -- New Capabilities in 2.0.0 {#sec-features}

Between the v1.0.8 and v2.0.0 CRAN submissions, 47 GitHub issues closed
inside the 512-commit range (2025-07-26 to 2026-07-09). Most of those
closes were not new capability: fixes to existing behavior, issue-hygiene
closes verifying functionality that already existed before the range
(issue #34, for example, found `qcStudbook()` already integrated into
`modInput` from pre-migration work and simply closed the paper trail about
five months later, with no code change), and internal process items (lint
configuration, roxygen documentation harmonization, `codecov` config
consolidation) that belong to no reader-facing section of this article.
@tbl-features curates the 13 of 47 (28%) that shipped a capability an
analyst or maintainer can actually use -- a genuinely new statistic, a
new Shiny tab, or a materially different default -- each traced to its
closing issue number and the session(s) that implemented it.

```{r}
#| label: tbl-features
#| tbl-cap: >
#|   Curated new capabilities shipped in the v1.0.8 -> v2.0.0 range: 13
#|   of 47 closed GitHub issues judged genuinely feature-shaped, as of
#|   2026-07-09.
features <- read.csv("data/feature-highlights.csv", na.strings = "NA")
kableExtra::kbl(
  features[, c("feature", "issue", "session_range", "description")],
  col.names = c("Capability", "Issue", "Session(s)", "What it does"),
  align = c("l", "l", "c", "l")
) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::column_spec(1, width = "14em") |>
  kableExtra::column_spec(4, width = "24em")
```

@tbl-features groups into three loose clusters, covered below: parent
identification, Genetic Value Analysis uncertainty, and two new Shiny
dashboards.

### Parent identification grows principled and species-aware

The single largest cluster of new capability -- five of the thirteen
curated features -- reworked how the package identifies and reasons
about parents. It started narrowly: issue #31 replaced a fixed
dam-exclusion heuristic in `getPotentialParents()` with a window derived
from each species' actual gestation length (Session 74, commit
`0eeee3f6`), and the **Potential Parents** Shiny tab (issue #48, Session
80-81) turned that package-level logic into an app feature for the first
time -- previously, `getPotentialParents()` had no UI at all.

That work exposed a bigger gap: the package had no first-class notion of
species. Issue #46 (Session 165-169) added a `species` pedigree column
and made the gestation window and its UI defaults species-keyed instead
of assuming rhesus macaque throughout. Issue #73 (Session 182-189) went
further, adding a 14-species reproductive-parameter table (minimum
breeding ages, gestation length) with user-configurable overrides,
feeding both Genetic Value Analysis and the Potential Parents tab. Issue
#119 (Session 303-307, five slices) closed the loop: a single flat
`minParentAge = 2` default -- inconsistent with the species-aware
breeding-age table introduced for #73 -- was replaced end to end with
sex- and species-specific `minSireAge`/`minDamAge` floors across QC,
candidate-matching, and the Shiny UI.

### Genetic Value Analysis becomes more honest about uncertainty

Four features changed what Genetic Value Analysis reports, not by adding
a new tab but by making existing statistics more accurate or more
explicit about their own limits. Issue #9 (Session 176-181, three
slices) stopped animals with an unknown parent from silently inflating
their own mean kinship and falsely top-ranking as "unique" founders --
their parentage is now classified and the displayed rank corrected
accordingly. Issue #76 (Session 191-192), filed as a direct follow-on to
#9, carried the same logic to genome uniqueness: a both-unknown-parentage
"Undetermined" founder now reports `gu = 0` instead of a number with no
real biological signal behind it. Issue #82 (Session 208-209) added a
`fgSE` column next to `fg` in `reportGV()`'s output, so the founder
genome equivalents estimate ships with its own sampling uncertainty
rather than a bare point value. Issue #118 (Session 310-313, four
slices) added the largest single statistic of the range: effective
population size (Ne) estimates, computed three ways (gene diversity,
demographic sex-ratio, and variance effective size) -- a colony-level
number distinct from, and generally smaller than, simple census count.

### Two new dashboards, and a long-dormant module unlocked

The remaining features are new Shiny surfaces. Issue #112 (Session
280-283, four slices) added the **Genetic Diversity** tab: a
`modGeneticDiversity` module rendering a group x metric red/yellow/green
heatmap that combines four independent signals (proportion of low
genetic value, Indian-origin status, production status, and inbreeding
risk) into one at-a-glance view per breeding group. Issues #47 and #49
(Session 83-84) mounted `modORIPReporting` -- a module that had existed,
complete but never wired into the app, since the original Shiny-module
migration (@sec-modules) -- as an ONPRC-only grant-reporting tab, gated
by `shouldShowOripTab()` so it does not appear for other sites. Together
with the Potential Parents tab above, three of the ten modules listed in
@tbl-modules trace their existence, or their activation, to this
section's work rather than to the migration itself.

Two smaller fixes round out the curated set (@tbl-features): issue #35
(Session 68-69) extended the Pedigree Browser's focal-animal trim to
include descendants, not just ancestors; and issue #44 (Session 70-72)
replaced eight scattered literal `"U"`-prefix checks with a single
configurable auto-generated-ID format, defaulting to the prior behavior
unconfigured.

## Section 3 -- Testing at Scale {#sec-testing}

Growth in features (@sec-features) is only as trustworthy as the tests
behind it. Across the same v1.0.8 -> v2.0.0 range, the test suite grew
from 132 to 257 `.R` files under `tests/testthat/` -- test files plus the
shared helpers, fixtures, and setup script that support them -- a 95%
increase against a 512-commit range that itself roughly doubled the
package's feature surface. More consequential than the raw count,
though, is what changed in *kind*: a substantial browser-driven,
end-to-end test harness went from present-but-inert to executable, and
then from executable-but-shallow to exercising real data-bearing
workflows.

```{r}
#| label: tbl-testing-growth
#| tbl-cap: >
#|   Test-suite growth at five checkpoints across the v1.0.8 -> v2.0.0
#|   range, as of 2026-07-09. "Files in tests/testthat/" counts every
#|   `.R` file in the directory (test files plus helpers/fixtures/setup);
#|   "shinytest2/AppDriver-referencing" is the subset that names the
#|   browser-testing package or its driver class.
growth <- read.csv("data/testing-growth.csv")
growth$commit_display <- substr(growth$sha, 1, 8)
kableExtra::kbl(
  growth[, c("label", "date", "commit_display", "test_file_count",
             "shinytest2_referencing_files")],
  col.names = c("Checkpoint", "Date", "Commit", "Files in tests/testthat/",
                "shinytest2/AppDriver-referencing"),
  align = c("l", "c", "l", "r", "r")
) |>
  kableExtra::kable_styling(full_width = FALSE)
```

```{r}
#| label: fig-testing-growth
#| fig-cap: >
#|   Test-suite growth over the same five checkpoints as @tbl-testing-growth.
#|   The shinytest2/AppDriver-referencing count is near-zero relative to
#|   the total and is plotted against the same axis for comparison, not
#|   because it is negligible -- see prose for what that subset represents.
#| fig-width: 7
#| fig-height: 4
growth$label <- factor(growth$label, levels = growth$label)
growth_long <- data.frame(
  label = rep(growth$label, 2),
  count = c(growth$test_file_count, growth$shinytest2_referencing_files),
  series = rep(c("Files in tests/testthat/", "shinytest2/AppDriver-referencing"),
               each = nrow(growth))
)
ggplot2::ggplot(growth_long, ggplot2::aes(x = label, y = count, color = series,
                                           group = series)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::geom_text(ggplot2::aes(label = count), vjust = -1, size = 3.2,
                      show.legend = FALSE) +
  ggplot2::scale_color_manual(values = c("Files in tests/testthat/" = "#3B6E8F",
                                          "shinytest2/AppDriver-referencing" = "#C9622C")) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.05, 0.12))) +
  ggplot2::labs(x = NULL, y = "File count", color = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "bottom",
                 axis.text.x = ggplot2::element_text(angle = 20, hjust = 1))
```

@tbl-testing-growth and @fig-testing-growth ground the growth claim in
five checkpoints rather than just the two endpoints: the v1.0.8 CRAN
submission (`4548aa1b`, 2025-07-25, before this project adopted the
SESSION_RUNNER methodology), the start of Session 1 (`6fd87749`,
2026-05-30), the two commits bracketing Phase 9 of the module migration
(@sec-modules), and the v2.0.0 CRAN submission (`8ca8bb24`, 2026-07-09).
The shinytest2/AppDriver-referencing count is already 25 files at
Session 1's start -- that scaffold predates this project's methodology
entirely; it was built on the module branch (commit `7da01afe`) during
the same pre-Session-1 work that produced the modular app itself, but,
as the next section describes, it did not run.

### From a dormant scaffold to an executable, hardened harness

A shinytest2 browser end-to-end suite existed in the tree from early in
the module migration, but it did not work: the driver helper functions
its 23 test files called -- `create_app_driver()`,
`navigate_to_tab()`, `get_html_safe()`, and four others -- were never
defined. Every call would have thrown "could not find function" if the
suite had run; instead, an opt-in gate added at Session 19
(`create_test_app()`) made it skip cleanly by default, so the gap was
invisible rather than failing loudly. Issue #39 tracked closing it.

A Session 30 planning pass found the gap was larger than the parent
migration plan's Phase 8 had assumed -- 6 undefined helpers plus one
undefined constant, not the 3 the plan named -- and decomposed the work
into a four-session sub-plan
(`docs/planning/phase8-e2e-harness-subplan.md`, sub-phases 8a-8d,
Session 31-34). 8a defined the helpers browser-free; 8b proved Chrome
and the modular app itself would boot under `shinytest2`, the first-ever
browser run, and rewired CI to run the suite on a nightly schedule plus
manual dispatch rather than per pull request (browser tests are slower
and more flake-prone than the existing fast unit CI, which remained the
per-PR gate); 8c and 8d brought the remaining 20 files to
green-or-clean-skip. Issue #39 closed at the end of Session 34
(2026-06-06) on the sub-plan's own explicit completion bar: **executable
and CI-green opt-in**, not behavioral validation.

That bar was deliberate, not an oversight. The sub-plan documented,
in the same session that closed #39, why an executable suite does not
by itself mean the suite validates behavior: `navbarPage` renders every
tab's static markup into the DOM at boot, hidden by CSS, so the
suite's dominant pattern -- navigate to a tab, then `grepl()` a keyword
against the whole page body -- passes once the app merely boots,
regardless of which tab is actually selected. Coverage at the end of
8d was boot-level. A follow-on issue, #40, was filed at that same
session to close that specific gap.

Closing #40 took a further seven-slice pass (8e-1 through 8e-7,
Session 37-50, 2026-06-07 to 2026-06-10) that replaced boot-level
`expect_true(TRUE)` tautologies -- 41 of them, suite-wide -- with
assertions tied to the actually-selected pane, fixed a wrong-tab
navigation defect in the Summary Statistics family (7 of 8 tests
targeted "Genetic Value Analysis" instead of their own tab), and, in
its last three slices (8e-6a/b/c), rewired the pedigree-upload,
Genetic Value Analysis, and breeding-group E2E blocks to drive the
package's real analytical pipelines end to end rather than asserting
against static markup. A final slice, 8e-7 (Session 50), addressed an
operational flake rather than a behavioral gap: running all 23 files in
one CI process had produced roughly one transient Chrome/driver error
per five full-tier runs; partitioning the run into 13 per-module groups,
each a fresh R process, capped any single process at three files.
Issue #40 closed the following day, 2026-06-11.

As of 2026-07-09, both issues are closed and the harness the sub-plan
describes is the one currently checked in: `.github/workflows/shinytest2.yaml`
runs nightly (07:00 UTC) and on manual dispatch, opts every test in via
`NPRC_RUN_E2E`, and runs the per-module partition 8e-7 added -- distinct
from `R-CMD-check.yaml` and `test-coverage.yaml`, which remain the fast,
per-pull-request gate. That partition was exact against the 23
`test-{app,e2e}-*.R` files that existed when 8e-7 closed (2026-06-11), but
was not kept current as the suite grew: two more E2E files were added
afterward -- `test-e2e-potential-parents-module.R` (Session 82, issue #48)
and `test-e2e-orip-module.R` (Session 86, issues #47/#49), covering
exactly the two new Shiny tabs @sec-features describes -- without a
matching group added to the workflow. By 2026-07-08 the opt-in tier held
26 files against 13 groups covering only 24 of them, and those two
orphaned files never ran in CI -- a real gap, not a stale figure this
article repeated uncritically.

Session 337 (2026-07-09) closed that gap under the same TDD discipline
this section describes: a new regression test
(`tests/testthat/test_shinytest2_workflow_coverage.R`) parses the
workflow's group array and asserts every tracked file matches exactly one
group; two groups were added for the previously orphaned files; and the
workflow's header comment was rewritten to state the invariant without a
hardcoded file/group count, so it cannot go stale the same way again. A
live `workflow_dispatch` run confirmed the fix on real infrastructure --
all 15 groups green, including the two new ones
(run [29057393786](https://github.com/rmsharp/nprcgenekeepr/actions/runs/29057393786),
commit `c5ccf69b`).

## Section 4 -- An AI-Assisted Development Process {#sec-methodology}

Sections 1 through 3 describe *what* changed between v1.0.8 and v2.0.0. This
section describes *how*: every session in that work was directed by
**Claude Code** (Claude CLI) operating under a written protocol --
`SESSION_RUNNER.md`, `SAFEGUARDS.md`, and this package's own `CLAUDE.md` --
rather than open-ended prompting. Within the same 512-commit range as
Sections 1-3, **328 numbered sessions** (Session 1, `6fd87749`, 2026-05-30,
through Session 328, the exact v2.0.0-submission commit itself, `8ca8bb24`)
account for 502 of those commits; the other 10 predate the methodology --
informal module-branch experimentation plus the framework's own bootstrap
commit (`299060c0`, 2026-04-02).

```{r}
#| label: tbl-process-metrics
#| tbl-cap: >
#|   Engineering-process metrics, as extracted at the Phase A data freeze
#|   (Session 331, 2026-07-09). `CHANGELOG.md`, `PROJECT_LEARNINGS.md`,
#|   `HANDOFFS.md`, and `SESSION_NOTES.md` are live, append-only files that
#|   kept growing in every session after this extraction ran -- including
#|   the four sessions (331-334) that drafted the rest of this article --
#|   so these counts are a snapshot, not this article's own present state.
metrics <- read.csv("data/process-metrics.csv")
metrics$label <- c(
  "Sessions in range (Session 1 -> Session 328)",
  "Pre-methodology commits in range",
  "Session-1-to-range-end commits",
  "Total commits in range",
  "CHANGELOG.md ledger entries",
  "PROJECT_LEARNINGS.md entries",
  "Complete HANDOFFS.md receipts",
  "Self-assessments reporting 0 stakeholder corrections",
  "Self-assessments reporting >=1 stakeholder correction"
)
kableExtra::kbl(
  metrics[, c("label", "value")],
  col.names = c("Metric", "Count"),
  align = c("l", "r")
) |>
  kableExtra::kable_styling(full_width = FALSE)
```

### A written protocol, not open-ended prompting

Each session follows the same three-phase shape. **Orient** reads
`SAFEGUARDS.md` and `SESSION_NOTES.md`, checks open GitHub issues, runs a
project health dashboard (`methodology_dashboard.py`), and reconciles
`CHANGELOG.md`/`HANDOFFS.md` against `git log` for any commit a prior
session left unrecorded -- then reports what it found and stops for
direction; no work begins before the owner responds. **Execute** commits to
exactly **one deliverable** per session ("1 and done") and claims it in
`SESSION_NOTES.md`/`HANDOFFS.md` before any technical work starts, so a
session that crashes mid-task still leaves a trace. **Close out** evaluates
the previous session's handoff, self-assesses against a written score,
records new learnings, writes the next session's handoff, and appends a
dated entry to `CHANGELOG.md` before committing. `SESSION_RUNNER.md` names
the failure to do that last step **"unrecorded action"** -- one of 27 named,
numbered failure modes the protocol maintains, each added after a session
was observed failing that specific way.

### Strict TDD, gated by explicit confirmation

For production R code -- not this article's own drafting sessions, see
below -- `CLAUDE.md`'s Development Process Contract requires **strict
red-green-refactor TDD**: RED writes only failing tests, no implementation
code; GREEN writes the minimum implementation that passes them, no new
functionality; REFACTOR improves structure with no behavior change, all
tests still passing. @fig-tdd-cycle diagrams the cycle. What makes it more
than a style guideline is the enforcement mechanism: the assistant must
declare its current phase at the top of every response, and every phase
transition (`PRE-RED->RED`, `RED->GREEN`, `GREEN->REFACTOR`) is gated by a
structured `AskUserQuestion` prompt -- not a prose question -- with a header
of the exact form `TDD: <FROM>-><TO>`, an option spelling out the concrete
next-phase actions and verification, and an explicit hold/alternative
option. Skipping a phase, or writing implementation code during RED, is a
declared protocol violation the assistant must acknowledge and correct.

::: {#fig-tdd-cycle}
```{mermaid}
stateDiagram-v2
    [*] --> RED: AskUserQuestion gate (PRE-RED -> RED)
    RED --> GREEN: AskUserQuestion gate (RED -> GREEN)
    GREEN --> REFACTOR: AskUserQuestion gate (GREEN -> REFACTOR)
    REFACTOR --> [*]

    note right of RED
      Write failing tests only.
      No implementation code.
    end note
    note right of GREEN
      Minimum code to pass.
      No new functionality, no refactor.
    end note
    note right of REFACTOR
      Improve structure only.
      No behavior change; tests stay green.
    end note
```

The RED -> GREEN -> REFACTOR cycle `CLAUDE.md`'s Development Process
Contract requires for production R code, with an explicit `AskUserQuestion`
confirmation gate at every transition. Documentation and article-drafting
sessions -- including the three sessions that produced Sections 1-3 and the
session that produced this one -- are an explicitly declared exception (TDD
phase "N/A", stated at the top of every response) rather than a silent
omission.
:::

This article's own drafting sessions illustrate the exception rather than
the rule: none touched `R/` or `tests/`, so each declared **TDD phase: N/A**
explicitly in every response instead of running the cycle against prose it
does not apply to -- a small instance of the same discipline extending to
correctly scoping *when* a rule applies, not applying it uniformly
regardless of fit.

### Self-assessment and a durable correction ledger

@fig-self-score-trend plots every complete `HANDOFFS.md` self-assessment
score against session, for the sessions with one available.

```{r}
#| label: fig-self-score-trend
#| fig-cap: >
#|   Session self-assessed scores (0-10 scale, `HANDOFFS.md`'s `self_score`
#|   field), the 7 sessions with a complete receipt as of the Phase A data
#|   freeze. **Partial window, not a trend across the range:** receipts
#|   began at Session 324, one session before the `CHANGELOG.md`
#|   ledger-format resolution (Session 325) -- these 7 sessions are roughly
#|   2% of the 328-session range, and Sessions 329-330 were committed after
#|   the range's own end commit (`8ca8bb24`), on the same calendar day. Mean
#|   score across the 7: 8.57.
#| fig-width: 7
#| fig-height: 4
scores <- read.csv("data/self-score-trend.csv")
scores$session_num <- as.integer(sub("^S", "", scores$session))
scores <- scores[order(scores$session_num), ]
scores$session <- factor(scores$session, levels = scores$session)
ggplot2::ggplot(scores, ggplot2::aes(x = session, y = self_score, group = 1)) +
  ggplot2::geom_line(linewidth = 0.9, color = "#3B6E8F") +
  ggplot2::geom_point(size = 2.5, color = "#3B6E8F") +
  ggplot2::geom_text(ggplot2::aes(label = self_score), vjust = -1.2, size = 3.2) +
  ggplot2::scale_y_continuous(limits = c(0, 10.5), breaks = seq(0, 10, 2)) +
  ggplot2::labs(x = NULL, y = "Self-assessed score (0-10)") +
  ggplot2::theme_minimal(base_size = 11)
```

The mechanism behind that self-scoring is a durable receipt, not a spoken
summary: each session writes a `HANDOFFS.md` block recording what it did,
what is next, key files, gotchas, and its own score, and the *next* session
scores that handoff in turn before starting its own work -- visible in
@tbl-process-metrics as 7 complete receipts against 328 sessions in range
(the receipt requirement itself dates to Session 324). The same
append-only discipline extends to `CHANGELOG.md`, and this article's own
drafting sessions are a live, verifiable example of it working: each of the
four sessions that wrote Sections 1-3 closed out with a `commit: pending`
placeholder in its own `HANDOFFS.md` receipt (a session cannot know its own
close-out commit's sha before making that commit), corrected the
placeholder with one small follow-up commit once the sha existed, and --
because that follow-up commit landed after the session's own `CHANGELOG.md`
entry was already written -- the *following* session's Orient step caught
and logged it in turn: `cc0f7798` (Session 331's fix, logged by
Session 332), `2278b46f` (Session 332's fix, logged by Session 333),
`ee690776` (Session 333's fix, logged by Session 334), and `5f0b81d2`
(Session 334's fix, logged by this article's fifth session). Four
consecutive sessions produced the same small gap; four consecutive Orient
steps caught it.

`PROJECT_LEARNINGS.md` -- 305 dated entries as of the same Phase A freeze,
each naming a concrete finding, file, mechanism, and verdict, plus a
glossary of named recurring reflexes cited by tag -- is this project's
running record of what went wrong and what fixed it. `SESSION_RUNNER.md`
maintains a parallel, numbered catalog of 27 named failure modes at the
methodology level -- among them "Eager to start" (skipping orientation),
"Minimal handoff" (a technically-present but functionally useless handoff),
and "Unrecorded action" (the gap the receipt-sha backfills above
illustrate) -- each added after a session was observed failing that way.
Neither catalog is edited to remove an entry once a session outgrows it;
corrections accumulate rather than get erased.

@tbl-process-metrics's stakeholder-correction figures are the least precise
numbers in this section, and are presented with that caveat rather than
smoothed over: across `SESSION_NOTES.md`'s self-assessments, 269 explicit
mentions report zero stakeholder corrections and 2 report exactly one --
99.3% zero, by mention count. That is a text-mention proxy, not an
independently re-verified per-session audit; a session's own self-assessment
is not an external check on itself. Read it as "the sessions that stated a
number mostly stated zero," not as an audited defect rate.

## Conclusion {#sec-conclusion}

The v1.0.8 -> v2.0.0 effort retired a long-standing architectural liability
-- two independently maintained Shiny applications drifting out of sync --
shipped thirteen curated new analysis capabilities, grew the test suite from
132 to 257 files while turning its browser-driven layer from
present-but-inert into executable and behavior-verifying, and did all of it
under a written, gated development process rather than ad hoc iteration.

For maintainers and collaborators, the practical implications are concrete.
New features have exactly one Shiny implementation to extend, not two to
keep in sync (@sec-modules). The `R/mod*.R` module boundary is now the unit
of both development and testing (@sec-modules, @sec-testing). And the
session-by-session record in `CHANGELOG.md`, `PROJECT_LEARNINGS.md`, and
`HANDOFFS.md` is a navigable history of *why* each change was made, not
just *what* changed (@sec-methodology) -- a resource future sessions, and
future maintainers, can read directly rather than reconstruct from commit
messages alone.

This work has been supported in part by NIH grants P51 RR13986 to the
Southwest National Primate Research Center and P51 OD011092 to the Oregon
National Primate Research Center.
