---
title: "Engineering nprcgenekeepr 2.0.0: Modular Architecture, Expanded Capability, and an AI-Assisted Development Process"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  echo = FALSE, warning = FALSE, message = FALSE
)
```

## Section 1 -- From Monolith to Modules: the Shiny Architecture Transformation {#sec-modules}

For most of its life, `nprcgenekeepr` shipped **two coexisting Shiny
applications** that implemented the same features twice. The original,
`inst/application/` -- a single `server.r` (1,304 lines) plus a short
`ui.r` (53 lines) that sourced eight separate `uitp*.R` panel files --
ran alongside a newer, module-based rewrite (`R/appUI.R`, `R/appServer.R`,
and a growing set of `R/mod*.R` files). The two had already begun to
drift: different tab identifiers, different default parameters, a bug
fixed in one but not the other. Maintaining both was tracked as this
project's single biggest source of friction for adding new features
(issue #27).

"Completing the conversion" meant bringing the modular app to full
feature parity with the monolith, validating it end to end, and then
retiring the monolith outright -- not a gradual coexistence, a hard cutover.
That work ran as a **nine-phase, vertical-slice migration** -- each phase
one TDD session (RED -> GREEN -> REFACTOR with phase gates), each phase
leaving a working application behind it -- from Session 22 to Session 35
(2026-06-03 to 2026-06-06).

```{r}
#| label: fig-commit-pace
#| fig-cap: >
#|   Commits per month across the full v1.0.8 -> v2.0.0 range
#|   (`4548aa1b`..`8ca8bb24`, 512 non-merge commits, as of 2026-07-09). The
#|   June 2026 spike overlaps the nine-phase modularization migration
#|   (Session 22-35, 2026-06-03 to 2026-06-06), but June also carried other,
#|   unrelated work -- this chart is overall project pace, not a count of
#|   module-migration commits alone.
#| fig-width: 7
#| fig-height: 4
timeline <- read.csv("data/commit-activity-timeline.csv")
timeline$month <- factor(timeline$month, levels = timeline$month)
ggplot2::ggplot(timeline, ggplot2::aes(x = month, y = commit_count)) +
  ggplot2::geom_col(fill = "#3B6E8F") +
  ggplot2::geom_text(ggplot2::aes(label = commit_count), vjust = -0.4,
                      size = 3.2) +
  ggplot2::labs(x = NULL, y = "Commits") +
  ggplot2::theme_minimal(base_size = 11)
```

@fig-commit-pace situates the migration inside the wider effort: the
project moved from occasional, pre-methodology commits in late 2025 to a
sustained pace once the SESSION_RUNNER methodology took hold in
Session 1 (2026-05-30), peaking in June 2026 -- the same month the
module migration ran -- before settling into a testing-and-release-hardening
pace in July.

### Architecture, before and after

@fig-architecture contrasts the two designs directly. The monolith's
`server.r` was a single large reactive block reading and writing shared
state through ordinary R variables in one scope. The modular app instead
composes independent `modXServer()` functions, each returning a named
list of `reactive()` closures; `appServer()` wires those returns into one
`shared <- reactiveValues(...)` object and passes `reactive(shared$...)`
down to the modules that need it -- an explicit, if still informal,
module contract (documented as XARCH-2 in the migration plan).

::: {#fig-architecture}
```{mermaid}
flowchart TB
  subgraph Before["Before -- inst/application/, port 6012"]
    direction TB
    ui1["ui.r (53 lines)<br/>sources 8 uitp*.R panel files"] --> srv1
    srv1["server.r (1,304 lines)<br/>one monolithic reactive block"]
  end
  subgraph After["After -- R/appUI.R + R/appServer.R, port 6013"]
    direction TB
    ui2["appUI.R<br/>composes 10 modXXXUI() calls"]
    srv2["appServer.R<br/>shared reactiveValues()"]
    mods["10 R/mod*.R modules:<br/>modInput, modPedigree, modPyramid,<br/>modGeneticValue, modSummaryStats,<br/>modBreedingGroups, modGvAndBgDesc,<br/>modORIPReporting, modGeneticDiversity,<br/>modPotentialParents"]
    ui2 --> mods
    srv2 --> mods
    mods -- "reactive() returns" --> srv2
  end
```

The legacy monolith (`inst/application/`, deleted at Phase 9, commit
`3db018d1`) versus the modular app now launched by `runGeneKeepR()`
(default port 6013, as of 2026-07-09). `runModularApp()` -- the modular
app's original launcher -- is now a soft-deprecated alias that calls
`runGeneKeepR()`, so pre-existing callers keep working.
:::

The cutover was declared in Phase 9 (Session 35): `runGeneKeepR()` became
the canonical launcher via `lifecycle::deprecate_soft()`-aliasing, and
`inst/application/` -- 17 files, including `server.r`, `ui.r`, and the
eight `uitp*.R` panels -- was deleted in that same, standalone commit
(`3db018d1`, with the alias and orphan cleanup following in
`24992e0b`/`53a9e5e0`/`a1618c48`). Because deletion was its own commit
rather than folded into a parity change, reverting the cutover -- had the
modular app proven insufficient -- would have been a single `git revert`,
not an unpicking of nine phases of work. As of 2026-07-09, that revert
was never needed.

### The modules today

```{r}
#| label: tbl-modules
#| tbl-cap: >
#|   The ten Shiny modules mounted in the modular app, as of 2026-07-09
#|   (`4,731` total lines across `R/mod*.R`, `R/appUI.R`, and
#|   `R/appServer.R`).
modules <- read.csv("data/module-inventory.csv")
modules <- modules[order(-modules$loc), ]
kableExtra::kbl(
  modules,
  col.names = c("Module", "Responsibility", "Lines of code", "Test files"),
  align = c("l", "l", "r", "r")
) |>
  kableExtra::kable_styling(full_width = FALSE)
```

@tbl-modules lists all ten `R/mod*.R` files that exist in the package as
of 2026-07-09; together with `R/appUI.R` and `R/appServer.R` they total
4,731 lines. Six of the ten -- `modInput`, `modPedigree`, `modPyramid`,
`modGeneticValue`, `modSummaryStats`, and `modBreedingGroups` -- came out
of the nine-phase migration itself. `modGvAndBgDesc` was mounted early in
that migration (Phase 2, Session 23). `modORIPReporting` exists and is
wired into both `appUI()` and `appServer()`, but only mounts at runtime
for an ONPRC site configuration (`shouldShowOripTab()`), matching the
migration plan's explicit decision to keep it unwired-but-undeleted
rather than build out its functionality. `modGeneticDiversity` and
`modPotentialParents` are **not** migration artifacts -- they were added
afterward, as new capabilities in their own right; Section 2 (new
features) covers them.

### The nine-phase migration

```{r}
#| label: tbl-phases
#| tbl-cap: >
#|   The nine-phase, vertical-slice migration from monolith to modular
#|   app, Session 22 to Session 35 (2026-06-03 to 2026-06-06).
phases <- read.csv("data/migration-phases.csv", na.strings = "NA")
phases$highlight <- c(
  paste("Founder table added to the Summary Stats tab; kinship download",
        "moved to the module-internal getKinshipMatrix(); dual-name",
        "z-score lookup fix."),
  paste("Mounted the GvAndBgDesc description tab; resolved an HTML-id",
        "collision with the Genetic Value tab."),
  paste("Re-exposed the genome-uniqueness threshold selector",
        "(default 4); added a subset/filter export."),
  "Genotype-file merge restored for both separate- and common-pedigree input modes.",
  paste("Breeding Groups: per-group downloads, a per-group kinship view,",
        "and a group selector (new \"Group Detail\" tab)."),
  paste("Breeding Groups: seed-group pre-seeding; exposed three",
        "previously inert controls; corrected a 100x default-iteration",
        "drift (1000 -> 10)."),
  paste("Focal-animal / LabKey pedigree build restored (live-EHR path;",
        "verification was environmentally limited to a mocked seam)."),
  paste("shinytest2 E2E harness enabled end-to-end -- expanded into a",
        "4-session subplan (8a-8d, issue #39) plus a 7-part hardening",
        "pass (8e-1..8e-7, issue #40, Session 37-50)."),
  paste("Declared the modular app canonical: runGeneKeepR() now launches",
        "it directly; inst/application/ (17 files) deleted;",
        "runModularApp() became the soft-deprecated alias.")
)
phases$commit_display <- ifelse(
  is.na(phases$commit_sha), "not quoted in source plan", phases$commit_sha
)
kableExtra::kbl(
  phases[, c("phase", "highlight", "risk", "session", "commit_display",
             "status")],
  col.names = c("Phase", "What shipped", "Risk", "Session(s)",
                "Commit sha", "Status"),
  align = c("c", "l", "c", "c", "l", "l")
) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::column_spec(2, width = "26em")
```

@tbl-phases summarizes all nine phases. Phases 1, 2, and 9 are anchored
to directly-verified commit shas; phases 3 through 7 are verified only by
their `CHANGELOG.md` session-close-out entries -- the source migration
plan does not quote a sha for those phases, and this article does not
invent one. Phase 8 deserves a specific correction against an earlier,
looser characterization of this migration ("phases 1-9 all marked
DONE"): it was not a single-session parity phase like its neighbors. It
expanded into its own four-session subplan (8a-8d, Session 31-34, issue
#39, closed at Session 34) and, after that, a further seven-part
hardening pass (8e-1 through 8e-7, Session 37-50, issue #40) --
documented in the subplan's own body text and in `CHANGELOG.md`, not
visible from the migration plan's phase header alone. Phase 9, by
contrast, was the irreversible step: declaring the modular app canonical
and deleting the monolith outright, which is why it carries a HIGH risk
rating even though it shipped in a single session.
