---
title: "Genetic Value Analysis"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview {#sec-overview}

The **genetic value analysis** is the heart of `nprcgenekeepr`. It ranks the
animals in a colony by how much each one contributes to the colony's genetic
diversity, so that managers can prioritize genetically valuable animals for
breeding and retention (Vinson and Raboin 2015). The ranking rests on two
complementary metrics:

- **mean kinship** -- the average kinship coefficient between an animal and
  all animals in the population (itself included). A *low* mean kinship is good:
  it means
  the animal is only distantly related to the rest of the colony, so its genes
  are under-represented and worth propagating.
- **genome uniqueness** -- the percentage of an animal's alleles that are
  *rare* in the population (carried by no more than a few animals), estimated by
  a gene-drop simulation. A *high* genome uniqueness is good: it means the animal
  carries alleles that would be lost if it were not bred.

`reportGV()` computes both, ranks the population, and also returns colony-level
diversity summaries. This article runs the whole analysis from R, using the
`examplePedigree` data set that ships with the package. The Shiny app
(`runGeneKeepR()`) drives the same `reportGV()` through its Genetic Value
Analysis tab; here we script it directly.

## Setup {#sec-setup}

```{r}
#| label: load
#| message: false
library(nprcgenekeepr)
set_seed(1L)
```

Genome uniqueness is estimated by a gene-drop simulation -- founder alleles are
dropped down the pedigree at random and the rare ones counted -- so the genome
uniqueness values (and the founder-genome-equivalent below) differ between runs
unless the seed is fixed. `reportGV()` does **not** set a seed internally, so we
call `set_seed()` once, up front, to make everything reproducible.
(`set_seed()` is a thin, version-stable wrapper around `set.seed()`.) Mean
kinship is computed deterministically from the pedigree and does not depend on
the seed.

## From studbook to a population {#sec-pipeline}

The analysis needs a quality-controlled pedigree and a defined **population of
interest** -- the animals whose genetic value we want to rank.

```{r}
#| label: qc
breederPed <- qcStudbook(examplePedigree,
  minParentAge  = 2.0,
  reportChanges = FALSE,
  reportErrors  = FALSE
)
```

`qcStudbook()` validates and standardizes the studbook (checking parentage,
sexes, dates, and minimum parent age). Next we choose the population of interest.
Here it is the animals that have **at least one known parent** (so they are not
founders) and are **still in the colony** (no exit date). We mark them with
`setPopulation()` and trim the pedigree to those animals plus the ancestors
needed to compute their kinships.

```{r}
#| label: focal
focalAnimals <- breederPed$id[!(is.na(breederPed$sire) &
  is.na(breederPed$dam)) &
  is.na(breederPed$exit)]
breederPed <- setPopulation(ped = breederPed, ids = focalAnimals)
trimmedPed <- trimPedigree(focalAnimals, breederPed)

length(focalAnimals)  # animals in the population of interest
dim(trimmedPed)        # those animals plus the ancestors kinship needs
```

## Running the analysis {#sec-run}

```{r}
#| label: gv
gv <- reportGV(trimmedPed,
  guIter   = 50, # gene-drop iterations; use >= 1000 in practice
  guThresh = 1,  # an allele is "unique" if carried by <= this many animals
  byID     = TRUE
)
names(gv)
```

> The small `guIter` above keeps this article quick to render. For real
> analyses use `guIter >= 1000`; the genome-uniqueness estimates stabilize only
> with many iterations.

`reportGV()` returns a list. The two elements you will use most are `report`
(the ranked table) and `kinship` (the pairwise kinship matrix, which
`groupAddAssign()` consumes when forming breeding groups -- see the *Forming
Breeding Groups* article). The rest summarize founder diversity (below).

## The ranking report {#sec-report}

`report` is a data frame with one row per animal in the population of interest,
already sorted from most to least genetically valuable.

```{r}
#| label: report
rpt <- gv[["report"]]
dim(rpt)
names(rpt)
head(rpt[, c("id", "sex", "age", "indivMeanKin", "zScores", "gu",
  "value", "rank")], 8)
```

The key columns:

| Column | Meaning |
|---|---|
| `indivMeanKin` | the animal's mean kinship (lower = less related to the colony = better) |
| `zScores` | mean kinship standardized across the population (mean 0, SD 1) |
| `gu` | genome uniqueness, as a percentage (higher = carries rarer alleles = better) |
| `totalOffspring` / `livingOffspring` | offspring counts |
| `value` | a `High Value` / `Low Value` (occasionally `Undetermined`) flag |
| `rank` | integer rank, 1 = most valuable |

The `rank` and `value` columns come from the ranking scheme in `orderReport()`
and `rankSubjects()`, which sorts the population into ordered tiers:

1. imported founders with no offspring;
2. animals with genome uniqueness above 10%, by **descending genome
   uniqueness** (ties broken by ascending mean kinship) -- this is where
   carrying rare alleles earns a high rank;
3. the remaining animals whose standardized mean kinship is at or below 0.25
   (i.e. no more related to the colony than about average), by **ascending mean
   kinship** -- least related ranked highest;
4. everyone else, flagged `Low Value`, by ascending mean kinship.

Tiers 1--3 are flagged `High Value`. Because tier 2 ranks on genome uniqueness
and tier 3 on mean kinship, the two metrics can disagree -- and the top of the
table shows it:

```{r}
#| label: dual
head(rpt[, c("id", "indivMeanKin", "gu", "rank")], 3)
```

An animal need not have the very lowest mean kinship to rank first; carrying
rare alleles (high genome uniqueness) can place it ahead of a slightly less
related animal. That is the point of using both metrics rather than either one
alone.

```{r}
#| label: value-counts
table(rpt$value)
```

## Colony-level diversity {#sec-diversity}

Alongside the per-animal ranking, `reportGV()` reports two whole-colony measures
of how much of the original founders' diversity survives (Lacy 1989):

```{r}
#| label: fefg
c(foundersKnown = gv$total,
  maleFounders  = gv$nMaleFounders,
  femaleFounders = gv$nFemaleFounders)
round(c(fe = gv$fe, fg = gv$fg, fgSE = gv$fgSE), 2)
```

- **Founder equivalents** (`fe = 1 / sum(p^2)`, where `p` is each founder's
  proportional genetic contribution) is the *effective* number of founders --
  the number of equally contributing founders that would give the same diversity
  as the actual, unequal contributions. It is deterministic, so it has no
  sampling standard error.
- **Founder genome equivalents** (`fg`) goes further, also accounting for the
  founder alleles *lost* to genetic drift over the generations (estimated from
  the same gene drop). Because some alleles are always lost, `fg` is smaller than
  `fe`. Because `fg` is a Monte Carlo estimate, `reportGV()` also returns `fgSE`,
  its sampling standard error (shown above and displayed inline as `fg +/- fgSE`
  wherever FG appears); it shrinks roughly as `1 / sqrt(iterations)`. At small
  iteration counts `fg` also carries a slight finite-sample (Jensen) bias that
  the standard error does not capture, so prefer more iterations when the
  precision of `fg` matters.

Together they say, in effect, "the colony's diversity is equivalent to this many
ideal founders" -- a smaller number signals a narrower gene pool.

## Key arguments {#sec-arguments}

| Argument | Default | Meaning |
|---|---|---|
| `ped` | -- | the (trimmed, QC'd) pedigree with a `population` column |
| `guIter` | `1000L` | gene-drop iterations (more = more stable genome uniqueness) |
| `guThresh` | `1L` | an allele is "unique" if carried by no more than this many animals |
| `pop` | `NULL` | animal IDs to treat as the population (default: the `population` column) |
| `byID` | `TRUE` | count an individual's homozygous alleles once when scoring uniqueness |

## See also {#sec-seealso}

- The **Studbook Quality Control** article -- `qcStudbook()` validates and
  standardizes the studbook this analysis consumes, the first step before ranking.
- The **Building a Focal-Animal Pedigree Offline** article -- build a focal-animal
  pedigree from files with no database, via `getFocalAnimalPedFromFile()`.
- The **Forming Breeding Groups** article -- `groupAddAssign()` uses the kinship
  matrix this analysis returns to assemble genetically diverse groups.
- The **Age-Sex Pyramid Plots** article -- picture the colony's age and sex
  structure with `getPyramidPlot()`.
- `rankSubjects()` -- the ranking scheme applied to a report.
- `kinship()` -- pairwise kinship coefficients from a pedigree.
- `runGeneKeepR()` -- the Shiny app that performs this workflow interactively.

**References.**

Lacy RC (1989). "Analysis of Founder Representation in Pedigrees: Founder
Equivalents and Founder Genome Equivalents." *Zoo Biology* 8(2):111-123.

Vinson A, Raboin MJ (2015). "A Practical Approach for Designing Breeding Groups
to Maximize Genetic Diversity in a Large Colony of Captive Rhesus Macaques
(*Macaca mulatta*)." *Journal of the American Association for Laboratory Animal
Science* 54(6):700-707.
