---
title: "Studbook Quality Control"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview {#sec-overview}

Every analysis in `nprcgenekeepr` -- genetic value, breeding-group formation,
the age-sex pyramid -- starts from a **quality-controlled pedigree**. Real
studbooks arrive with inconsistent column names, mixed date formats, sex codes
that disagree with how an animal is actually used as a parent, duplicated
records, and impossible parentages. `qcStudbook()` is the single function that
validates and standardizes a studbook before anything else touches it; the
*Genetic Value Analysis* and *Forming Breeding Groups* articles both call it as
their very first step.

`qcStudbook()` performs several families of checks:

- **required columns** -- the studbook must contain `id`, `sire`, `dam`,
  `sex`, and `birth`;
- **identifier validity** -- animal IDs must be alphanumeric (a period is not
  allowed, because it breaks formulas, file names, and namespaces across
  software environments);
- **sex consistency** -- an animal used as a sire must be male and one used as
  a dam must be female;
- **date validity** -- birth and exit dates must parse to real calendar dates;
- **duplicate detection** -- the same ID must not appear in two conflicting
  records;
- **minimum parent age** -- a parent must be at least `minParentAge` years old
  (default 2.0) at an offspring's birth.

It runs in one of two modes, controlled by `reportErrors`:

- **production mode** (`reportErrors = FALSE`, the default) returns a clean,
  standardized data frame -- silently correcting what it can safely repair and
  **stopping** on anything it cannot;
- **diagnostic mode** (`reportErrors = TRUE`) does not stop. It scans the whole
  studbook and returns a *list* of everything it found, so you can see every
  problem in one pass and fix the source data.

## Setup {#sec-setup}

```{r}
#| label: load
#| message: false
library(nprcgenekeepr)
```

`qcStudbook()` is deterministic -- it runs no random simulation -- so, unlike
the genetic-value and breeding-group analyses (which seed a gene-drop
simulation), this article needs no `set_seed()` call to be reproducible.

## Cleaning a real studbook {#sec-clean}

The package ships `examplePedigree`, a realistic 3,694-animal studbook. Running
it through `qcStudbook()` in production mode returns a standardized pedigree
ready for analysis.

```{r}
#| label: qc
ped <- qcStudbook(examplePedigree,
  minParentAge  = 2.0,
  reportChanges = FALSE,
  reportErrors  = FALSE
)
dim(ped)
names(ped)
head(ped[, c("id", "sire", "dam", "sex", "gen", "birth", "exit", "age")])
```

The result has standardized columns, founders (animals with no known parent)
marked generation 0, and an `age` column. Because `examplePedigree` already
carries an `age` column, `qcStudbook()` keeps it rather than recomputing ages
against the current date, so this article renders the same numbers every time.
When the input has no `age` column, ages are computed from `birth` to `exit` --
or to today's date for animals still in the colony.

## What QC standardizes {#sec-standardize}

Before testing for errors, `qcStudbook()` normalizes the studbook so downstream
code sees consistent names and codes. Column names are lower-cased and stripped
of spaces, periods, and underscores, and common aliases are renamed
(`egoid` / `ego_id` -> `id`, `sireid` -> `sire`, `damid` -> `dam`,
`birthdate` -> `birth`). Setting `reportChanges = TRUE` records exactly what was
renamed. The shipped `pedGood` data set has five columns -- four with
deliberately messy headers (`ego_id`, `sire.id`, `dam_id`, `birth_date`)
and an already-canonical `sex` that QC leaves untouched:

```{r}
#| label: changes
chg <- qcStudbook(pedGood, reportChanges = TRUE, reportErrors = TRUE)
Filter(length, chg$changedCols)
```

`qcStudbook()` also standardizes coded values: sex to `M` / `F` / `U`
(from `MALE` / `FEMALE` / `1` / `2` / ...; by default the hermaphrodite codes
`H` / `HERMAPHRODITE` / `4` are folded into `U`), status to
`ALIVE` / `DECEASED` / `SHIPPED` / `UNKNOWN`, and ancestry to
`INDIAN` / `CHINESE` / `HYBRID` / `JAPANESE` / `UNKNOWN` / `OTHER`. Character
dates are parsed to `Date`, and `UNKNOWN` parents become either `NA` or
auto-generated `Unnnn` placeholder IDs.

## Diagnosing problems without stopping {#sec-diagnose}

In diagnostic mode (`reportErrors = TRUE`) the function returns a list whose
named elements each hold one category of finding. A **clean** studbook returns
`NULL` -- nothing to report:

```{r}
#| label: clean
is.null(qcStudbook(pedGood, reportErrors = TRUE))
```

The package ships small, purpose-built data sets that each trigger one kind of
problem; they make good worked examples.

**Sex inconsistency.** In `pedFemaleSireMaleDam`, `s1` is recorded female but is
the sire of two offspring, and `d1` is recorded male but is used as a dam:

```{r}
#| label: sex
qcStudbook(pedFemaleSireMaleDam,
  reportErrors = TRUE)[c("femaleSires", "maleDams")]
```

**Invalid dates.** `pedInvalidDates` contains `"205-06-19"` (a three-digit year)
and `"2002-16-22"` (month 16); the offending row numbers are reported:

```{r}
#| label: dates
qcStudbook(pedInvalidDates, reportErrors = TRUE)$invalidDateRows
```

**Duplicate records.** `pedDuplicateIds` repeats animal `d1` in two rows:

```{r}
#| label: dups
qcStudbook(pedDuplicateIds, reportErrors = TRUE)$duplicateIds
```

**A missing required column.** `pedMissingBirth` has no `birth` column at all:

```{r}
#| label: missing
qcStudbook(pedMissingBirth, reportErrors = TRUE)$missingColumns
```

**An impossible parentage.** In `pedSameMaleIsSireAndDam`, `s1` is listed as
both a sire and a dam:

```{r}
#| label: sireanddam
qcStudbook(pedSameMaleIsSireAndDam, reportErrors = TRUE)$sireAndDam
```

Diagnostic mode also reports parents younger than `minParentAge` at an
offspring's birth (in `suspiciousParents`) and any ID containing a period
(in `invalidIdChars`).

## Production mode vs diagnostic mode {#sec-modes}

The two modes treat the same problems differently. Production mode fixes what it
can prove is safe and refuses to return a pedigree it cannot trust; diagnostic
mode never stops, so a single pass surfaces every issue at once.

| Problem | Production mode (`reportErrors = FALSE`) | Diagnostic mode (`reportErrors = TRUE`) |
|---|---|---|
| Messy column names | corrected | `changedCols` (with `reportChanges = TRUE`) |
| Female sire / male dam | **corrected** (sex flipped) | `femaleSires` / `maleDams` |
| Exact duplicate record | **removed** | `duplicateIds` |
| Missing required column | **stops** | `missingColumns` |
| Invalid date | **stops** | `invalidDateRows` |
| Sire is also a dam | **stops** | `sireAndDam` |
| Parent below `minParentAge` | **stops** | `suspiciousParents` |
| Period in an ID | **stops** | `invalidIdChars` |

Production mode silently corrects the safe cases. We can confirm it on
`pedFemaleSireMaleDam` -- `s1` becomes male and `d1` female -- and on
`pedDuplicateIds`, where the repeated row is dropped:

```{r}
#| label: corrections
qcStudbook(pedFemaleSireMaleDam)[, c("id", "sex")]
nrow(pedDuplicateIds)             # before
nrow(qcStudbook(pedDuplicateIds)) # the duplicate row is gone
```

Run on data it cannot safely repair, production mode stops with a message naming
the offending rows -- here, the invalid dates:

```{r}
#| label: stop
#| warning: false
tryCatch(
  qcStudbook(pedInvalidDates),
  error = function(e) cat(conditionMessage(e))
)
```

The recommended workflow follows from this: run once in diagnostic mode to see
everything, correct the source studbook, then run in production mode to get the
clean pedigree the analyses consume.

## Key arguments {#sec-arguments}

| Argument | Default | Meaning |
|---|---|---|
| `sb` | -- | the studbook data frame to check |
| `minParentAge` | `2.0` | minimum age (years) a parent must be at an offspring's birth; skipped when birth dates are missing |
| `reportChanges` | `FALSE` | record column-name corrections in `changedCols` |
| `reportErrors` | `FALSE` | diagnostic mode: scan everything and return a list of findings (or `NULL`) instead of a cleaned pedigree |

## See also {#sec-seealso}

- The **Genetic Value Analysis** article -- ranks a quality-controlled pedigree
  by genetic value.
- The **Forming Breeding Groups** article -- assembles breeding groups from a
  quality-controlled pedigree.
- `qcStudbook()` -- the function documented here.
- `runModularApp()` -- the Shiny app, whose Quality Control tab drives this same
  function interactively.

**Reference.**

Vinson A, Raboin MJ (2015). "A Practical Approach for Designing Breeding Groups
to Maximize Genetic Diversity in a Large Colony of Captive Rhesus Macaques
(*Macaca mulatta*)." *Journal of the American Association for Laboratory Animal
Science* 54(6):700-707.
